### 2022-09-10 manuscript.Rmd
sparar en github kopia. omoget utkast.

---
title: "Quantifying unmet ICU beds in tertiary hospitals in urban India using Prospective Score Matching "
output: html_document
author: "Maria Eriksson"
bibliography: bibliography.bib
csl: vancouver.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- This is an R Markdown document. Markdown is a simple formatting
syntax for authoring HTML, PDF, and MS Word documents. For more
details on using R Markdown see http://rmarkdown.rstudio.com -->

<!-- This is a comment and will not be present in the compiled
document. Try "knitting" this document by clicking the knit button up
to the left, or by running `rmarkdown::render("manuscript.Rmd")` in
the console and you should only see headings -->

<!-- Structure and most comments are from the Strengthening the
reporting of observational studies in epidemiology (STROBE) statement,
see https://www.strobe-statement.org/. For more explanations see the
paper Strengthening the Reporting of Observational Studies in
Epidemiology (STROBE):explanation and elaboration.  -->

<!-- Indicate the study’s design with a commonly used term in the
title, for example: "Associations between resuscitation interventions
and opportunities for improvement in adult trauma patients: A cohort
study" -->

Abstract
========

<!-- Provide in the abstract an informative and balanced summary of
what was done and what was found. Not more than 300 words. -->

Background
----------

Methods
-------

Results
-------

Conclusion
----------

Introduction
============

<!-- Explain the scientific background and rationale for the
investigation being reported. End by stating the aim or specific
objectives, including any prespecified hypotheses. The introduction
should not be longer than 1-1.5 pages, typically 4-5 paragraphs, each
6-8 lines long. Support every statement with a citation -->
Trauma/Injury is a substantial global health problem that disproportionately affects low and middle-income countries (LMIC), resulting in significant mortality and morbidity@traumaLMIC. Health care professionals in many LMIC testify that their medical assessment oftentimes is that a trauma patient should be admitted to the ICU, but aren’t because such resources are scarce, overloaded and unavailable. Their medical judgement is that the lives of these patients were salvageable, and that these patients died as a direct result of not being admitted to the ICU@stuffLMICdoctorssay.

In order to improve intra-hospital care, trauma quality improvement programs (TQIP) have been developed. The aim of TQIPs is to ensure equivalent care and improve trauma care by developing best practice guidelines@tqip. TQIPs utilizes a retrospective cohort study, analysing the current of trauma care patients and post-fact attempting to determine the appropriate level of care for each patient. From these analyses many aspects of trauma care can be improved and optimized, such as determining the appropriate amount of ICU beds and other trauma care resources.@tqip

The existing TQIP often originate in high-income countries, and are either not applicable nor easily transferred to the hospital conditions of low and middle income countries (LMIC). This study aims to make an TQIP assessment in an Indian context and fill part of that knowledge gap, and to test whether there is merit in the testimony of the Indian colleagues. if there is indeed a shortage of ICU beds in Indian hospitals, further quantify the scope of that shortage. If this analysis indeed indicate a shortage, it can be used to motivate entending ICU resources and reduce preventable trauma morbidity and mortality.

Methods 
=======

Study design
------------

<!-- Present key elements of study design -->

Setting
-------

<!-- Describe the setting, locations, and relevant dates, including
periods of recruitment, exposure, follow-up, and data collection -->

Participants
------------

<!-- Cohort study: Give the eligibility criteria, and the sources and
methods of selection of participants. Describe methods of
follow-up. For matched studies, give matching criteria and number of
exposed and unexposed -->

<!-- Case-control study: Give the eligibility criteria, and the
sources and + methods of case ascertainment and control
selection. Give the rationale for the choice of cases and
controls. For matched studies, give matching criteria and the number
of controls per case -->

<!-- Cross-sectional study: Give the eligibility criteria, and the
sources and methods of selection of participants -->

Variables and data sources/measurements
---------------------------------------

<!-- Clearly define all outcomes, exposures, predictors, potential
confounders, and effect modifiers. Give diagnostic criteria, if
applicable. For each variable of interest, give sources of data and
details of methods of assessment (measurement). Describe comparability
of assessment methods if there is more than one group -->

Bias 
----

<!-- Describe any efforts to address potential sources of bias -->

Study size 
----------

<!-- Explain how the study size was arrived at -->

Quantitative variables
----------------------

<!-- Explain how quantitative variables were handled in the
analyses. If applicable, describe which groupings were chosen and why
-->

Statistical methods
-------------------

<!--

(a) Describe all statistical methods, including those used to control
for confounding

(b) Describe any methods used to examine subgroups and interactions

(c) Explain how missing data were addressed

(d)

Cohort study: If applicable, explain how loss to follow-up was addressed

Case-control study:If applicable, explain how matching of cases and
controls was addressed

Cross-sectional study: If applicable, describe analytical methods
taking account of sampling strategy

(e) Describe any sensitivity analyses
-->

Results
=======

<!--

Participants
------------

(a) Report numbers of individuals at each stage of study—eg numbers
potentially eligible, examined for eligibility, confirmed eligible,
included in the study, completing follow-up, and analysed

(b) Give reasons for non-participation at each stage

(c) Consider use of a flow diagram

Descriptive data
----------------

(a) Give characteristics of study participants (eg demographic,
clinical, social) and information on exposures and potential
confounders

(b) Indicate number of participants with missing data for each
variable of interest

(c) Cohort study — Summarise follow-up time (eg, average and total
amount)

Outcome data
------------

Cohort study — Report numbers of outcome events or summary measures
over time

Case-control study — Report numbers in each exposure category, or
summary measures of exposure

Cross-sectional study — Report numbers of outcome events or summary
measures

Main results
------------

(a) Give unadjusted estimates and, if applicable, confounder-adjusted
estimates and their precision (eg, 95% confidence interval). Make
clear which confounders were adjusted for and why they were included

(b) Report category boundaries when continuous variables were
categorized

(c) If relevant, consider translating estimates of relative risk into
absolute risk for a meaningful time period

Other analyses 
--------------

Report other analyses done—eg analyses of subgroups and interactions,
and sensitivity analyses 
-->

You can include code in this document like this:

```{r main, echo=FALSE}
source("main.R") ## This "imports" the main script file of your project and run any code in it
```

You can also embed plots:

```{r plot, echo=FALSE}
plot(pressure)
```

You can also mix text and code, so called inline code, like this: `r 2+5`.

Discussion
==========

<!--

Key results
-----------
Summarise key results with reference to study objectives

Limitations
-----------

Discuss limitations of the study, taking into account sources of
potential bias or imprecision.  Discuss both direction and magnitude
of any potential bias

Interpretation
--------------

Give a cautious overall interpretation of results considering
objectives, limitations, multiplicity of analyses, results from
similar studies, and other relevant evidence

Generalisability
----------------

Discuss the generalisability (external validity) of the study results

-->

Conclusion
==========

<!-- Give a short conclusion, corresponding to the aim -->

References
==========

<!-- Do not edit by hand, references will be inserted and formatted automatically once you knit this document -->

### 2022-09-11 manuscript.Rmd
sparar en github kopia. ganska babbligt ännu.

@@
 knitr::opts_chunk$set(echo = TRUE)
+table_nums <- captioner::captioner(prefix = "Tab.")
 ```
@@
 TQIPs utilizes a retrospective cohort study, analysing the current of trauma care patients and post-fact attempting to determine the appropriate level of care for each patient.
-From these analyses many aspects of trauma care can be improved and optimized, such as determining the appropriate amount of ICU beds and other trauma care resources.@tqip
+From these analyses many aspects of trauma care can be improved and optimized, such as determining the appropriate amount of ICU beds and other trauma care resources @tqip.
 
-The existing TQIP often originate in high-income countries, and are either not applicable nor easily transferred to the hospital conditions of low and middle income countries (LMIC).
-This study aims to make an TQIP assessment in an Indian context and fill part of that knowledge gap, and to test whether there is merit in the testimony of the Indian colleagues.
-if there is indeed a shortage of ICU beds in Indian hospitals, further quantify the scope of that shortage.
-If this analysis indeed indicate a shortage, it can be used to motivate entending ICU resources and reduce preventable trauma morbidity and mortality.
+Existing TQIP often originate in high-income countries, and are either not applicable nor easily transferred to the hospital conditions of low and middle income countries (LMIC).
+This study aims to make an TQIP assessment in an Indian context to fill part of that knowledge gap, and to test whether there is merit in the testimony of the Indian colleagues.
+If there is indeed a shortage of ICU beds in Indian hospitals, the study shall further quantify and propose a scope of the unmet need for ICU beds for trauma patients at tertiary hospitals in urban India.
+If this analysis indeed indicate a shortage, it can be used to motivate extending ICU resources and reduce preventable trauma morbidity and mortality.
 
@@
 <!-- Present key elements of study design -->
+Observational retrospective cohort study from data collected at four tertiary hospitals in urban India.
 
@@
 periods of recruitment, exposure, follow-up, and data collection -->
+FRÅGA: kopierat rakt av härifrån https://github.com/titco/titco-I
+Men det är ju exakt vad som ska under den här rubriken. Får man göra så eller är det plagiat?
+The TITCO-I v1 dataset includes data on 16 000 patients admitted to four public university hospitals in urban India.
+The original dataset included 16 047 patients but 47 patients were deleted from the dataset for anonymization purposes, see the anonymization section for more details.
+At the time of the research Jai Prakash Narayan Apex Trauma Center (JPNATC), All India Institute of Medical Sciences, New Delhi, was a dedicated trauma centre with almost 180 beds.
+King Edward Memorial hospital (KEM), Mumbai, was a tertiary level hospital with no dedicated trauma ward.
+Lokmanya Tilak Municipal General Hospital (LTMGH), Mumbai was a tertiary level public university hospital with a dedicated trauma ward with 14 beds.
+The Institute of Post-Graduate Medical Education and Research and Seth Sukhlal Karnani Memorial Hospital (SSKM), Kolkata was a tertiary level public university hospital, but without a dedicated trauma ward.
+The cost of care in these hospitals was nominal and the patients admitted represented mainly a lower socioeconomic stratum of the population.
+The first patient was enrolled in July 2013 and the last patient in December 2015.
+The exact enrollment period varied across participating hospitals.
 
@@
 exposed and unexposed -->
+The TITCO1 data set includes all data recorded for trauma patients at the participating hospitals is included in the TITCO1 study.
 
@@
 
+There are 193 possible observations recorded for each patient. A general presentation of these observations is in table_nums('tab_1')
+
+tab.1_cap <- table_nums(name = "tab_1", 
+                        caption = "Data types collected in the TITCO1 dataset")
+
+                     knitr::kable(table1_1, "simple")
+
+datatvätt:
+FRÅGA: quantitativa variablerna har många NA som jag tror egentligen är NO, ex om pat har haft syrgas vid spO2 mätning.
+kan jag tänka lite fritt här när det känns uppenbart eller ska jag förankra?
+också finns det ex obs där numeriska värdet för spO2 saknas men ändå data yes/no om pat fått syrgas och jag vet inte om sån data är meningsfull?
+FRÅGA: vad betyder GCS verbal V1?
+det är inte felregistrerat det är ett alternativ i codebook.
+jag har hittat ett förslag om att det är pat som inte kan ge verbalt svar pga intub eller surgical airway, men det finns pat som har varken eller och GCS verbalt = 1v
+
+The quantitative data include recorded parameters such as vital parameters, lab results, GCS total score and cumulative time in surgery or ICU, if applicable.
+Measurement or assessment techniques are standardized.
+
+The qualitative data include categorical data such as coarse categorization or the accident mechanism and injury, details about arrival circumstances, and whether or not a patient had received a medical examination or intervention.
+
+Text data are freehand examination results and operation categorizations, as well as assigned ICD codes.
+These data are vast and contain a lot of useful information, but also introduce a lot of challenges.
+There is a lack of standardization, user dependence, heterogenous grouping caused by using too general ICD codes, and datamining challenges for this type of data.
+If it may cause inaccuracies or practical challenges, it shall be omitted.
+
+The intermediate outcome of interest is ICU admission, irregardless of duration, and to identify the predictors that warrant ICU admission.
+Therefore, the properties of the patients that were admitted into the ICU will be examined.
+The variables with the highest propensity to warrant a ICU admission to be ICU - predictors - will be identified.
+Using these predictors and propensity score matching will identify patients that weren’t admitted into the ICU, but presented with a set of predictors - a propensity score - very similar to those that were admitted.
+The ultimate outcome of this analysis is to quantify this group of patients and consequently quantify the unmet need for ICU beds.
+
 Bias 
@@
 <!-- Describe any efforts to address potential sources of bias -->
+The TITCO1 dataset are observational data collected from a clinical environment and not an experimental setting.
+Producing a randomized control trial study for these types of lifesaving medical interventions for trauma patients would obviously be morally indefensible.
+Medical interventions, examinations and ICU admission are not randomized, they are a consequence of a patients medical condition.
+There exists a causal effect between the independent and the dependent variables: a patient is sent to the ICU precisely because of the severity of their condition.
+
+This causality introduces biases that must be addressed in the statistical analysis.
+Traditionally, multivariable regression has been used tocontrol for confounding in these cases but methods based on propensity scores have become popular alternatives.
+Propensity Score Matching will be descibed in detail in chapter Statistical methods.
 
@@
 <!-- Explain how the study size was arrived at -->
+The data of 16047 trauma patients were collected at the participating hospitals over 2.5 years, as described in chapter Settings
+47 entries have been randomly excluded for anonymization purposes, and for the purpose of this study the 120 patients who were dead on arrival have been excluded aswell.
 
@@
 -->
+Fråga: hur diffar detta från vad som finns i Variables and data sources/measurements
 
@@
 for confounding
+-->
+Propensity score matching will be used in this analysis.
+Propensity score matching (PSM) is typically performed to determine the efficacy of a conditionalized intervention - causality between the dependent variables and available independent variables - where there is a need to offset or balance the effect of this causality.
+<!-- 
+The goal of propensity score matching is usually to identify suitable control subjects that did not receive an intervention, even though they fairly well met the same inclusion criteria that the subjects that did receive the intervention had.
+Control subjects who are similar enough to the intervention group in the most relevant aspects for receiving treatment.
+-->
+In this context PSM could provide tools for an unbiased examination of the efficacy of ICU care on mortality.
+When simply comparing the cohort that did go to the ICU to the cohort that did not, it would appear that ICU care actually increased mortality.
+In the TITCO1 dataset the survival rate of patients who did not go to the ICU was 81% versus 68% for patients who did.
+This is of course an incorrect conclusion, the higher mortality rates in the ICU is because patients who are admitted here are worse off.
+If, however, the conditions that typically warrant ICU admission could be identified, a subset of the TITCO1 dataset that fairly well met those criteria but did not go to the ICU could be identified using propensity score matching.
+
+Identifying a subset of patients that should have received ICU care but didn't is the end point for this thesis.
+In the typical PSM context this subset is further used as the control group to determine the efficacy of ICU care.
+
+The first step is to identify predictor variables: independent variables most predictive of the outcome or dependent variable.
+In this context it means to first identify the conditions and medical interventions received that correlate to ICU care.
+All properties of the patients that were sent to the ICU will be compared to the patents who didn't, in order to identify properties that distinguishes this cohort.
+
+Second step is to calculate propensity scores for all patients.
+Each patient in the TITCO1 data set gets a propensity score based on their medical conditions and interventions, specifically their individual values of the predictor variables.The propensity score determines the likelihood that a patient will receive ICU care, given their condition.
+
+Third step is PSM.
+PSM identify members of the subset that did not get the intervention but had the same predictors as the subgroup that did get the intervention.
+The purpose of the propensity score matching is identifying patients that weren’t admitted into the ICU, but presented with a set of predictors - a propensity score - very similar to those that were admitted.
 

### 2022-09-15 manuscript.Rmd
current version

@@
 Trauma/Injury is a substantial global health problem that disproportionately affects low and middle-income countries (LMIC), resulting in significant mortality and morbidity@traumaLMIC.
+The annual mortality from injury in India amount to one million casualties @onemilliondead.
 Health care professionals in many LMIC testify that their medical assessment oftentimes is that a trauma patient should be admitted to the ICU, but aren’t because such resources are scarce, overloaded and unavailable.
@@
 This study aims to make an TQIP assessment in an Indian context to fill part of that knowledge gap, and to test whether there is merit in the testimony of the Indian colleagues.
-If there is indeed a shortage of ICU beds in Indian hospitals, the study shall further quantify and propose a scope of the unmet need for ICU beds for trauma patients at tertiary hospitals in urban India.
+The study shall propose a scope of the unmet need for ICU beds for trauma patients at tertiary hospitals in urban India.
 If this analysis indeed indicate a shortage, it can be used to motivate extending ICU resources and reduce preventable trauma morbidity and mortality.
+
+Trauma disproportionately affects young, otherwise healthy individuals.
+They are often the primary breadwinners for their families and therefore the detrimental impact of trauma go beyond mortality or substantial loss of disease-free years of life for the affected individuals, it also has a considerable financial and social harm on their family members and communities.
+@traumaLMIC
+
+Aim
+-----------
+Predict the unmet need for ICU beds for trauma patients at tertiary hospitals in urban India by applying propensity score matching.
+
+The theory being examined in this thesis is that there exist subset of the TITCO1 dataset that didn't go to the ICU, but are very similar to the patients that did go to the ICU, in the covariates that caused ICU-treated patients to receive this intervention.
+The thesis is that these patients were not admitted to the IC for lack of such resources.
+These patients will be identified using propensity score matching.
+The size of this group in relation to the group that were admitted to the ICU indicate the magnitude of the unmet need of ICU beds.
+If, for every patient that did receive ICU care, there was another patient with similar predictors that wasn’t admitted due to scarcity of ICU resources, that would mean the ICU resources are 50% of what they ought to be.
 
@@
 exposed and unexposed -->
-The TITCO1 data set includes all data recorded for trauma patients at the participating hospitals is included in the TITCO1 study.
+The TITCO1 data set includes all data recorded for trauma patients at the participating hospitals. The patients have the common denominators than geographical location and trauma.  
+FRÅGA: finns en socioekonomisk selektion ? läs: är detta sjukhus dit vissa socioekonomiska grupper hamnar?
 
@@
 
-The intermediate outcome of interest is ICU admission, irregardless of duration, and to identify the predictors that warrant ICU admission.
-Therefore, the properties of the patients that were admitted into the ICU will be examined.
-The variables with the highest propensity to warrant a ICU admission to be ICU - predictors - will be identified.
-Using these predictors and propensity score matching will identify patients that weren’t admitted into the ICU, but presented with a set of predictors - a propensity score - very similar to those that were admitted.
-The ultimate outcome of this analysis is to quantify this group of patients and consequently quantify the unmet need for ICU beds.
-
 Bias 
@@
 The TITCO1 dataset are observational data collected from a clinical environment and not an experimental setting.
-Producing a randomized control trial study for these types of lifesaving medical interventions for trauma patients would obviously be morally indefensible.
 Medical interventions, examinations and ICU admission are not randomized, they are a consequence of a patients medical condition.
@@
 <!-- Explain how the study size was arrived at -->
-The data of 16047 trauma patients were collected at the participating hospitals over 2.5 years, as described in chapter Settings
+The data of 16047 trauma patients were collected at the participating hospitals over 2.5 years, as described in chapter Settings.
 47 entries have been randomly excluded for anonymization purposes, and for the purpose of this study the 120 patients who were dead on arrival have been excluded aswell.
+
+Återkom: XX patients were discarded because of missing data
 
@@
 Propensity score matching will be used in this analysis.
-Propensity score matching (PSM) is typically performed to determine the efficacy of a conditionalized intervention - causality between the dependent variables and available independent variables - where there is a need to offset or balance the effect of this causality.
+Propensity score matching (PSM) is best explained using one typical application: identifying a control group when examining the efficacy of an intervention for a specific outcome, where there is causality between the independent variables for the intervention, and sometimes also the outcome.
+There is a need to offset or balance the effect of this causality, but a randomized case-control study could be impractical or immoral.
+
+In this context, it could be to examine the efficacy of ICU care on survival.
+If the correlation between the independent variables and the intervention and outcome isn't accounted for, a simple comparison between the cohort that did go to the ICU to the cohort that did not, would find that ICU care actually increased mortality.
+In the TITCO1 dataset the survival rate of patients who did not go to the ICU was 81% versus 68% for patients who did.
+This is of course an incorrect conclusion, this data has treatment selection bias as patient covariates often confound treatment selection.
+The higher mortality observed in patients who received ICU care is because patients were selected for the intervention because they presented certain medical conditions that also carry a high mortality - even with the intervention ICU care.
+Producing a randomized control trial study for these types of lifesaving medical interventions for trauma patients would obviously go against ethical guidelines of research, and another approach to produce a control group is necessary.
+Propensity score matching is a statistical procedure for reducing this bias by assembling a sample from within the data set, in which confounding factors are balanced between the cohort that got the intervention and the cohort that did not.
 <!-- 
@@
 -->
-In this context PSM could provide tools for an unbiased examination of the efficacy of ICU care on mortality.
-When simply comparing the cohort that did go to the ICU to the cohort that did not, it would appear that ICU care actually increased mortality.
-In the TITCO1 dataset the survival rate of patients who did not go to the ICU was 81% versus 68% for patients who did.
-This is of course an incorrect conclusion, the higher mortality rates in the ICU is because patients who are admitted here are worse off.
-If, however, the conditions that typically warrant ICU admission could be identified, a subset of the TITCO1 dataset that fairly well met those criteria but did not go to the ICU could be identified using propensity score matching.
+The size of this subset in relation to the subset of patients that did receive ICU care is the end point for this analysis.
+This proportion is indicative of the unmet need of ICU beds, as described in chapter Aim.
+In the typical PSM appliction, this subset is used as the control group to determine the efficacy of ICU care for some outcome, eg mortality.
 
-Identifying a subset of patients that should have received ICU care but didn't is the end point for this thesis.
-In the typical PSM context this subset is further used as the control group to determine the efficacy of ICU care.
-
-The first step is to identify predictor variables: independent variables most predictive of the outcome or dependent variable.
-In this context it means to first identify the conditions and medical interventions received that correlate to ICU care.
+The first step is to identify predictor variables: covariates most predictive of the dependent variable.
+In this context it means to first identify the conditions and medical interventions received that correlate best to being offered ICU care.
 All properties of the patients that were sent to the ICU will be compared to the patents who didn't, in order to identify properties that distinguishes this cohort.
@@
 Second step is to calculate propensity scores for all patients.
-Each patient in the TITCO1 data set gets a propensity score based on their medical conditions and interventions, specifically their individual values of the predictor variables.The propensity score determines the likelihood that a patient will receive ICU care, given their condition.
+Each patient in the TITCO1 data set gets a propensity score based on their medical conditions and interventions, specifically their individual values of the predictor variables.
+The propensity score determines the likelihood that a patient will receive ICU care, given their condition.
 
 Third step is PSM.
-PSM identify members of the subset that did not get the intervention but had the same predictors as the subgroup that did get the intervention.
+PSM is a standardized mathematical approach to identify members of the subset that did not get the intervention but had the same predictors as the subgroup that did get the intervention.
 The purpose of the propensity score matching is identifying patients that weren’t admitted into the ICU, but presented with a set of predictors - a propensity score - very similar to those that were admitted.
@@
 (c) Explain how missing data were addressed 
+The data completion rate is presented in table_nums('tab_1').
+As the base assumption, all quantitative variables collected for each patient are considered potential recordings.
+Data that are missing are assumed missing because care for a patient did not motivate the collection of the data.
+For example, if serum creatinine was not recorded, it is assumed that the collection of this variable wasn't relevant for care of the patient.
+
+There are some variables that should reasonably have been collected and documented given the patient condition, but omission will not be cause of dismissal unless these variables are predictors.
+
+All patients with incomplete observations for any predictors will be discarded.
+For relevant qualitative variables, missing data did not automatically warrant dismissal, if the missing data could be indirectly proven.
+For example if the quantitative yes/no for an examination like a CT scan was missing, it was deemed a 'yes' if there was a CT scan result recorded.
+If both were missing, the patient data was dismissed.
+All vital parameters are considered mandatory, irregardless of patient condition so patients without data of heart rate, systolic blood pressure or respiration rate at any time of their hospital stay, have been discarded.
+FRÅGA: right?
+
+All quantitative data from the original dataset that were entered as zero but where a zero value is physiologically impossible were replaced with NA values.
+Zero values for variables where it is possible to have zero values, such as respiration rate and heart rate were left as zeroes to not tamper with the data, although a zero value indicate that the variable probably wasn't measured.
+These amount to 250 and 155 entries respectively for the first observation within an hour and less than 100 for both for the second observation after 24 h, so the choice to include or exclude these observations won’t skew the data considerably.
 

### 2022-09-15 main.R

@@ -18,7 +18,7 @@
 ## them). They all need to be sourced however when you compile your
 ## manuscript file or run this file as a job, as that happens in a
 ## clean R session.
-noacsr::source_all_functions()
+#noacsr::source_all_functions()
 library(tidyverse)
 library(tableone)
 
@@ -26,9 +26,21 @@ library(tableone)
 headerproperties <- read.csv("headerproperties.csv")
 titcodata <- read.csv("titcodata.csv")
 
+#FRÅGA: jag kan inte hitta hur man definierar eller filtrerar på row headers, men jag kan göra det här superjobbiga sättet
+typefilt=filter(headerrows, name == "type") #gör en dataframe av type-raden
+typetrans <- data.table::transpose(typefilt) #transponera den. jo. humor me. jag måste transponera eftersom jag bara vet hur man filtrerar på kolumnnamn
+timers=filter(typetrans, V1 == "time" | V1 == "date") #NU kan jag applicera den här arbetshästen
+counttimers=length(timers)
+#och det kan jag bara göra för att göra counts jag har ju tappat alla associationer inom datasetet
+
 #ladda vektorerna med kvantitativa och kvalitativa datan
 load("quantitativecolumnname.rdata")
 load("qualitativecolumnname.rdata")
+load("textcolumnname.rdata")
+
+#sortera bort DOA
+alive=filter(titcodata, incl != 2)
+titcodata=alive
 
 ## Whatever you do next, maybe clean data?
 #definiera dataseten
@@ -38,10 +50,12 @@ nolicu=filter(titcodata, licu <0.5)
 #välj kvant kolumnerna
 quantitativeLicu <- licu[, which((names(licu) %in% quantitativecolumnname)==TRUE)]
 quantitativeNolicu <- nolicu[, which((names(nolicu) %in% quantitativecolumnname)==TRUE)]
+quantitativeTitco <- titcodata[, which((names(titcodata) %in% quantitativecolumnname)==TRUE)]
 
 #gör stat analysen
 quantitatestatlicu=CreateTableOne(data = quantitativeLicu,includeNA=FALSE)
 quantitatestatnolicu=CreateTableOne(data = quantitativeNolicu,includeNA=FALSE)
+
 
 #gör t.test för alla kvant kolumner
 columncount=length(quantitativecolumnname) #hur många kolumner ska vi rulla

### 2022-09-16 manuscript.Rmd
tabellerna, basdatan,

@@
 The annual mortality from injury in India amount to one million casualties @onemilliondead.
+Half of trauma mortality in India occur in-hospital, and in-hospital trauma mortality in Indian hospitals was double that of high-income countries @indiantrauma.
 Health care professionals in many LMIC testify that their medical assessment oftentimes is that a trauma patient should be admitted to the ICU, but aren’t because such resources are scarce, overloaded and unavailable.
 Their medical judgement is that the lives of these patients were salvageable, and that these patients died as a direct result of not being admitted to the ICU@stuffLMICdoctorssay.
+Globally, an estimated two million lives could be saved annually, if the appropriate and timely trauma care provided by hospitals in many high-income countries were available and provided in low to-middle-income countries (LMICs) @livessaved
 
@@
 Trauma disproportionately affects young, otherwise healthy individuals.
-They are often the primary breadwinners for their families and therefore the detrimental impact of trauma go beyond mortality or substantial loss of disease-free years of life for the affected individuals, it also has a considerable financial and social harm on their family members and communities.
+They are often the primary breadwinners for their families and therefore the detrimental impact of trauma go beyond mortality or substantial loss of disease-free years of life for the affected individuals, it also carries a considerable financial and social harm on their family members and communities.
 @traumaLMIC
@@
 These patients will be identified using propensity score matching.
+
 The size of this group in relation to the group that were admitted to the ICU indicate the magnitude of the unmet need of ICU beds.
@@
 exposed and unexposed -->
-The TITCO1 data set includes all data recorded for trauma patients at the participating hospitals. The patients have the common denominators than geographical location and trauma.  
+The TITCO1 data set includes all data recorded for trauma patients of the emergency departments at the participating hospitals.
+Trauma covers injury and with a mechanism of road traffic, railway, fall, burn or assault.
 FRÅGA: finns en socioekonomisk selektion ? läs: är detta sjukhus dit vissa socioekonomiska grupper hamnar?
+
+47 patients were excluded for anonymization purposes, 120 patients were excluded because they were dead on arrival, and further XX pat were excluded because of missing data.
 
@@
 
-(c) Explain how missing data were addressed 
+Missing Data
+-------------
 The data completion rate is presented in table_nums('tab_1').

### 2022-09-16 main.R

@@ -25,6 +25,9 @@ library(tableone)
 ## Import data
 headerproperties <- read.csv("headerproperties.csv")
 titcodata <- read.csv("titcodata.csv")
+#jag har manuellt ändrat och städat lite i dessa så jag sparar dem så jag kan committa dem:
+save(headerproperties, file = "headerproperties.RData")
+save(titcodata, file = "titcodata.RData")
 
 #FRÅGA: jag kan inte hitta hur man definierar eller filtrerar på row headers, men jag kan göra det här superjobbiga sättet
 typefilt=filter(headerrows, name == "type") #gör en dataframe av type-raden
@@ -67,8 +70,31 @@ for (a in 1:columncount) {
   tval=c[["statistic"]][["t"]]
   pvalues[a,1]=pval
   tvalues[a,1]=tval
+
+  
   
 }
 
-df_t <- data.table::transpose(quantitativecolumnname)
+
+df_t <- data.table::transpose(quantitativelong)
+
+#det här är tabellen med P pch T-värden
 statresult<-data.frame(df_t,pvalues,tvalues) #jag vet inte hur jag får headers att stå p och t-värde, men det blir en tabell
+colnames(statresult)<-data.frame("variable short","p-värde","t-värde")
+
+b=quantitatestatlicu[["ContTable"]][["Overall"]]
+myDf <- as.data.frame(b)
+ICU_mean=myDf$mean
+ICU_sigma=myDf$sd
+myDfmeanlicu <- as.data.frame(ICU_mean)
+myDfsdlicu <- as.data.frame(ICU_sigma)
+
+b=quantitatestatnolicu[["ContTable"]][["Overall"]]
+myDf <- as.data.frame(b)
+noICU_mean=myDf$mean
+noICU_sigma=myDf$sd
+myDfmeannolicu <- as.data.frame(noICU_mean)
+myDfsdnolicu <- as.data.frame(noICU_sigma)
+
+#det här är mean och std av ICU vs no-ICU
+table2<- data.frame(df_t,myDfmeanlicu,myDfsdlicu,myDfmeannolicu,myDfsdnolicu)

### 2022-09-18 manuscript.Rmd
dagens version

@@
 
+Ethical considerations
+----------------------
+The data has been thoroughly anonymized.
+Each patient have been given an ID number, time frames have been shifted to make identification by admission and discharge or death dates impossible, and a further 47 patients have been omitted from the data set to shift the data frames and disable tracking.
+
+The study is observational, the care provided to each patient has been evidence-based best available care.
+The data are standard details that would have been recorded anyway, and collection has not delayed or interfered with care nor added any cost.
+
+FRÅGA: jag kan inte se att consent tagits?
+
 Aim
@@
 The Institute of Post-Graduate Medical Education and Research and Seth Sukhlal Karnani Memorial Hospital (SSKM), Kolkata was a tertiary level public university hospital, but without a dedicated trauma ward.
-The cost of care in these hospitals was nominal and the patients admitted represented mainly a lower socioeconomic stratum of the population.
+The cost of care in these hospitals was nominal.
 The first patient was enrolled in July 2013 and the last patient in December 2015.
@@
 Trauma covers injury and with a mechanism of road traffic, railway, fall, burn or assault.
-FRÅGA: finns en socioekonomisk selektion ? läs: är detta sjukhus dit vissa socioekonomiska grupper hamnar?
+The patients represent mainly the lower stratum of socioeconomic stratum of their communities.
 
@@
 of assessment methods if there is more than one group -->
-
 There are 193 possible observations recorded for each patient. A general presentation of these observations is in table_nums('tab_1')
@@
 
-This causality introduces biases that must be addressed in the statistical analysis.
-Traditionally, multivariable regression has been used tocontrol for confounding in these cases but methods based on propensity scores have become popular alternatives.
-Propensity Score Matching will be descibed in detail in chapter Statistical methods.
+If the purpose of this thesis were to examine the efficacy of ICU care on eg.
+mortality, this bias would have to be addressed using some statistical tool, but it is not.
+The purpose of this thesis is use these same methods to assemble a control group but not for the purpose of examining the efficacy of ICU care on eg.
+mortality, but to identify and quantify the subset of patients that should have been admitted to the ICU but weren't.
+
+Propensity Score Matching will be described in detail in chapter Statistical methods.
+
+FRÅGA: Bör jag tänka på någon bias iallafall?
+Jag kan tänka mig att det finns skillnader mellan sjukhusen som bör undersökas (de har väl inte nödvändigtvis exakt samma proportioner ICU-sängar givet upptagningsområde, och att ett sjukhus därför kan vara mycket mer generös med ICU-sängarna än ett annat?), men vad för typ av statistisk bias är det isåfall?
+Det finns en selection bias i att det är lägre socioekonomisk stratum, men hur relevant är den?
+
+Propensity score matching in itself introduces a source of bias.
+The researcher chooses the predictor variables and so can introduce a self-fulfilling prophecy.
+For this reason the dataset shall be randomly split into two parts of 40% and 60% of the patients.
+The 40% shall be used to produce the propensity score matching, and the 60% shall be used to verify it.
+Fråga: typ?
 
@@
 
+The two subgroups were patients admitted to the ICU, and the patients who weren't. Admission to the iCU is a binary yes or no, length of stay has not been taken into Consideration.
+
 Missing Data
@@
 -->
+
+Predictor selection 
+-------------------
+The quantitative data did not produce any clear results see table_nums('tab_2')
+
+tab.2_cap <- table_nums(name = "tab_2", 
+                        caption = "Means and standard devations of quantitative variables of ICU and non-ICU admitted patients, respectively")
+
+                     knitr::kable(table3, "simple"),
+
+This is perhaps because of a crowd out effect.
+For instance, the total Glascow coma scale score is high and quite similar for the non-ICU subset versus the ICU subset, the difference between the average of the groups is 2 points.
+However, considered individually, low scores in either category is a high propensity to admit a patient to the ICU, as demonstrated in the comparison of quantitative parameters.
+
+P-values and T-values of non-ICU vs ICU patients:
+
+tab.3_cap <- table_nums(name = "tab_3", 
+                        caption = "P- and t-values of the quantitative values in comparison of non-ICU and ICU patients")
+
+                     knitr::kable(statresult, "simple"),
+
+FRÅGA: är p-värdena rimliga tycker du? de är superlåga och min känsla - med 5 hp statistik i ryggen - är att std-avvikelsen är för stor för det?
+
+For the qualitative data, the propensity of each parameter to get a patient admitted to the ICU is in relation to the proportion of patients that were admitted to the ICU from the whole TITCO1 data set, about a third.
+If considerably more than a third of each qualitative parameter are admitted to the ICU, then that parameter is of interest.
+The proportions are presented in tab.4_cap <- table_nums(name = "tab_4",
+                        caption = "Proportion of patients admitted to the ICU for all variants of all quuantitative variables")
+
+                     knitr::kable(qualitativetable, "simple"),
 

### 2022-09-19 createtable.R
en liten kodsnutt som skapat några tabeller och som ganska väl illustrerar alla extra rader jag använder för att få till tabeller. jag transponerar fram och tillbaka för att kunna göra olika saker, byter objekttyper fram och tuillbaka osv osv
och det är alltid återvändsgränder, ex om jag vill räkna något som ex hur många pat som haft intubation går jag rakt på det och tar fram bara den siffran.

helst skulle jag ju skapa en dataframe med de patienterna och sen göra en count på den dataframen. då får jag siffran jag ville ha men jag har också en dataframe som jag sen kan trolla vidare med om ja

@@ -1,0 +1,75 @@
+
+
+# for (a in 1:7) {
+#   
+# currentname = headerproperties[a,1]
+# 
+# currentnamerow=filter(headerrows, name == currentname) #gör en dataframe av type-raden
+# currentnamerowtrans <- data.table::transpose(currentnamerow) #transponera den. jo. humor me. jag måste transponera eftersom jag bara vet hur man filtrerar på kolumnnamn
+# timers=filter(typetrans, V1 == "time" | V1 == "date") #NU kan jag applicera den här arbetshästen
+# counttimers=length(timers)
+# #och det kan jag bara göra för att göra counts jag har ju tappat alla associationer inom datasetet
+# 
+# }
+
+headerrows <- head(headerproperties)
+typefilt=filter(headerrows, name == "type") #gör en dataframe av 
+typetrans <- data.table::transpose(typefilt) #transponera den. jo.
+types=unique(typetrans)
+
+typenames<- data.frame(matrix(ncol = 1, nrow = 6))
+typecount<- data.frame(matrix(ncol = 1, nrow = 6))
+
+for (a in 1:6) {
+
+  currenttype=types[a,1]
+  
+thistype=filter(typetrans, V1 == currenttype) #NU kan jagapplicera den här arbetshästen
+thistypetrans <- data.table::transpose(thistype) #transpose
+
+counttimes=length(thistypetrans)
+
+typenames[a,1]=currenttype
+typecount[a,1]=counttimes
+
+collect <- data.frame(typenames, typecount)
+
+}
+
+
+#fler kolumner
+#hur ifyllda är kolunerna?
+quantitativeTitco <- titcodata[, which((names(titcodata) %in% quantitativecolumnname)==TRUE)]
+qualitativeTitco <- titcodata[, which((names(titcodata) %in% qualitativecolumnname)==TRUE)]
+textTitco <- titcodata[, which((names(titcodata) %in% textcolumnname)==TRUE)]
+
+completionquant=1-sum(is.na(quantitativeTitco))/(nrow(titcodata)*length(quantitativecolumnname)) #completion rate quant
+completionquant=round(completionquant, digits = 2)
+completionqual=1-sum(is.na(qualitativeTitco))/(nrow(titcodata)*length(qualitativecolumnname))
+completionqual=round(completionqual, digits = 2)
+
+#för text får vi göra lite annorlunda:
[…]

### 2022-09-19 manuscript.Rmd
tabellerna med nu.
det gör lite ont i ögonen att titta på dem just nu tyvärr, formatering har varit prio 2 jag ville bara presentera datan från förra veckan för diskussionen idag 19/9

@@
 
-tab.1_cap <- table_nums(name = "tab_1", 
-                        caption = "Data types collected in the TITCO1 dataset")
+```{r}
+setwd("C:/Users/maria/Downloads/Packt Learning RStudio for R Statistical Computing 2012 RETAIL eBook-repackb00k/unmet-ICU-beds")
+load("table1_1.rdata")
 
-                     knitr::kable(table1_1, "simple")
+knitr::kable(table1_1, caption = "Data types collected in the TITCO1 dataset")
+```
 
@@
 
-tab.2_cap <- table_nums(name = "tab_2", 
-                        caption = "Means and standard devations of quantitative variables of ICU and non-ICU admitted patients, respectively")
+```{r}
+setwd("C:/Users/maria/Downloads/Packt Learning RStudio for R Statistical Computing 2012 RETAIL eBook-repackb00k/unmet-ICU-beds")
+load("table3.rdata")
 
-                     knitr::kable(table3, "simple"),
+knitr::kable(table3, caption = "Means and standard devations of quantitative variables of ICU and non-ICU admitted patients, respectively")
+```
 
@@
 
-tab.3_cap <- table_nums(name = "tab_3", 
-                        caption = "P- and t-values of the quantitative values in comparison of non-ICU and ICU patients")
+```{r}
+setwd("C:/Users/maria/Downloads/Packt Learning RStudio for R Statistical Computing 2012 RETAIL eBook-repackb00k/unmet-ICU-beds")
+load("statresult.rdata")
 
-                     knitr::kable(statresult, "simple"),
+knitr::kable(table3, caption = "P- and t-values of the quantitative values in comparison of non-ICU and ICU patients")
+```
 
@@
 If considerably more than a third of each qualitative parameter are admitted to the ICU, then that parameter is of interest.
-The proportions are presented in tab.4_cap <- table_nums(name = "tab_4",
-                        caption = "Proportion of patients admitted to the ICU for all variants of all quuantitative variables")
+The proportions are presented in
 
-                     knitr::kable(qualitativetable, "simple"),
+```{r}
+setwd("C:/Users/maria/Downloads/Packt Learning RStudio for R Statistical Computing 2012 RETAIL eBook-repackb00k/unmet-ICU-beds")
+load("qualitativetable.rdata")
+
+knitr::kable(qualitativetable, caption = "Proportion of patients admitted to the ICU for all variants of all quuantitative variables")
+```
 

### 2022-09-19 manuscript.Rmd
formiddagens version

@@
 The study is observational, the care provided to each patient has been evidence-based best available care.
-The data are standard details that would have been recorded anyway, and collection has not delayed or interfered with care nor added any cost.
+The data are standard details collected by a dedicated officer from the research team, and collection has not delayed or interfered with care.
 
@@
 
+COPYPASTE: One dedicated project officer collected data on site for approximately eight hours, five days per week.
+The shifts were rotated, so that the project officers worked morning, evening, and night shifts, and then had a day off.
+This way all possible shifts were covered during the course of a month.
+Data for patients arriving during the project officers shifts were collected by direct observation in the area where trauma patients were received, and the project officers were allowed to ask the health care staff for values of parameters not entered into the patient's records.
+For example, if the systolic blood pressure was measured (the project officer could see it being measured) but was not documented in the patient record, the project officer was allowed to ask for its value.
+Data for patients arriving to hospital outside the project officers' shifts were recorded from patient records.
+These patients were identified from the nurses' log books in the emergency department.
+
 Participants
@@
 Trauma covers injury and with a mechanism of road traffic, railway, fall, burn or assault.
+Patients with isolated limb injury and patients who were dead on arrival were not included.
 The patients represent mainly the lower stratum of socioeconomic stratum of their communities.
 
-47 patients were excluded for anonymization purposes, 120 patients were excluded because they were dead on arrival, and further XX pat were excluded because of missing data.
+47 patients were excluded for anonymization purposes,  Further XX pat were excluded because of missing data.
 
@@
 mortality, this bias would have to be addressed using some statistical tool, but it is not.
-The purpose of this thesis is use these same methods to assemble a control group but not for the purpose of examining the efficacy of ICU care on eg.
+The purpose of this thesis is use these same methods to assemble a control group not for the purpose of examining the efficacy of ICU care on eg.
 mortality, but to identify and quantify the subset of patients that should have been admitted to the ICU but weren't.
@@
 -------------------
+
+Predictor variables will be selected using both clinical judgement and data analysis.
+
+Clinical experience compels inclusion of vital parameters, irregardless if there is any statistical motivation. Such variables are hart rate, systolic blood pressure and respiration rate.
+
+The TITCO1 data set is analyzed mathemathically in pursuit of suitable predictor variables.
+The data set is divided into two subgroups, the cohort that were admitted to the ICU and the cohort that wasn't.
+In comparison between these subgroups, identify variables that distinguishes the former from the latter.
+
+Mean and standard deviation for the quantitative data is analysed for the ICU and non-ICU cohort respectively.
+p- and t-values when comparing these cohorts to show if there are any statistically significant differences.
+
+For the qualitative data, the propensity of each parameter to get a patient admitted to the ICU is in relation to the proportion of patients that were admitted to the ICU from the whole TITCO1 data set, about a third.
+If considerably more than a third of each qualitative parameter are admitted to the ICU, then that parameter is of interest.
+
+Some qualitative parameters such as intubation are recorded after both 1 and 24 hours, if applicable.
+The record at 1 hour has two options indicating the patient is intubated, "before arrival" meaning the patient was intubated at a their first health care facility and "yes" meaning the patient was intubated within the first hour.
+Because these two options both mean the patient was intubated after 1 hour, an aggregate parameter has been created for this, and a further parameter "intubated after 24 hours" that summarize the patients who were intubated on arrival, the patients who were intubated within the first hour and the patients who were intubated between 1 and 24 hours.
+
+FRÅGA: det finns pat som är intuberade "before arrival" men som inte är noterade som transferred från ett annat sjukhus.
+De har kommit med motor rickshaw.
+Är det rimligt att anta att de är transferred från ett sjukhus iallafall för vem har annars intuberat dessa patienter?
+
+The parameter "taken to the operation theatre" is a heterogenous qualitative parameter, where only asmall proportion of patients had been taken within the first hour but a substantial proportion of patients had been taken to the operation theatre after 24 hours.
+There is a need to consider the free text field operation type to filter out for operation types with higher propensity for ICU admission from operations with lower propensity.
+
+Results
+=======
+
 The quantitative data did not produce any clear results see table_nums('tab_2')
@@
 
-For the qualitative data, the propensity of each parameter to get a patient admitted to the ICU is in relation to the proportion of patients that were admitted to the ICU from the whole TITCO1 data set, about a third.
-If considerably more than a third of each qualitative parameter are admitted to the ICU, then that parameter is of interest.
-The proportions are presented in
+The proportions of the qualitative data are presented in
 
@@
 ```
-
-Results
-=======
 
@@
 of any potential bias
+-->
 
+In order to limit the scope of this thesis work, a few quite relevant factors could not be considered.
+One of the most important factors was temporality.
+Emergency rooms everywhere face significant variations in patient loads over the week or special occasions such as the Diwali festival.
+A different approach completely but to the same effect would have been to consider the ICU wards at the participating hospitals and chronologically determine which patients did not receive ICU care because these wards were at capacity.
+Though intuitively it would seem possible to do given the TITCO1 dataset, it would have been very difficult to do practically.
+
+Length of stay in hours is ICU care was recorded in the TITCO1 data set, but for the purpose of this thesis ICU admission has been reducedn to a binary 'yes' or 'no'.
+Because it is not uncommon to discontinue ICU care prematurely in favor of patients with more dire needs, the next step would be to consider such circumstances.
+
+<!--
 Interpretation

### 2022-09-19 manuscript.Rmd
inför möte

@@
 
-The parameter "taken to the operation theatre" is a heterogenous qualitative parameter, where only asmall proportion of patients had been taken within the first hour but a substantial proportion of patients had been taken to the operation theatre after 24 hours.
+<!-- 
+The parameter "taken to the operation theatre" is a heterogenous qualitative parameter, where only a small proportion of patients had been taken within the first hour but a substantial proportion of patients had been taken to the operation theatre after 24 hours.
 There is a need to consider the free text field operation type to filter out for operation types with higher propensity for ICU admission from operations with lower propensity.
+-->
+The parameter "taken to the operation theatre" is a heterogenous qualitative parameter, and operations performed in the first hour tend to be more acute than operations performed within the first day of hospitalization.
+There is a need to consider the free text field operation type to filter out for operation types with higher propensity for ICU admission from operations with lower propensity.
+
+The free hand text based information available contain information about examination findings, surgery types and ICD codes.
+Many of these data fields contain crucial information about patient condition, but also present a few challenges as described in chapter: Variables and data sources/measurements.
+These text based data fields will be examined for entries that are more prevalent in the ICU cohort than the non-ICU cohort, and combined with clinical experience to select suitable predictors.
 

### 2022-09-19 main.R

@@ -19,7 +19,11 @@
 ## manuscript file or run this file as a job, as that happens in a
 ## clean R session.
 #noacsr::source_all_functions()
+
+setwd("C:/Users/maria/Downloads/Packt Learning RStudio for R Statistical Computing 2012 RETAIL eBook-repackb00k/unmet-ICU-beds")
+
 library(tidyverse)
+library(dplyr)
 library(tableone)
 
 ## Import data
@@ -28,6 +32,7 @@ titcodata <- read.csv("titcodata.csv")
 #jag har manuellt ändrat och städat lite i dessa så jag sparar dem så jag kan committa dem:
 save(headerproperties, file = "headerproperties.RData")
 save(titcodata, file = "titcodata.RData")
+headerrows <- head(headerproperties)
 
 #FRÅGA: jag kan inte hitta hur man definierar eller filtrerar på row headers, men jag kan göra det här superjobbiga sättet
 typefilt=filter(headerrows, name == "type") #gör en dataframe av type-raden
@@ -40,6 +45,7 @@ counttimers=length(timers)
 load("quantitativecolumnname.rdata")
 load("qualitativecolumnname.rdata")
 load("textcolumnname.rdata")
+load("quantitativelong.rdata")
 
 #sortera bort DOA
 alive=filter(titcodata, incl != 2)
@@ -95,6 +101,12 @@ noICU_mean=myDf$mean
 noICU_sigma=myDf$sd
 myDfmeannolicu <- as.data.frame(noICU_mean)
 myDfsdnolicu <- as.data.frame(noICU_sigma)
+mean_diff=noICU_mean-ICU_mean
+mean_diffdf<- as.data.frame(mean_diff)
 
 #det här är mean och std av ICU vs no-ICU
-table2<- data.frame(df_t,myDfmeanlicu,myDfsdlicu,myDfmeannolicu,myDfsdnolicu)
+table3<- data.frame(df_t,myDfmeanlicu,myDfsdlicu,myDfmeannolicu,myDfsdnolicu, mean_diffdf)
+
+
+save(table3, file = "table3.RData")
+save(statresult, file = "statresult.RData")

### 2022-09-19 manuscript.Rmd
efter mötet

@@
 
+NOTE2SELF: för projektredogörelsen feta ut och ha med paragraferna här som temporaries och säg till när de är klara för kommentering intro och metoddelen, 
+Fråga Johanna: 3-5 sidor ja men det är en rekommendation och formatet med omfattande bakgrund passar inte den här typen av studie.
+jag kan men jag vet inte om det är meningsfullt att producera 2 sidor text baraföratt?
+I instruktionen skriver KI själva att det är en kontrollmekanism bara "koordinatorn får en uppfattning om progress i projektet,".
+Kan jag bara motivera utformningen så är jag inte orolig
 <!-- Explain the scientific background and rationale for the
@@
 6-8 lines long. Support every statement with a citation -->
+
+Background
+----------
 Trauma/Injury is a substantial global health problem that disproportionately affects low and middle-income countries (LMIC), resulting in significant mortality and morbidity@traumaLMIC.
@@
 
-Ethical considerations
-----------------------
-The data has been thoroughly anonymized.
-Each patient have been given an ID number, time frames have been shifted to make identification by admission and discharge or death dates impossible, and a further 47 patients have been omitted from the data set to shift the data frames and disable tracking.
-
-The study is observational, the care provided to each patient has been evidence-based best available care.
-The data are standard details collected by a dedicated officer from the research team, and collection has not delayed or interfered with care.
-
-FRÅGA: jag kan inte se att consent tagits?
-
 Aim
@@
 
-The theory being examined in this thesis is that there exist subset of the TITCO1 dataset that didn't go to the ICU, but are very similar to the patients that did go to the ICU, in the covariates that caused ICU-treated patients to receive this intervention.
+The theory being examined in this thesis is that there exist subset of the TITCO-I dataset that didn't go to the ICU, but are very similar to the patients that did go to the ICU, in the covariates that caused ICU-treated patients to receive this intervention.
 The thesis is that these patients were not admitted to the IC for lack of such resources.
@@
 periods of recruitment, exposure, follow-up, and data collection -->
-FRÅGA: kopierat rakt av härifrån https://github.com/titco/titco-I
-Men det är ju exakt vad som ska under den här rubriken. Får man göra så eller är det plagiat?
-The TITCO-I v1 dataset includes data on 16 000 patients admitted to four public university hospitals in urban India.
-The original dataset included 16 047 patients but 47 patients were deleted from the dataset for anonymization purposes, see the anonymization section for more details.
-At the time of the research Jai Prakash Narayan Apex Trauma Center (JPNATC), All India Institute of Medical Sciences, New Delhi, was a dedicated trauma centre with almost 180 beds.
-King Edward Memorial hospital (KEM), Mumbai, was a tertiary level hospital with no dedicated trauma ward.
-Lokmanya Tilak Municipal General Hospital (LTMGH), Mumbai was a tertiary level public university hospital with a dedicated trauma ward with 14 beds.
-The Institute of Post-Graduate Medical Education and Research and Seth Sukhlal Karnani Memorial Hospital (SSKM), Kolkata was a tertiary level public university hospital, but without a dedicated trauma ward.
-The cost of care in these hospitals was nominal.
-The first patient was enrolled in July 2013 and the last patient in December 2015.
-The exact enrollment period varied across participating hospitals.
+16 000 patients admitted to four public university hospitals in urban India are included in the TITCO-I v1 dataset.
+Initially the data set included 16047 patients, but to improve anonymization 47 patients were deleted from the dataset as described in the chapter anonymization.
+The participating hospitals were King Edward Memorial hospital (KEM) and Lokmanya Tilak Municipal General Hospital (LTMGH), Mumbai, Jai Prakash Narayan Apex Trauma Center (JPNATC), All India Institute of Medical Sciences, New Delhi and The Institute of Post-Graduate Medical Education and Research and Seth Sukhlal Karnani Memorial Hospital (SSKM), Kolkata.
+When the data was collected JPNATC had almost 180 beds in a dedicated trauma centre, KEM had no specific trauma ward and was a tertiary referral hospital, LTMGH had a specific trauma ward with 14 beds and was a tertiary referral public university hospital and SSKM had no specific trauma ward and was a public university tertiary referral hospital.
 
-COPYPASTE: One dedicated project officer collected data on site for approximately eight hours, five days per week.
-The shifts were rotated, so that the project officers worked morning, evening, and night shifts, and then had a day off.
-This way all possible shifts were covered during the course of a month.
-Data for patients arriving during the project officers shifts were collected by direct observation in the area where trauma patients were received, and the project officers were allowed to ask the health care staff for values of parameters not entered into the patient's records.
-For example, if the systolic blood pressure was measured (the project officer could see it being measured) but was not documented in the patient record, the project officer was allowed to ask for its value.
-Data for patients arriving to hospital outside the project officers' shifts were recorded from patient records.
-These patients were identified from the nurses' log books in the emergency department.
+The data was collected between July 2013 and December 2015 but the collection period for any individual hospitals did not cover this period.
+
+A dedicated project officer involved in the data collection was on site five days per week, approximately eight hours a day.
+The project officers covered night, evening and morning shifts in a way so all shift variants were covered every month of collection.
+Project officers worked in the area where trauma patients arrived to the hospital and collected data for patients arriving during their shift through direct observation.
+If the officer observed an examination being performed but not documented in the patient records were allowed to ask health care staff for the result.
+If a trauma patient arrived when a project officer was not on site, they were identified from the log books of the nurses on staff in the emergency department and their data was collected from patient records.
 
@@
 exposed and unexposed -->
-The TITCO1 data set includes all data recorded for trauma patients of the emergency departments at the participating hospitals.
+The eligibility criteria for the TITCO-I data set were patients who presented with trauma at the emergency departments of the participating hospitals.
 Trauma covers injury and with a mechanism of road traffic, railway, fall, burn or assault.
-Patients with isolated limb injury and patients who were dead on arrival were not included.
-The patients represent mainly the lower stratum of socioeconomic stratum of their communities.
-
-47 patients were excluded for anonymization purposes,  Further XX pat were excluded because of missing data.
+Patients who were dead on arrival were excluded, as were patients who only had isolated limb injuries.
+A majority of patients came from low socioeconomic circumstances in their respective communities.
 
@@
 sources and methods of selection of participants -->
+Ethical considerations
+----------------------
+Ethical approval for the collection of the data was obtained from the boards of ethical approval in all four hospitals.
+
+SSKM, LTMGH, KEM and JPNATC name of ethical review boards and approval reference numbers were IPGME&R Research Oversight Committee (IEC/279), Ethics Committee of the Staff and Research Society (IEC/11/13), Institutional Ethics Committee (IEC(I)/OUT/222/14), Institute Ethics Committee All India Institute of Medical Sciences (EC/NP-279/2013 RP-Ol/2013), respectively.
+
+The ethical review boards of all hospitals granted a waiver of informed consent.
+The trauma patients of in this study included patients in poor physiological, psychological and physical state, oftentimes unconscious or with reduced level of consciousness.
+The data collected for this study were merely routine data that would have been collected under normal circumstances and did not involve any additional examinations.
+The collection did not in any way interfere, delay or alter the care provided.
+Therefore the research team and the ethical review boards concluded that obtaining informed consent would only unneccesarily burden patients and/or family members.
+Personal details such as names, insurance details or contact information were not recorded.
+
+Anonymization
+-------------
+Hospital identification number
+Hospital name has been replaced with a identification number, and the conversion key has not been saved so there is not possible to reverse the conversion.
+Despite this, a high degree of anonymization can not be expected because the hospitals do have some distinguishable features and these are reflected in the data collected.
+For example,  70 % of the railroad accidents in the data set is in one of the hospitals, and one of the hosapitals is in close proximity to a big railroad hub.
+
+FRÅGA: Kan man och är det meningsfullt att analysera om sjukhus med dedikerade trauma wards får fler patienter körda till dem som sen hamnar på ICU?
+För när jag tittade lite på om man kunde luska ut vilket sjukhus som var vilket via just tran-parametern så var skillnaden tydlig.
+På ett sjukhus hamnar 96% av patienterna som transporterats från andra health facilities på IVA, och på ett annat sjukhus är det bara 4% och då gissar jag att det förra är ett sjukhus med en dedikerad trauma ward och det senare inte är det?
+det är vad det är för anonymiseringen det är lite begränsade möjligheter att anonymisera bra som jag illustrerar med järnvägsexemplet, men jag ställer frågan för om detta är en bias jag borde ta hänsyn till eller iallafall acknowledge ?
+Jag ställer i princip samma fråga i bias-kapitlet.
+
+Patient identification numbers
+Patient names were never recorded, patients were identified by a identification number in the original unencrypted data set.
+In the anonymization process the identification number has been replaced with another random identification number, and the conversion key has not been saved so there is not possible to reverse the conversion.
+
+Date observations
+Dates recorded in the TITCO-I data set were shifted using a random offset of somewhere between 0 and 5 weeks for each patient, but without losing any relevant information.
+The offset was designed to not dissipate details that could be of potential use for researchers such as day of week and rough indication of season.
+The internal relations between dates have been preserved so that the timing of all recordings and discharge in chronological relation to admission are all intact.
+
+It is important to note that all patients have sequence numbers that corresponded to the chronological order in which they arrived at any of the participating hospitals, and that sequence number has been kept intact.
+Keeping this sequence number variable intact allows for many useful operations such as examinations of trends in the data, while it is not possible to undo the date randomization by the sequence number variable, such an equation system is still underdetermined.
+Unfortunately this means it is also not possible to reconstruct the patient volumes at any given hospital on any given day and from that rate of resource engagement.
+In addition, 47 patient data were randomly deleted to prevent identification via the sequence number variable.
+
+Age
+Because very young and very old patients are so few, they have been grouped together to prevent identification via the age parameter.
+
+Fråga: man har buntat ihop 89+ arna för anonymisering, men 0-åringarna är bara 7 och ganska sårbara indentifieringsmässigt
+
+KOPPAT tills jag förstår och kan omformulera
+Injury descriptors and codes
+A random selection of 1% of injury descriptors entered into each of the variables e_1, e_2, e_3, e_4, e_5, e_6, e_7, e_8, e_9, e_10, e_11, e_12, xray_1, xray_2, xray_3, xray_4, xray_5, xray_6, xray_7, xray_8, xray_9, xray_10, xray_11, fast_1, fast_2, fast_3, fast_4, fast_5, fast_6, fast_7, fast_8, fast_9, fast_10, fast_11, ct_1, ct_2, ct_3, ct_4, ct_5, ct_6, ct_7, ct_8, ct_9, ct_10, ct_11, ct_12, ct_13, op_1, op_2, op_3, op_4, op_5, op_6, op_7, op_8, op_9, op_10, op_11 were set to missing.
+Corresponding ICD-10 codes were also set to missing.
+This was done to prevent identification of patients based on the list of recorded injuries.
+KOPPAT
 
@@
 
-knitr::kable(table1_1, caption = "Data types collected in the TITCO1 dataset")
+knitr::kable(table1_1, caption = "Data types collected in the TITCO-I dataset")
 ```
@@
 <!-- Describe any efforts to address potential sources of bias -->
-The TITCO1 dataset are observational data collected from a clinical environment and not an experimental setting.
+The TITCO-I dataset are observational data collected from a clinical environment and not an experimental setting.
 Medical interventions, examinations and ICU admission are not randomized, they are a consequence of a patients medical condition.
@@
 
-If the purpose of this thesis were to examine the efficacy of ICU care on eg.
-mortality, this bias would have to be addressed using some statistical tool, but it is not.
-The purpose of this thesis is use these same methods to assemble a control group not for the purpose of examining the efficacy of ICU care on eg.
-mortality, but to identify and quantify the subset of patients that should have been admitted to the ICU but weren't.
-
-Propensity Score Matching will be described in detail in chapter Statistical methods.
+Because the purpose of this thesis is to statistically assess how many patients that should have gone to the ICU but didn't, this particular selection bias does not skew the outcome and therefore does not have to be adressed.
 
 FRÅGA: Bör jag tänka på någon bias iallafall?
-Jag kan tänka mig att det finns skillnader mellan sjukhusen som bör undersökas (de har väl inte nödvändigtvis exakt samma proportioner ICU-sängar givet upptagningsområde, och att ett sjukhus därför kan vara mycket mer generös med ICU-sängarna än ett annat?), men vad för typ av statistisk bias är det isåfall?
+Jag kan tänka mig att det finns skillnader mellan sjukhusen som jag skriver lite om innan (de har väl inte nödvändigtvis exakt samma proportioner ICU-sängar givet upptagningsområde, och att ett sjukhus därför kan vara mycket mer generös med ICU-sängarna än ett annat?
+ICU admission rate skiljer sig ju en del och sjukhus med dedikerade trauma wards får fler svårt skakade patienter - inte minst ditkörda från andra sjukhus - som åker in på ICU, än sjukhus utan dessa resurser.
+en trauma patient med viss karakteristik påkörd utanför sjukhus A vs utanför sjukhus D verkar göra skillnad.), men vad för typ av statistisk bias är det isåfall?
 Det finns en selection bias i att det är lägre socioekonomisk stratum, men hur relevant är den?
+Och om detta gäller hela data setet så påverkar det väl mera hur generaliserbara resultaten är, snarare än introducerar en bias jag måste addressera?
 
 Propensity score matching in itself introduces a source of bias.
-The researcher chooses the predictor variables and so can introduce a self-fulfilling prophecy.
+The researcher chooses the predictor variables and so can introduce an overdetermined set and a self-fulfilling prophecy.
 For this reason the dataset shall be randomly split into two parts of 40% and 60% of the patients.
@@
 The data of 16047 trauma patients were collected at the participating hospitals over 2.5 years, as described in chapter Settings.
-47 entries have been randomly excluded for anonymization purposes, and for the purpose of this study the 120 patients who were dead on arrival have been excluded aswell.
+47 entries have been randomly excluded for anonymization purposes.
 
@@
 -->
-Fråga: hur diffar detta från vad som finns i Variables and data sources/measurements
+Fråga: hur diffar detta från vad som finns i Variables and data sources/measurements? Har jag missförstått rubriken variables eller den här rubriken?
 
@@
 If the correlation between the independent variables and the intervention and outcome isn't accounted for, a simple comparison between the cohort that did go to the ICU to the cohort that did not, would find that ICU care actually increased mortality.
-In the TITCO1 dataset the survival rate of patients who did not go to the ICU was 81% versus 68% for patients who did.
+In the TITCO-I dataset the survival rate of patients who did not go to the ICU was 81% versus 68% for patients who did.
 This is of course an incorrect conclusion, this data has treatment selection bias as patient covariates often confound treatment selection.
@@
 Second step is to calculate propensity scores for all patients.
-Each patient in the TITCO1 data set gets a propensity score based on their medical conditions and interventions, specifically their individual values of the predictor variables.
+Each patient in the TITCO-I data set gets a propensity score based on their medical conditions and interventions, specifically their individual values of the predictor variables.
 The propensity score determines the likelihood that a patient will receive ICU care, given their condition.
@@
 
-Clinical experience compels inclusion of vital parameters, irregardless if there is any statistical motivation. Such variables are hart rate, systolic blood pressure and respiration rate.
+Clinical experience compels inclusion of vital parameters, irregardless if there is any statistical motivation. Such variables are heart rate, systolic blood pressure and respiration rate.
 
-The TITCO1 data set is analyzed mathemathically in pursuit of suitable predictor variables.
+The TITCO-I data set is analyzed mathemathically in pursuit of suitable predictor variables.
 The data set is divided into two subgroups, the cohort that were admitted to the ICU and the cohort that wasn't.
@@
 
-For the qualitative data, the propensity of each parameter to get a patient admitted to the ICU is in relation to the proportion of patients that were admitted to the ICU from the whole TITCO1 data set, about a third.
+For the qualitative data, the propensity of each parameter to get a patient admitted to the ICU is in relation to the proportion of patients that were admitted to the ICU from the whole TITCO-I data set, about a third.
 If considerably more than a third of each qualitative parameter are admitted to the ICU, then that parameter is of interest.
@@
 There is a need to consider the free text field operation type to filter out for operation types with higher propensity for ICU admission from operations with lower propensity.
+FRÅGA: borde man ta operation within 24 hours och komplettera med text based operationstyp?
 
@@
 
-The quantitative data did not produce any clear results see table_nums('tab_2')
+The quantitative data results see table_nums('tab_2')
 
@@
 
-This is perhaps because of a crowd out effect.
+For many parameters the difference is small, perhaps because of a crowd out effect.
 For instance, the total Glascow coma scale score is high and quite similar for the non-ICU subset versus the ICU subset, the difference between the average of the groups is 2 points.
@@
 
+TASK4SELF: släng upp alla variabler i google sheetset
+.r fil för missing cleaning av titcodatasetet
+onsdag kolla kommentarerna löpande. 
+kolla på textbaserade parametrar
+släng upp alla tasks i slack, och instruera martin och johanna att kommentera i slack
 ```{r}
@@
 
-knitr::kable(table3, caption = "P- and t-values of the quantitative values in comparison of non-ICU and ICU patients")
+knitr::kable(statresult, caption = "P- and t-values of the quantitative values in comparison of non-ICU and ICU patients")
 ```
-
 FRÅGA: är p-värdena rimliga tycker du? de är superlåga och min känsla - med 5 hp statistik i ryggen - är att std-avvikelsen är för stor för det?
@@
 A different approach completely but to the same effect would have been to consider the ICU wards at the participating hospitals and chronologically determine which patients did not receive ICU care because these wards were at capacity.
-Though intuitively it would seem possible to do given the TITCO1 dataset, it would have been very difficult to do practically.
+Though intuitively it would seem possible to do given the TITCO-I dataset, it would have been very difficult to do practically.
 
-Length of stay in hours is ICU care was recorded in the TITCO1 data set, but for the purpose of this thesis ICU admission has been reducedn to a binary 'yes' or 'no'.
+Length of stay in hours is ICU care was recorded in the TITCO-I data set, but for the purpose of this thesis ICU admission has been reducedn to a binary 'yes' or 'no'.
 Because it is not uncommon to discontinue ICU care prematurely in favor of patients with more dire needs, the next step would be to consider such circumstances.

### 2022-09-20 manuscript.Rmd
morgonen 20/9 version för kommentering

@@
 I instruktionen skriver KI själva att det är en kontrollmekanism bara "koordinatorn får en uppfattning om progress i projektet,".
-Kan jag bara motivera utformningen så är jag inte orolig
+Kan jag bara motivera utformningen så är jag inte orolig jag la frågan till kursadmin
 <!-- Explain the scientific background and rationale for the
@@
 
-Existing TQIP often originate in high-income countries, and are either not applicable nor easily transferred to the hospital conditions of low and middle income countries (LMIC).
+Existing TQIP often originate in high-income countries (HIC), and are either not applicable nor easily transferred to the hospital conditions of low and middle income countries (LMIC) @frailty @gunshots  @onemilliondead.
+Applying TQIPs outside of their original context or even within a subset of it's own origin is difficult even without taking income bracket in account, for instance firearm violence is a considerable bigger portion of trauma patients in some US contexts, than the national average @gunshot.
+For trauma patients, there exist significant structural differences between HIC and LMIC mainly regarding injury mechanisms and severity, health care resources as well as patient make-up @frailty @indiantrauma @traumaLMIC @onemilliondead.
+These differences make a direct transferral of HIC TQIP models onto LMIC irrelevant or misleading and there is a need to develop TQIP assessment applicable to local conditions.
 This study aims to make an TQIP assessment in an Indian context to fill part of that knowledge gap, and to test whether there is merit in the testimony of the Indian colleagues.
@@
 
-The theory being examined in this thesis is that there exist subset of the TITCO-I dataset that didn't go to the ICU, but are very similar to the patients that did go to the ICU, in the covariates that caused ICU-treated patients to receive this intervention.
+The theory being examined in this masters thesis is that there exist subset of the TITCO-I dataset that didn't go to the ICU, but are very similar to the patients that did go to the ICU, in the covariates that caused ICU-treated patients to receive this intervention.
 The thesis is that these patients were not admitted to the IC for lack of such resources.
@@
 
-KOPPAT tills jag förstår och kan omformulera
-Injury descriptors and codes
-A random selection of 1% of injury descriptors entered into each of the variables e_1, e_2, e_3, e_4, e_5, e_6, e_7, e_8, e_9, e_10, e_11, e_12, xray_1, xray_2, xray_3, xray_4, xray_5, xray_6, xray_7, xray_8, xray_9, xray_10, xray_11, fast_1, fast_2, fast_3, fast_4, fast_5, fast_6, fast_7, fast_8, fast_9, fast_10, fast_11, ct_1, ct_2, ct_3, ct_4, ct_5, ct_6, ct_7, ct_8, ct_9, ct_10, ct_11, ct_12, ct_13, op_1, op_2, op_3, op_4, op_5, op_6, op_7, op_8, op_9, op_10, op_11 were set to missing.
-Corresponding ICD-10 codes were also set to missing.
-This was done to prevent identification of patients based on the list of recorded injuries.
-KOPPAT
+Text based input
+1% of the text-based inputs in ICD code fields and examination or operation findings were randomly deleted to obstruct identification by this variable.
 
@@
 P-values and T-values of non-ICU vs ICU patients:
-
-TASK4SELF: släng upp alla variabler i google sheetset
-.r fil för missing cleaning av titcodatasetet
-onsdag kolla kommentarerna löpande. 
-kolla på textbaserade parametrar
-släng upp alla tasks i slack, och instruera martin och johanna att kommentera i slack
 ```{r}

### 2022-09-20 manuscript.Rmd
20 sep kväll

@@
 NOTE2SELF: för projektredogörelsen feta ut och ha med paragraferna här som temporaries och säg till när de är klara för kommentering intro och metoddelen, 
-Fråga Johanna: 3-5 sidor ja men det är en rekommendation och formatet med omfattande bakgrund passar inte den här typen av studie.
-jag kan men jag vet inte om det är meningsfullt att producera 2 sidor text baraföratt?
+Fråga Johanna: 3-5 sidor ja men det är en rekommendation och formatet med omfattande bakgrund passar inte den här typen av studie och det blir ganska tydligt vad du menar med det när jag försöker producera texten.
+Mina kurskamraters projekt handlar om specifika sjukdomar eller medicinska interventioner eller liknande och då är det naturligt och viktigt att beskriva dem utförligt, men jag har inte riktigt något motsvarande min punchline är just i metoddelen.
+Jag skulle kunna skriva ännu mer om indisk traumastatistik, hur dåligt det är med trauma, recycla resonemanget om hur bl.a.
+motorcykelolyckor har olika karakteristik i Indien vs nordamerika men jag vet inte om det är meningsfullt att producera 2 sidor text baraföratt.
 I instruktionen skriver KI själva att det är en kontrollmekanism bara "koordinatorn får en uppfattning om progress i projektet,".
@@
 
-Trauma disproportionately affects young, otherwise healthy individuals.
-They are often the primary breadwinners for their families and therefore the detrimental impact of trauma go beyond mortality or substantial loss of disease-free years of life for the affected individuals, it also carries a considerable financial and social harm on their family members and communities.
+Trauma is a main cause of mortality and morbidity to otherwise healthy and young individuals, and dispropotionately affect this demographic.
+People in this demographic are also oftentimes the primary or sole breadwinners for their families.
+Therefore the detrimental impact of trauma go beyond mortality or substantial loss of disease-free years of life for the affected individuals, it also carries a considerable financial and social harm on their family members and communities.
 @traumaLMIC
@@
 The theory being examined in this masters thesis is that there exist subset of the TITCO-I dataset that didn't go to the ICU, but are very similar to the patients that did go to the ICU, in the covariates that caused ICU-treated patients to receive this intervention.
-The thesis is that these patients were not admitted to the IC for lack of such resources.
-These patients will be identified using propensity score matching.
+The thesis is that these patients were not admitted to the ICU only for lack of such resources, not by a medical assessment.
+In statistics, such a group would be akin to an non-experimental conrol group, and these patients will be identified using propensity score matching.
 
@@
 Unfortunately this means it is also not possible to reconstruct the patient volumes at any given hospital on any given day and from that rate of resource engagement.
+Nor is it possible to distinguish if patients who were made to wait for their first assessment were so because their injuries were deemed as less severe or if it was because health care staff was overwhelmed which is an important distinction.
+
 In addition, 47 patient data were randomly deleted to prevent identification via the sequence number variable.

### 2022-09-21 manuscript.Rmd
21 sep inga stora ändringar

@@
 ---
-title: "Quantifying unmet ICU beds in tertiary hospitals in urban India using Prospective Score Matching "
-output: html_document
+title: 'Quantifying unmet ICU beds in tertiary hospitals in urban India using Prospective
+  Score Matching '
 author: "Maria Eriksson"
+output:
+  word_document: default
+  html_document: default
 bibliography: bibliography.bib
@@
 Epidemiology (STROBE):explanation and elaboration.  -->
+
+Abbreviations
+============
+ER - Emergency room
+HIC - High-income countries
+LMIC - Low- and middle-income countries
 
@@
 
-NOTE2SELF: för projektredogörelsen feta ut och ha med paragraferna här som temporaries och säg till när de är klara för kommentering intro och metoddelen, 
-Fråga Johanna: 3-5 sidor ja men det är en rekommendation och formatet med omfattande bakgrund passar inte den här typen av studie och det blir ganska tydligt vad du menar med det när jag försöker producera texten.
-Mina kurskamraters projekt handlar om specifika sjukdomar eller medicinska interventioner eller liknande och då är det naturligt och viktigt att beskriva dem utförligt, men jag har inte riktigt något motsvarande min punchline är just i metoddelen.
-Jag skulle kunna skriva ännu mer om indisk traumastatistik, hur dåligt det är med trauma, recycla resonemanget om hur bl.a.
-motorcykelolyckor har olika karakteristik i Indien vs nordamerika men jag vet inte om det är meningsfullt att producera 2 sidor text baraföratt.
-I instruktionen skriver KI själva att det är en kontrollmekanism bara "koordinatorn får en uppfattning om progress i projektet,".
-Kan jag bara motivera utformningen så är jag inte orolig jag la frågan till kursadmin
+<!-- 
+FYI: utmanat volymkraven, de är motstridiga med kravet "Innehåller endast sådan information som är nödvändig för att förstå och motivera studiens syfte och metodik i stort"
+-->
+
 <!-- Explain the scientific background and rationale for the
@@
 ----------
-Trauma/Injury is a substantial global health problem that disproportionately affects low and middle-income countries (LMIC), resulting in significant mortality and morbidity@traumaLMIC.
+Trauma/Injury is a substantial global health problem that disproportionately affects low and middle-income countries (LMIC), resulting in significant mortality and morbidity @traumaLMIC.
 The annual mortality from injury in India amount to one million casualties @onemilliondead.
@@
 Health care professionals in many LMIC testify that their medical assessment oftentimes is that a trauma patient should be admitted to the ICU, but aren’t because such resources are scarce, overloaded and unavailable.
-Their medical judgement is that the lives of these patients were salvageable, and that these patients died as a direct result of not being admitted to the ICU@stuffLMICdoctorssay.
+Their medical judgement is that the lives of these patients were salvageable, and that these patients died as a direct result of not being admitted to the ICU @stuffLMICdoctorssay.
 Globally, an estimated two million lives could be saved annually, if the appropriate and timely trauma care provided by hospitals in many high-income countries were available and provided in low to-middle-income countries (LMICs) @livessaved
 
-In order to improve intra-hospital care, trauma quality improvement programs (TQIP) have been developed.
-The aim of TQIPs is to ensure equivalent care and improve trauma care by developing best practice guidelines@tqip.
+In order to improve and ensure equivalence of intra-hospital care, trauma quality improvement programs (TQIP) have been developed.
+TQIP assessments collect trauma data from participating health care providers and apply statistical analysis to the data.
+The aim is to find discrepancies in treatment for patients presenting with similar conditions.
+TQIPs aim to ensure equivalent care and improve trauma care by developing best practice guidelines @tqip, and to assert that health care providers provide the same care under same circumstances as their peers do.
 TQIPs utilizes a retrospective cohort study, analysing the current of trauma care patients and post-fact attempting to determine the appropriate level of care for each patient.
 From these analyses many aspects of trauma care can be improved and optimized, such as determining the appropriate amount of ICU beds and other trauma care resources @tqip.
+Fråga: rätt uppfattat? bygger dessa analyser alltid på propensity score matching approach ?
 
-Existing TQIP often originate in high-income countries (HIC), and are either not applicable nor easily transferred to the hospital conditions of low and middle income countries (LMIC) @frailty @gunshots  @onemilliondead.
+Knowlege gap
+------------
+Existing TQIP often originate in high-income countries (HIC), and are either not applicable nor easily transferred to the hospital conditions of low and middle income countries (LMIC) @frailty @gunshot  @onemilliondead.
 Applying TQIPs outside of their original context or even within a subset of it's own origin is difficult even without taking income bracket in account, for instance firearm violence is a considerable bigger portion of trauma patients in some US contexts, than the national average @gunshot.
-For trauma patients, there exist significant structural differences between HIC and LMIC mainly regarding injury mechanisms and severity, health care resources as well as patient make-up @frailty @indiantrauma @traumaLMIC @onemilliondead.
+For trauma patients, there exist significant structural differences between HIC and LMIC.
+These differences are in factors such as injury mechanisms and severity, health care resources as well as patient make-up @frailty @indiantrauma @traumaLMIC @onemilliondead.
+One example of this is the differences in helmet use for motorcyclists and their passengers, as well as differences of displacement in vehicle fleets in LMIC versus HIC.
+These factors have tremendous impact on the type and severity of injuries patients in the ER have, and subsequently on their care and outcome.
 These differences make a direct transferral of HIC TQIP models onto LMIC irrelevant or misleading and there is a need to develop TQIP assessment applicable to local conditions.
-This study aims to make an TQIP assessment in an Indian context to fill part of that knowledge gap, and to test whether there is merit in the testimony of the Indian colleagues.
-The study shall propose a scope of the unmet need for ICU beds for trauma patients at tertiary hospitals in urban India.
-If this analysis indeed indicate a shortage, it can be used to motivate extending ICU resources and reduce preventable trauma morbidity and mortality.
 
@@
 People in this demographic are also oftentimes the primary or sole breadwinners for their families.
-Therefore the detrimental impact of trauma go beyond mortality or substantial loss of disease-free years of life for the affected individuals, it also carries a considerable financial and social harm on their family members and communities.
-@traumaLMIC
+Therefore the detrimental impact of trauma go beyond mortality or substantial loss of disease-free years of life for the affected individuals, it also carries a considerable financial and social harm on their family members and communities @traumaLMIC.
+.
 
 Aim
------------
-Predict the unmet need for ICU beds for trauma patients at tertiary hospitals in urban India by applying propensity score matching.
+----
+Predict the unmet need for ICU beds for trauma patients at tertiary hospitals in urban India.
+
+This study aims to make an TQIP assessment in an Indian context to fill part of the knowledge gap, and to test whether there is merit in the testimony of the Indian colleagues.
+The TQIP assessment shall propose a scope of the unmet need for ICU beds for trauma patients at tertiary hospitals in urban India.
+If this analysis indeed indicate a shortage, it can be used to motivate extending ICU resources and reduce preventable trauma morbidity and mortality.
 
@@
 The participating hospitals were King Edward Memorial hospital (KEM) and Lokmanya Tilak Municipal General Hospital (LTMGH), Mumbai, Jai Prakash Narayan Apex Trauma Center (JPNATC), All India Institute of Medical Sciences, New Delhi and The Institute of Post-Graduate Medical Education and Research and Seth Sukhlal Karnani Memorial Hospital (SSKM), Kolkata.
-When the data was collected JPNATC had almost 180 beds in a dedicated trauma centre, KEM had no specific trauma ward and was a tertiary referral hospital, LTMGH had a specific trauma ward with 14 beds and was a tertiary referral public university hospital and SSKM had no specific trauma ward and was a public university tertiary referral hospital.
+When the data was collected all hospitals were tertiary hospitals.
+JPNATC had almost 180 beds in a dedicated trauma centre, KEM had no specific trauma ward, LTMGH was a public university hospital with a specific trauma ward with 14 beds and SSKM was a  public university hospital with no specific trauma ward.
 
-The data was collected between July 2013 and December 2015 but the collection period for any individual hospitals did not cover this period.
+The data was collected between July 2013 and December 2015 but the collection period for any individual hospitals did not cover necessarily this whole period.
 
@@
 också finns det ex obs där numeriska värdet för spO2 saknas men ändå data yes/no om pat fått syrgas och jag vet inte om sån data är meningsfull?
-FRÅGA: vad betyder GCS verbal V1?
-det är inte felregistrerat det är ett alternativ i codebook.
-jag har hittat ett förslag om att det är pat som inte kan ge verbalt svar pga intub eller surgical airway, men det finns pat som har varken eller och GCS verbalt = 1v
 

### 2022-09-21 manuscript.Rmd
21 kväll

@@
 ---
-title: 'Quantifying unmet ICU beds in tertiary hospitals in urban India using Prospective
+title: 'Quantifying unmet ICU beds in tertiary hospitals in urban India using Propensity
   Score Matching '
@@
 output:
+  html_document: default
   word_document: default
-  html_document: default
 bibliography: bibliography.bib
@@
 Trauma/Injury is a substantial global health problem that disproportionately affects low and middle-income countries (LMIC), resulting in significant mortality and morbidity @traumaLMIC.
+Despite having only 1% of the global vehicle fleet, a disproportionate 10% of global traffic casualties occur in India @indiantrauma.
 The annual mortality from injury in India amount to one million casualties @onemilliondead.
 Half of trauma mortality in India occur in-hospital, and in-hospital trauma mortality in Indian hospitals was double that of high-income countries @indiantrauma.
+
 Health care professionals in many LMIC testify that their medical assessment oftentimes is that a trauma patient should be admitted to the ICU, but aren’t because such resources are scarce, overloaded and unavailable.
@@
 In comparison between these subgroups, identify variables that distinguishes the former from the latter.
+The quantitative and qualitative properties were considered separately.
 
@@
 For instance, the total Glascow coma scale score is high and quite similar for the non-ICU subset versus the ICU subset, the difference between the average of the groups is 2 points.
-However, considered individually, low scores in either category is a high propensity to admit a patient to the ICU, as demonstrated in the comparison of quantitative parameters.
+However, low scores in either category is a high propensity to admit a patient to the ICU, as demonstrated in the comparison of quantitative parameters.
 

### 2022-09-21 manuscript.Rmd
Comment on intro and methods

@@
 ----------
-Trauma/Injury is a substantial global health problem that disproportionately affects low and middle-income countries (LMIC), resulting in significant mortality and morbidity @traumaLMIC.
+Trauma/Injury<!-- I suggest not using these interchangeably, as they're different things.
+I would define trauma as physcial injury and the body's response to that injury --> is a substantial global health problem that disproportionately affects low and middle-income countries (LMIC), resulting in significant mortality and morbidity @traumaLMIC.
 Despite having only 1% of the global vehicle fleet, a disproportionate 10% of global traffic casualties occur in India @indiantrauma.
@@
 
-Health care professionals in many LMIC testify that their medical assessment oftentimes is that a trauma patient should be admitted to the ICU, but aren’t because such resources are scarce, overloaded and unavailable.
+Health care professionals in many LMIC testify that their medical assessment oftentimes is that a trauma patient should be admitted to the ICU, but are not<!-- Avoid contractions in academic writing--> because such resources are scarce, overloaded and unavailable.
 Their medical judgement is that the lives of these patients were salvageable, and that these patients died as a direct result of not being admitted to the ICU @stuffLMICdoctorssay.
@@
 
-In order to improve and ensure equivalence of intra-hospital care, trauma quality improvement programs (TQIP) have been developed.
+<!-- You can definitely write more about the health care system in India, the trauma system, the availability of ICUs, how common ICU care is in trauma, why trauma patients need ICU care, how long they stay, outcomes etc -->
+
+In order to improve and ensure equivalence <!-- Equivalence how?
+Across hospitals?
+Within trauma systems?-->of intra-hospital care, trauma quality improvement programs (TQIP) have been developed.
 TQIP assessments collect trauma data from participating health care providers and apply statistical analysis to the data.
@@
 Fråga: rätt uppfattat? bygger dessa analyser alltid på propensity score matching approach ?
+<!-- I'm not sure if this last paragraph is relevant to your work, but you can keep it for now.
+TQIP is a broad term that one hand includes all kinds of processes to improve the quality of trauma care, such as mortality and morbidity conferences, using audit filters, comparing risk adjusted outcomes over time and between centres, on the other hand is a term that the American College of Surgeons use for their specific TQIP package, which is then much more closely defined.
+So no, I would say that these analyses rarely involves propensity score matching -->
 
@@
 
+<!-- I would rephrase to "Quantify the unmet need..." -->
+
+<!-- The three paragraphs below can be removed, as the information presented is not really part of the aim. Some of it can be moved to introduction and some to methods -->
 This study aims to make an TQIP assessment in an Indian context to fill part of the knowledge gap, and to test whether there is merit in the testimony of the Indian colleagues.
@@
 
+<!-- I would not focus too much on the TQIP aspect of this, instead I would focus on the likely shortage of ICU care in this setting and that there are few methods to quantify this -->
+
 Methods 
@@
 <!-- Present key elements of study design -->
-Observational retrospective cohort study from data collected at four tertiary hospitals in urban India.
+Observational retrospective cohort study using data from the Towards Improved Trauma Care Outcomes (TITCO) in India cohort.
 
@@
 periods of recruitment, exposure, follow-up, and data collection -->
-16 000 patients admitted to four public university hospitals in urban India are included in the TITCO-I v1 dataset.
+16 000 patients admitted to four public university hospitals in urban India are included in the TITCO dataset.
 Initially the data set included 16047 patients, but to improve anonymization 47 patients were deleted from the dataset as described in the chapter anonymization.
@@
 exposed and unexposed -->
-The eligibility criteria for the TITCO-I data set were patients who presented with trauma at the emergency departments of the participating hospitals.
+The eligibility criteria for the TITCO dataset were patients who presented with trauma at the emergency departments of the participating hospitals.
 Trauma covers injury and with a mechanism of road traffic, railway, fall, burn or assault.
 Patients who were dead on arrival were excluded, as were patients who only had isolated limb injuries.
-A majority of patients came from low socioeconomic circumstances in their respective communities.
+A majority of patients came from low socioeconomic circumstances in their respective communities. <!-- This last sentence can be moved to Setting -->
 
@@
 -------------
+
+<!-- You don't have to include all this in the thesis or a paper, definitely enough if you only provide a brief summary or cite the information on GitHub -->
 Hospital identification number
@@
 Jag ställer i princip samma fråga i bias-kapitlet.
+
+<!-- I don't think you should include centre in your propensity score, because as you say one centre had almost exclusively ICU beds for these patients, so that variable would almost perfectly discriminate ICU care.
+-->
 
@@
 
-Fråga: man har buntat ihop 89+ arna för anonymisering, men 0-åringarna är bara 7 och ganska sårbara indentifieringsmässigt
+Fråga: man har buntat ihop 89+ arna för anonymisering, men 0-åringarna är bara 7 och ganska sårbara indentifieringsmässigt<!-- Good point.
+However age was recorded in months and that information was removed.
+Also, you don't know when, in time, they were admitted.
+-->
 
@@
 <!-- Describe any efforts to address potential sources of bias -->
-The TITCO-I dataset are observational data collected from a clinical environment and not an experimental setting.
+The TITCO dataset are observational data collected from a clinical environment and not an experimental setting.
 Medical interventions, examinations and ICU admission are not randomized, they are a consequence of a patients medical condition.
@@
 
-Because the purpose of this thesis is to statistically assess how many patients that should have gone to the ICU but didn't, this particular selection bias does not skew the outcome and therefore does not have to be adressed.
+Because the purpose of this thesis is to statistically assess how many patients that should have gone to the ICU but did not, this particular selection bias does not skew the outcome and therefore does not have to be adressed.
 
@@
 
+<!-- You can definitely discuss the bias introduced by the hospitals being different -->
+
 Propensity score matching in itself introduces a source of bias.
@@
 The 40% shall be used to produce the propensity score matching, and the 60% shall be used to verify it.
+<!-- The other way around, 60% to build the model, 40% to assess unmet need -->
 Fråga: typ?
@@
 The data of 16047 trauma patients were collected at the participating hospitals over 2.5 years, as described in chapter Settings.
-47 entries have been randomly excluded for anonymization purposes.
+47 entries have been randomly excluded for anonymization purposes.<!-- I would say that it's enough to state that you will use all 16000 observations in the dataset -->
 
-Återkom: XX patients were discarded because of missing data
+Återkom: XX patients were discarded because of missing data <!-- This information should be in Results, not Methods -->
 
@@
 -->
-Fråga: hur diffar detta från vad som finns i Variables and data sources/measurements? Har jag missförstått rubriken variables eller den här rubriken?
+Fråga: hur diffar detta från vad som finns i Variables and data sources/measurements? Har jag missförstått rubriken variables eller den här rubriken? <!-- You can remove this -->
 
@@
 There is a need to offset or balance the effect of this causality, but a randomized case-control study could be impractical or immoral.
+<!-- I don't think the example below helps and I would consider removing it.
+Focus instead on describing the process of developing the propensity score: building a model that estimates the probability of receiving ICU care and then using this probability to identify observations with similar probabilites but different treatment -->
 
 In this context, it could be to examine the efficacy of ICU care on survival.
-If the correlation between the independent variables and the intervention and outcome isn't accounted for, a simple comparison between the cohort that did go to the ICU to the cohort that did not, would find that ICU care actually increased mortality.
-In the TITCO-I dataset the survival rate of patients who did not go to the ICU was 81% versus 68% for patients who did.
+If the correlation between the independent variables and the intervention and outcome is not accounted for, a simple comparison between the cohort that did go to the ICU to the cohort that did not, would find that ICU care actually increased mortality.
+In the TITCO dataset the survival rate of patients who did not go to the ICU was 81% versus 68% for patients who did.
 This is of course an incorrect conclusion, this data has treatment selection bias as patient covariates often confound treatment selection.
@@
 In this context it means to first identify the conditions and medical interventions received that correlate best to being offered ICU care.
-All properties of the patients that were sent to the ICU will be compared to the patents who didn't, in order to identify properties that distinguishes this cohort.
+All properties of the patients that were sent to the ICU will be compared to the patents who did not, in order to identify properties that distinguishes this cohort.
 
-Second step is to calculate propensity scores for all patients.
-Each patient in the TITCO-I data set gets a propensity score based on their medical conditions and interventions, specifically their individual values of the predictor variables.
+The second step is to calculate propensity scores for all patients.
+Each patient in the TITCO data set gets a propensity score based on their medical conditions and interventions, specifically their individual values of the predictor variables.
 The propensity score determines the likelihood that a patient will receive ICU care, given their condition.
@@
 PSM is a standardized mathematical approach to identify members of the subset that did not get the intervention but had the same predictors as the subgroup that did get the intervention.
-The purpose of the propensity score matching is identifying patients that weren’t admitted into the ICU, but presented with a set of predictors - a propensity score - very similar to those that were admitted.
+The purpose of the propensity score matching is identifying patients that were not admitted to the ICU, but presented with a set of predictors - a propensity score - very similar to those that were admitted.
+
+<!-- I would not match on the variable values but on the probability, estimated using for example logistic regression. So:
+
+1. Randomly split the data into a training (60%) and test sample (40%)
+2. Select predictors using the training data
+3. Build a multivariable logistic regression model using the training data with ICU care as the outcome and the predictors from step 1 as independent variables.
+4. Use the multivariable model to estimate the probability of receiving ICU care in the test data and match patients using this probability
+5. Compare the number of patients who received ICU care with the number who didn't
 
 (b) Describe any methods used to examine subgroups and interactions
+
+-->
 

### 2022-09-22 manuscript.Rmd
intermediär version

--- 2022-09-21 manuscript.Rmd
@@
 ============
+ACS - American college of surgeons
 ER - Emergency room
@@
 LMIC - Low- and middle-income countries
+PSM - propensity score matching
+RCT - Randomized case-control study
+TITCO -Towards Improved Trauma Care Outcomes 
+TQIP - Trauma Care Improvement Program
 
@@
 
-<!-- 
-FYI: utmanat volymkraven, de är motstridiga med kravet "Innehåller endast sådan information som är nödvändig för att förstå och motivera studiens syfte och metodik i stort"
--->
-
 <!-- Explain the scientific background and rationale for the
@@
 
-Background
-----------
-Trauma/Injury is a substantial global health problem that disproportionately affects low and middle-income countries (LMIC), resulting in significant mortality and morbidity @traumaLMIC.
+Trauma globally and in India
+----------------------------
+Trauma is a substantial global health problem that disproportionately affects low and middle-income countries (LMIC), resulting in significant mortality and morbidity @traumaLMIC.
 Despite having only 1% of the global vehicle fleet, a disproportionate 10% of global traffic casualties occur in India @indiantrauma.
@@
 
-Health care professionals in many LMIC testify that their medical assessment oftentimes is that a trauma patient should be admitted to the ICU, but aren’t because such resources are scarce, overloaded and unavailable.
+Mortality due to accidents in India amounted to 421 104 people in 2019, which is 31.5 per 100.000 inhabitants.
+The traumatic mechanisms of injury from accidents covered in this thesis are road accidents, falls, assault and burns and these make up about half of the accident mortality @nationalcrime.
+
+Trauma is a main cause of mortality and morbidity to otherwise healthy and young individuals, and dispropotionately affect this demographic.
+Most casualties affected the age group 30-45 years, followed by the age bracket 18-30 years @nationalcrime.
+People in this demographic are also oftentimes the primary or sole breadwinners for their families.
+Therefore the detrimental impact of trauma go beyond mortality or substantial loss of disease-free years of life for the affected individuals, it also carries a considerable financial and social harm on their family members and communities @traumaLMIC @nationalcrime
+
+Health care professionals in many LMIC testify that their medical assessment oftentimes is that a trauma patient should be admitted to the ICU, but are not because such resources are scarce, overloaded and unavailable.
 Their medical judgement is that the lives of these patients were salvageable, and that these patients died as a direct result of not being admitted to the ICU @stuffLMICdoctorssay.
@@
 
+A diverse mix of public, for-profit and not-for-profit private actors provide health services in India.
+According to a WHO 2022 report, India has 76 709 non-ventilator supported and 39 476 ventilator-supported ICU beds @ICUbeds.
+This is 8.2 per 100 000 inhabitants.
+The WHO report also highlight the probable shortage of ICU resources in India, though in the context of the COVID-19 pandemic.
+
+Methods to examine trauma care and ensure equivalence
+-----------------------------------------------------
 In order to improve and ensure equivalence of intra-hospital care, trauma quality improvement programs (TQIP) have been developed.
-TQIP assessments collect trauma data from participating health care providers and apply statistical analysis to the data.
+The pilot and most notable example is the TQIP developed by the American College of Surgeons ACS.
+Since the implementation of the ACS pilot, TQIP has evolved to a broad term that include many kinds of processes to improve the quality of trauma care, such as mortality and morbidity conferences, using audit filters or comparing risk adjusted outcomes over time and between centres.
 The aim is to find discrepancies in treatment for patients presenting with similar conditions.
-TQIPs aim to ensure equivalent care and improve trauma care by developing best practice guidelines @tqip, and to assert that health care providers provide the same care under same circumstances as their peers do.
-TQIPs utilizes a retrospective cohort study, analysing the current of trauma care patients and post-fact attempting to determine the appropriate level of care for each patient.
+TQIPs aim to ensure patients get equivalent care within and across hospitals, to benchmark performance of health care providers versus their peers and to improve trauma care by developing best practice guidelines @tqip.
+
+One aspect of TQIP assessments is to collect trauma data from participating health care providers and apply statistical analysis to the data.
+From this data, TQIPs utilizes a retrospective cohort study, analysing the current of trauma care patients and post-fact attempting to determine the appropriate level of care for each patient.
 From these analyses many aspects of trauma care can be improved and optimized, such as determining the appropriate amount of ICU beds and other trauma care resources @tqip.
-Fråga: rätt uppfattat? bygger dessa analyser alltid på propensity score matching approach ?
 
-Knowlege gap
-------------
-Existing TQIP often originate in high-income countries (HIC), and are either not applicable nor easily transferred to the hospital conditions of low and middle income countries (LMIC) @frailty @gunshot  @onemilliondead.
-Applying TQIPs outside of their original context or even within a subset of it's own origin is difficult even without taking income bracket in account, for instance firearm violence is a considerable bigger portion of trauma patients in some US contexts, than the national average @gunshot.
+It is tempting to apply existing models of TQIP assessments on Indian trauma data to identify causes of the observed disparities in trauma care performance in HIC and LMIC, and propose measures to harmonize performance.
+Unfortunately as TQIP often originate in high-income countries (HIC), they are either not applicable nor easily transferred to the hospital conditions of LMIC @frailty @gunshot  @onemilliondead.
+Applying TQIPs outside of their original context or even within a subset of its own origin is difficult even without taking income bracket in account, for instance firearm violence is a considerable bigger portion of trauma patients in some US contexts, than the national average @gunshot.
 For trauma patients, there exist significant structural differences between HIC and LMIC.
@@
 
-Trauma is a main cause of mortality and morbidity to otherwise healthy and young individuals, and dispropotionately affect this demographic.
-People in this demographic are also oftentimes the primary or sole breadwinners for their families.
-Therefore the detrimental impact of trauma go beyond mortality or substantial loss of disease-free years of life for the affected individuals, it also carries a considerable financial and social harm on their family members and communities @traumaLMIC.
-.
+Kommentar på Mgerdin kommentar: jag snodde din definition här.
+Jag håller med om att detta inte är den mest relevanta informationen men för "kontrollmekanism"-arbetet som ska in på måndag här har KI en checklista och de vill ha 3-5 sidor bakgrund trots att vi jobbar med max 1-1.5 så jag måste feta ut.
+Detsamma för kommentaren i Aim jag är medveten om att det är felplacerat eller irrelevant.
+
+Knowlege gap
+------------
+There are compelling reasons to believe there is a shortage of ICU beds in India, but the scope of this shortage and the potential for reduction of trauma mortality if this shortage would be remedied is unknown.
+
+WHO and local health care workers believe there is a shortage of ICU beds in Indian hospitals, and that the lives of many patients would have been salvageable if the appropriate ICU resources had been available.
+The differences of in-hospital trauma mortality in Indian hospitals was double that of high-income countriesin, and part of this difference could be explained by a shortage of ICU beds in India @indiantrauma
 
@@
 ----
-Predict the unmet need for ICU beds for trauma patients at tertiary hospitals in urban India.
 
-This study aims to make an TQIP assessment in an Indian context to fill part of the knowledge gap, and to test whether there is merit in the testimony of the Indian colleagues.
-The TQIP assessment shall propose a scope of the unmet need for ICU beds for trauma patients at tertiary hospitals in urban India.
-If this analysis indeed indicate a shortage, it can be used to motivate extending ICU resources and reduce preventable trauma morbidity and mortality.
+Quantify the unmet need for ICU beds in tertiaty hospitals in urban India.
 
-The theory being examined in this masters thesis is that there exist subset of the TITCO-I dataset that didn't go to the ICU, but are very similar to the patients that did go to the ICU, in the covariates that caused ICU-treated patients to receive this intervention.
-The thesis is that these patients were not admitted to the ICU only for lack of such resources, not by a medical assessment.
-In statistics, such a group would be akin to an non-experimental conrol group, and these patients will be identified using propensity score matching.
-
-The size of this group in relation to the group that were admitted to the ICU indicate the magnitude of the unmet need of ICU beds.
-If, for every patient that did receive ICU care, there was another patient with similar predictors that wasn’t admitted due to scarcity of ICU resources, that would mean the ICU resources are 50% of what they ought to be.
+Apply statistical analysis to a large data set of Indian trauma patients and determine how many out of the patients that were not admitted to the ICU reasonably should have been admitted given their condition.
+Their numbers in proportion to the number of ICU-admitted patients is indicative of the relative shortage of ICU beds.
 
@@
 periods of recruitment, exposure, follow-up, and data collection -->
-16 000 patients admitted to four public university hospitals in urban India are included in the TITCO-I v1 dataset.
+16 000 patients admitted to four public university hospitals in urban India are included in the TITCO dataset @titco.
 Initially the data set included 16047 patients, but to improve anonymization 47 patients were deleted from the dataset as described in the chapter anonymization.
@@
 JPNATC had almost 180 beds in a dedicated trauma centre, KEM had no specific trauma ward, LTMGH was a public university hospital with a specific trauma ward with 14 beds and SSKM was a  public university hospital with no specific trauma ward.
+
+The participating hospitals are all public entities, so a majority of patients came from low socioeconomic circumstances in their respective communities.
 
@@
 exposed and unexposed -->
-The eligibility criteria for the TITCO-I data set were patients who presented with trauma at the emergency departments of the participating hospitals.
-Trauma covers injury and with a mechanism of road traffic, railway, fall, burn or assault.
+The eligibility criteria for TITCO were patients who presented with injuries from trauma at the emergency departments of the participating hospitals with a mechanism of road traffic, railway, fall, burn or assault.
 Patients who were dead on arrival were excluded, as were patients who only had isolated limb injuries.
-A majority of patients came from low socioeconomic circumstances in their respective communities.
 
@@
 ----------------------
-Ethical approval for the collection of the data was obtained from the boards of ethical approval in all four hospitals.
+
+This study does not requlre ethical permission, it refers to data collected and shared openly.
+
+Ethical approval for the collection of the data used was obtained from the boards of ethical approval in all four hospitals.
 
@@
 For example,  70 % of the railroad accidents in the data set is in one of the hospitals, and one of the hosapitals is in close proximity to a big railroad hub.
-
-FRÅGA: Kan man och är det meningsfullt att analysera om sjukhus med dedikerade trauma wards får fler patienter körda till dem som sen hamnar på ICU?
-För när jag tittade lite på om man kunde luska ut vilket sjukhus som var vilket via just tran-parametern så var skillnaden tydlig.
-På ett sjukhus hamnar 96% av patienterna som transporterats från andra health facilities på IVA, och på ett annat sjukhus är det bara 4% och då gissar jag att det förra är ett sjukhus med en dedikerad trauma ward och det senare inte är det?
-det är vad det är för anonymiseringen det är lite begränsade möjligheter att anonymisera bra som jag illustrerar med järnvägsexemplet, men jag ställer frågan för om detta är en bias jag borde ta hänsyn till eller iallafall acknowledge ?
-Jag ställer i princip samma fråga i bias-kapitlet.
 
@@
 Age
-Because very young and very old patients are so few, they have been grouped together to prevent identification via the age parameter.
-
-Fråga: man har buntat ihop 89+ arna för anonymisering, men 0-åringarna är bara 7 och ganska sårbara indentifieringsmässigt
+Very old patients are few, they have been grouped together to prevent identification via the age parameter.
 
@@
 
-Because the purpose of this thesis is to statistically assess how many patients that should have gone to the ICU but didn't, this particular selection bias does not skew the outcome and therefore does not have to be adressed.
+There exist a potential bias in the routines, resources and patient makeup of the participating hospitals.
+Becasue they are all tertiary hospitals, about two thirds of the patients have been referred from other health care providers, meaning that there is a selection bias in the patient populations.
+Further, hospitals with dedicated trauma wards have a propensity to receive transferred patients or police escorted patients with worse condition and prognosis.
 
-FRÅGA: Bör jag tänka på någon bias iallafall?
-Jag kan tänka mig att det finns skillnader mellan sjukhusen som jag skriver lite om innan (de har väl inte nödvändigtvis exakt samma proportioner ICU-sängar givet upptagningsområde, och att ett sjukhus därför kan vara mycket mer generös med ICU-sängarna än ett annat?
-ICU admission rate skiljer sig ju en del och sjukhus med dedikerade trauma wards får fler svårt skakade patienter - inte minst ditkörda från andra sjukhus - som åker in på ICU, än sjukhus utan dessa resurser.
-en trauma patient med viss karakteristik påkörd utanför sjukhus A vs utanför sjukhus D verkar göra skillnad.), men vad för typ av statistisk bias är det isåfall?
-Det finns en selection bias i att det är lägre socioekonomisk stratum, men hur relevant är den?
-Och om detta gäller hela data setet så påverkar det väl mera hur generaliserbara resultaten är, snarare än introducerar en bias jag måste addressera?
+As stated in chapter Settings, most patients have low socioeconomic status.
+Low socioeconomic status is associated with higher level of malnutrition and comorbidities and would have a negtive effect on vitals and the capacity of patients to tolerate certain interventions.
+This circumstance doesn't affect the analysis and internal validity, but reduce the generalizability of the results.
+<!-- (ÅTERKOM: har inte råd att refereas?) -->
 
-Propensity score matching in itself introduces a source of bias.
-The researcher chooses the predictor variables and so can introduce an overdetermined set and a self-fulfilling prophecy.
+Because the purpose of this thesis is to statistically assess how many patients that should have gone to the ICU but did not, these particular selection biases are not expected to skew the outcome, and therefore does not have to be adressed.
+
+Hospitals with trauma wards could have lower admission criteria to the ICU than hospitals with no trauma ward.
+Because examining admission criteria to the ICU is part of the analysis it may be necessary to examine and address this bias by comparing ICU admitted patients between hospitals.
+Propensity score matching can be used to identify differences in ICU admission criteria between hospitals and find patients that were admitted but need not have been.
+These patients are excluded from predictor selection to reduce this bias.
+This is part of the development and training of the PSM model.
+
+Propensity score matching in itself introduces a source of bias from unfit assumptions.
+The terminology and potential sources of bias are explained chapter Study Design.
+In choosing predictor variables the researcher could introduce an overdetermined set and a self-fulfilling prophecy.
 For this reason the dataset shall be randomly split into two parts of 40% and 60% of the patients.
-The 40% shall be used to produce the propensity score matching, and the 60% shall be used to verify it.
-Fråga: typ?
+The 60% shall be used to produce the propensity score matching, and the remaining 40% shall be used to verify it.
 
@@
 <!-- Explain how the study size was arrived at -->
-The data of 16047 trauma patients were collected at the participating hospitals over 2.5 years, as described in chapter Settings.
-47 entries have been randomly excluded for anonymization purposes.
-
-Återkom: XX patients were discarded because of missing data
-
-Quantitative variables
-----------------------
-
-<!-- Explain how quantitative variables were handled in the
-analyses. If applicable, describe which groupings were chosen and why
--->
-Fråga: hur diffar detta från vad som finns i Variables and data sources/measurements? Har jag missförstått rubriken variables eller den här rubriken?
+The data of 16000 trauma patients were collected at the participating hospitals over 2.5 years, as described in chapter Settings.
 
@@
 -------------------
-
 <!--
@@
 -->
-Propensity score matching will be used in this analysis.
-Propensity score matching (PSM) is best explained using one typical application: identifying a control group when examining the efficacy of an intervention for a specific outcome, where there is causality between the independent variables for the intervention, and sometimes also the outcome.
-There is a need to offset or balance the effect of this causality, but a randomized case-control study could be impractical or immoral.
+Propensity score matching (PSM) is useful method for identifying and assembling a control group from within a data set where there is causality between the independent and dependent variables, and a randomized case-control (RCT) study to offset or balance this bias is impractical, immoral or otherwise improper.
+Propensity score is a statistical variable, understood as the likelihood of a subject receiving treatment conditional on observations of relevant covariates of that subject before treatment.
+In RCTs it is known from study design, but it is generally unknown in observational data.
+The rationale for PSM is that while it is unknown, this variable can be fairly well estimated using logistic regression.
 
-In this context, it could be to examine the efficacy of ICU care on survival.
-If the correlation between the independent variables and the intervention and outcome isn't accounted for, a simple comparison between the cohort that did go to the ICU to the cohort that did not, would find that ICU care actually increased mortality.
-In the TITCO-I dataset the survival rate of patients who did not go to the ICU was 81% versus 68% for patients who did.
-This is of course an incorrect conclusion, this data has treatment selection bias as patient covariates often confound treatment selection.
-The higher mortality observed in patients who received ICU care is because patients were selected for the intervention because they presented certain medical conditions that also carry a high mortality - even with the intervention ICU care.
-Producing a randomized control trial study for these types of lifesaving medical interventions for trauma patients would obviously go against ethical guidelines of research, and another approach to produce a control group is necessary.
-Propensity score matching is a statistical procedure for reducing this bias by assembling a sample from within the data set, in which confounding factors are balanced between the cohort that got the intervention and the cohort that did not.
-<!-- 
-The goal of propensity score matching is usually to identify suitable control subjects that did not receive an intervention, even though they fairly well met the same inclusion criteria that the subjects that did receive the intervention had.
-Control subjects who are similar enough to the intervention group in the most relevant aspects for receiving treatment.
--->
-The size of this subset in relation to the subset of patients that did receive ICU care is the end point for this analysis.
-This proportion is indicative of the unmet need of ICU beds, as described in chapter Aim.
-In the typical PSM appliction, this subset is used as the control group to determine the efficacy of ICU care for some outcome, eg mortality.
+The theory being examined in this masters thesis is that there exist a subset of the TITCO dataset that did not go to the ICU, but are very similar to the patients that did go to the ICU, in the covariates that caused ICU-treated patients to receive this intervention.
+The thesis is that these patients were not admitted to the ICU only for lack of such resources, not by a medical assessment.
 
-The first step is to identify predictor variables: covariates most predictive of the dependent variable.
-In this context it means to first identify the conditions and medical interventions received that correlate best to being offered ICU care.
-All properties of the patients that were sent to the ICU will be compared to the patents who didn't, in order to identify properties that distinguishes this cohort.
+Translated into this context this means that in this data set exist a selection bias, there is causality between the independent variables medical conditions and interventions and the dependent variable ICU admittance.
+Patients were admitted to the ICU as a consequence of their medical condition.
+Performing a randomized case-control study and randomly assigning or withholding a live-saving treatment would obviously be immoral, but we believe that in the TITCO data set exist patients who were not admitted into the ICU by volition but for scarcity of such resources.
+Such a group would be akin to an non-experimental control group and PSM can be used to identify and assemble it.
+If such a subset can be identified by PSM, then it naturally follows that there exists shortage of ICU beds in tertiary hospitals in India.
+It also follows that the shortage can be quantified by the proportionality of this subset versus the subset that did receive ICU care.
+If, for every patient that did receive ICU care, there was another patient with similar predictors that wasn’t admitted due to scarcity of ICU resources, that would mean the ICU resources are 50% of what they ought to be.
 
-Second step is to calculate propensity scores for all patients.
-Each patient in the TITCO-I data set gets a propensity score based on their medical conditions and interventions, specifically their individual values of the predictor variables.
+The propensity score model is developed using the Training Data and tested using the Validation Data.
+The workflow contain a series of steps and the work flow should understood as linear, but iterative.
+The output of each step is the input for later steps.
+Therefore the output of the last step reflect the cumulative error from unfit assumptions made the second and fourth step.
+The output is tested against Validation Data.
+By reverting back and adjusting assumptions in an iterative process, the model is continuously improved and the accuracy and validity is improved.
+
+The first step of PSM is to randomly split the data set into two parts, 60% and 40%.
+The 60% part is a Training Part used to delvelop and train the PSM, and the 40% part is the Validation Part.
+This is done because the inherent bias of the PSM method described in chapter Bias.
+Training Data is used in steps 2-4 and the output is then validated using Validation Data.
+
+Second step is to identify predictor variables: covariates most predictive of the dependent variable, as described in chapter Predictor selection.
+In this context it means to first identify the conditions and medical interventions received that correlate best to being admitted to ICU care.
+This selection is based on clinical experience as well as analysis of the Training Data.
+In the data analysis all properties of the patients that were sent to the ICU will be compared to the patents who did not, in order to identify properties that distinguishes this cohort., Familiarity with the collection of source data is crucial to minimize the risk to choose irrelevant predictors and introduce error.
+
+Third step is to build a standard multivariable logistic regression model with ICU care as the outcome and the predictors selected in the second step as independent variables.
+This model is used to calculate the standard variable propensity score for all patients.
 The propensity score determines the likelihood that a patient will receive ICU care, given their condition.
 
-Third step is PSM.
-PSM is a standardized mathematical approach to identify members of the subset that did not get the intervention but had the same predictors as the subgroup that did get the intervention.
-The purpose of the propensity score matching is identifying patients that weren’t admitted into the ICU, but presented with a set of predictors - a propensity score - very similar to those that were admitted.
+Fourth step is PSM.
+In PSM members of the subset that did not get ICU care but had similar predictors - and therefore similar propensity scores - as the subgroup that did get ICU care, are matched to patients who did go to the ICU.
+There are a few different ways to perform this matching but they are all imperfect because propensity scores are unique and perfect matching is unattainable There are control mechanisms and limitations but a risk of applying too fix or too loose matching criteria the outcome could inadvertedly vastly under- or overestimate the  need of ICU beds remain.
+
+The same procedure is repeated on the Validation Data using the same predictors and the same matching criteria, and this should produce a cohort comparable to the one trom Training Data.
 
@@
 
-The two subgroups were patients admitted to the ICU, and the patients who weren't. Admission to the iCU is a binary yes or no, length of stay has not been taken into Consideration.
+The two subgroups were patients admitted to the ICU, and the patients who were not. Admission to the iCU is a binary yes or no, length of stay has not been taken into Consideration.
 
@@
 Data that are missing are assumed missing because care for a patient did not motivate the collection of the data.
-For example, if serum creatinine was not recorded, it is assumed that the collection of this variable wasn't relevant for care of the patient.
+For example, if serum creatinine was not recorded, it is assumed that the collection of this variable was not relevant for care of the patient.
 
@@
 All quantitative data from the original dataset that were entered as zero but where a zero value is physiologically impossible were replaced with NA values.
-Zero values for variables where it is possible to have zero values, such as respiration rate and heart rate were left as zeroes to not tamper with the data, although a zero value indicate that the variable probably wasn't measured.
+Zero values for variables where it is possible to have zero values, such as respiration rate and heart rate were left as zeroes to not tamper with the data, although a zero value indicate that the variable probably was not measured.
 These amount to 250 and 155 entries respectively for the first observation within an hour and less than 100 for both for the second observation after 24 h, so the choice to include or exclude these observations won’t skew the data considerably.
@@
 The TITCO-I data set is analyzed mathemathically in pursuit of suitable predictor variables.
-The data set is divided into two subgroups, the cohort that were admitted to the ICU and the cohort that wasn't.
+The data set is divided into two subgroups, the cohort that were admitted to the ICU and the cohort that was not.
 In comparison between these subgroups, identify variables that distinguishes the former from the latter.
@@
 
+Missing data
+------------
+Återkom: XX patients were discarded because of missing data
+
+Predictor selection
+-------------------
 The quantitative data results see table_nums('tab_2')

### 2022-09-23 main.R
commit

@@ -20,7 +20,9 @@
 ## clean R session.
 #noacsr::source_all_functions()
 
-setwd("C:/Users/maria/Downloads/Packt Learning RStudio for R Statistical Computing 2012 RETAIL eBook-repackb00k/unmet-ICU-beds")
+#setwd("C:/Users/maria/Downloads/Packt Learning RStudio for R Statistical Computing 2012 RETAIL eBook-repackb00k/unmet-ICU-beds")
+
+setwd("C:/Users/maria/Downloads/unmet-ICU-beds")
 
 library(tidyverse)
 library(dplyr)
@@ -80,7 +82,7 @@ for (a in 1:columncount) {
   
   
 }
-
+#ändra något
 
 df_t <- data.table::transpose(quantitativelong)
 

### 2022-09-23 manuscript.Rmd
main efter revidering

@@
 
+<!-- 
+FYI: utmanat volymkraven, de är motstridiga med kravet "Innehåller endast sådan information som är nödvändig för att förstå och motivera studiens syfte och metodik i stort"
+-->
+
 <!-- Explain the scientific background and rationale for the
@@
 This is 8.2 per 100 000 inhabitants.
-The WHO report also highlight the probable shortage of ICU resources in India, though in the context of the COVID-19 pandemic.
+This same report also highlight the probable shortage of ICU resources in India, though in the context of the COVID-19 pandemic.
 

### 2022-09-23 manuscript.Rmd
Update manuscript.Rmd
kväll 23 sep
pre pull

--- 2022-09-22 manuscript.Rmd
@@
 
-A diverse mix of public, for-profit and not-for-profit private actors provide health services in India.
+A diverse mix of public, for-profit and not-for-profit private actors provide health services in India, where private providers dominate.
 According to a WHO 2022 report, India has 76 709 non-ventilator supported and 39 476 ventilator-supported ICU beds @ICUbeds.
-This is 8.2 per 100 000 inhabitants.
-The WHO report also highlight the probable shortage of ICU resources in India, though in the context of the COVID-19 pandemic.
 
@@
 
-One aspect of TQIP assessments is to collect trauma data from participating health care providers and apply statistical analysis to the data.
-From this data, TQIPs utilizes a retrospective cohort study, analysing the current of trauma care patients and post-fact attempting to determine the appropriate level of care for each patient.
-From these analyses many aspects of trauma care can be improved and optimized, such as determining the appropriate amount of ICU beds and other trauma care resources @tqip.
-
-It is tempting to apply existing models of TQIP assessments on Indian trauma data to identify causes of the observed disparities in trauma care performance in HIC and LMIC, and propose measures to harmonize performance.
-Unfortunately as TQIP often originate in high-income countries (HIC), they are either not applicable nor easily transferred to the hospital conditions of LMIC @frailty @gunshot  @onemilliondead.
-Applying TQIPs outside of their original context or even within a subset of its own origin is difficult even without taking income bracket in account, for instance firearm violence is a considerable bigger portion of trauma patients in some US contexts, than the national average @gunshot.
+TQIP often originate in high-income countries (HIC), they are either not applicable nor easily transferred to the hospital conditions of LMIC @frailty @gunshot  @onemilliondead.
 For trauma patients, there exist significant structural differences between HIC and LMIC.
 These differences are in factors such as injury mechanisms and severity, health care resources as well as patient make-up @frailty @indiantrauma @traumaLMIC @onemilliondead.
-One example of this is the differences in helmet use for motorcyclists and their passengers, as well as differences of displacement in vehicle fleets in LMIC versus HIC.
-These factors have tremendous impact on the type and severity of injuries patients in the ER have, and subsequently on their care and outcome.
 These differences make a direct transferral of HIC TQIP models onto LMIC irrelevant or misleading and there is a need to develop TQIP assessment applicable to local conditions.
-
-Kommentar på Mgerdin kommentar: jag snodde din definition här.
-Jag håller med om att detta inte är den mest relevanta informationen men för "kontrollmekanism"-arbetet som ska in på måndag här har KI en checklista och de vill ha 3-5 sidor bakgrund trots att vi jobbar med max 1-1.5 så jag måste feta ut.
-Detsamma för kommentaren i Aim jag är medveten om att det är felplacerat eller irrelevant.
 
@@
 ------------
-There are compelling reasons to believe there is a shortage of ICU beds in India, but the scope of this shortage and the potential for reduction of trauma mortality if this shortage would be remedied is unknown.
+There are compelling reasons to believe there is a shortage of ICU beds in India, but the scope of this shortage as well as the potential for reduction of trauma mortality if this shortage would be remedied is unknown.
 
-WHO and local health care workers believe there is a shortage of ICU beds in Indian hospitals, and that the lives of many patients would have been salvageable if the appropriate ICU resources had been available.
-The differences of in-hospital trauma mortality in Indian hospitals was double that of high-income countriesin, and part of this difference could be explained by a shortage of ICU beds in India @indiantrauma
+WHO and local health care workers believe there is a shortage of ICU beds in Indian hospitals, and that the lives of many patients would have been salvageable if the appropriate ICU resources had been available @ICUbeds.
+The differences of in-hospital trauma mortality in Indian hospitals was double that of high-income countries, and part of this difference could be explained by a shortage of ICU beds in India @indiantrauma .
 
@@
 All vital parameters are considered mandatory, irregardless of patient condition so patients without data of heart rate, systolic blood pressure or respiration rate at any time of their hospital stay, have been discarded.
-FRÅGA: right?
 

### 2022-09-26 clean.R
datatvätt

@@ -1,0 +1,135 @@
+library(noacsr)
+library(tidyverse)
+library(dplyr)
+setwd("C:/Users/maria/Downloads/Packt Learning RStudio for R Statistical Computing 2012 RETAIL eBook-repackb00k/unmet-ICU-beds")
+
+headerproperties <- read.csv("headerpropertiestrans.csv") #från rad 70 har datan fär "type" hamnat i "note"-kolumnen. det har jag löst manuellt
+titcodata <- read.csv("titcodatafull.csv") #untampered version
+
+
+#filter
+#ur headerprops, filtrera fram alla rader i type-kolumnen som är är c("qualitative", "quantitative")), och skapa en array av motsvarande rader i name-kolumnen
+# variables <- headerproperties %>% filter(type %in% c("qualitative", "quantitative")) %>% select(name)
+#name är header-raden i titco
+# # Välj kolumner ur datat baserat på listan av variabler
+#name-arrayen variables kan du nu applicera på titcodatan, och välja alla kolumner som finns nämnd i arrayen. variables$name byter objekttyp till chr 1:56 som måste till. all_of gör inget
+# titcodata %>% select(all_of(variables$name))
+
+#remove patients who died between arrival and admission
+liveonadmission<- titcodata %>% filter(incl %in% c("0","1", "3"))
+#skriv över
+
+titcodata=liveonadmission
+
+#ersätt alla fysiologiskt orimliga nollor i kvantitativa datat med NA
+#steg ett vilka är det?
+cantbezero=data.frame("hb","hc","sc","bun") #välj manuellt ut vilka kvawnt variabler som inte någonsin kan vara noll Hb, serumkreatinin, hematokrit, blod urea nitrogen
+cantbezerotrans<-data.table::transpose(cantbezero) #transponera
+colnames(cantbezerotrans)<-data.frame("name") #sätt dit headern manuellt så den fattar
+
+#filtrera fram motsvarande kolumner
+cantebezeromat=titcodata %>% select(all_of(cantbezerotrans$name))
+cantebezeromat[cantebezeromat==0] = NA
+#och skriv över. kolumn för kolumn det borde finnas ett lättare sätt?
[…]

### 2022-09-26 textfiles.R

@@ -1,0 +1,64 @@
+setwd("C:/Users/maria/Downloads/Packt Learning RStudio for R Statistical Computing 2012 RETAIL eBook-repackb00k/unmet-ICU-beds")
+
+#setwd("C:/Users/maria/Downloads/unmet-ICU-beds")
+
+library(tidyverse)
+library(dplyr)
+library(tableone)
+
+#build strings for the ICU codes
+
+resultmatrix = data.frame(matrix(, nrow = 3, ncol = 11))
+resultmatrixe = data.frame(matrix(, nrow = 1, ncol = 12))
+resultmatrixct = data.frame(matrix(, nrow = 1, ncol = 13))
+resultarray = data.frame(matrix(, nrow = 1, ncol = 11))
+
+#bygg först en liten dataframe med alla ICD koder
+#nej vi måste ha tre pga om jag vill använda dem sen för att filtrera ut kolumner accepterar inte r missing values och de är tre olika längder
+#jag behöver bygga en array inte en matris av resultaten här annars vill det inte förstå, så
+f=1 #counter
+uniquexraytot= data.frame(matrix(, nrow = 11*200, ncol =1 )) #resultatmatris, extra stor
+for (a in 1:11) {
+xray=paste("xray_",a,"_icd",sep="")
+fast=paste("fast_",a,"_icd",sep="")
+op=paste("op_",a,"_icd",sep="")
+
+resultmatrix[1,a]=xray
+resultmatrix[2,a]=fast
+resultmatrix[3,a]=op
+
+#vilka unika har vi?
+uniquexray=unique(titcodata[xray])
+uniquexraytot[f:nrow(uniquexray),1]<- data.frame(uniquexray) #data.frame(uniquexray[1:nrow(uniquexray),1])
+f=f+nrow(uniquexray)
+#DU VAR HÄR
+
+}
+
+for (a in 1:12) {
+  e=paste("e_",a,"_icd",sep="")
+  resultmatrixe[1,a]=e
+}
+
+for (a in 1:12) {
+  ct=paste("ct_",a,"_icd",sep="")
+  resultmatrixct[1,a]=ct
+}
+# 
+# resultmatrix[1,12]="e_12_icd"
+# resultmatrix[4,12]="ct_12_icd"
+# resultmatrix[4,13]="ct_12_icd"
+
+resultarray=resultmatrix[1,1:11]
+
+resultarraytrans<-data.table::transpose(resultarray) #transponera
[…]

### 2022-09-26 textfiles.R
icd kod

@@ -6,19 +6,25 @@ library(tidyverse)
 library(dplyr)
 library(tableone)
 
+titcodata <- read.csv("titcodatafull.csv")
+load("licu.rdata")
+titcodata=licu #fult
+
 #build strings for the ICU codes
 
 resultmatrix = data.frame(matrix(, nrow = 3, ncol = 11))
-resultmatrixe = data.frame(matrix(, nrow = 1, ncol = 12))
+resultmatrixext = data.frame(matrix(, nrow = 1, ncol = 12))
 resultmatrixct = data.frame(matrix(, nrow = 1, ncol = 13))
 resultarray = data.frame(matrix(, nrow = 1, ncol = 11))
 
 #bygg först en liten dataframe med alla ICD koder
 #nej vi måste ha tre pga om jag vill använda dem sen för att filtrera ut kolumner accepterar inte r missing values och de är tre olika längder
 #jag behöver bygga en array inte en matris av resultaten här annars vill det inte förstå, så
-f=1 #counter
-uniquexraytot= data.frame(matrix(, nrow = 11*200, ncol =1 )) #resultatmatris, extra stor
+rowno=1 #counter
+loop=0
+uniqueoptot= data.frame(matrix(, nrow = 11*200, ncol =1 )) #resultatmatris, extra stor
 for (a in 1:11) {
+  loop=loop+1
 xray=paste("xray_",a,"_icd",sep="")
 fast=paste("fast_",a,"_icd",sep="")
 op=paste("op_",a,"_icd",sep="")
@@ -28,35 +34,127 @@ resultmatrix[2,a]=fast
 resultmatrix[3,a]=op
 
 #vilka unika har vi?
-uniquexray=unique(titcodata[xray])
-uniquexraytot[f:nrow(uniquexray),1]<- data.frame(uniquexray) #data.frame(uniquexray[1:nrow(uniquexray),1])
-f=f+nrow(uniquexray)
+uniqueop=unique(titcodata[op])
+startat=rowno
+endat=rowno+nrow(uniqueop)-1
+howmanyrows=nrow(uniqueop)
+uniqueoptot[startat:endat,1]<- data.frame(matrix(uniqueop[1:howmanyrows,1])) #data.frame(uniquexray[1:nrow(uniquexray),1])
+rowno=rowno+nrow(uniqueop)
 #DU VAR HÄR
 
 }
 
+uniqueexttot= data.frame(matrix(, nrow = 11*200, ncol =1 ))
 for (a in 1:12) {
[…]

### 2022-09-27 textfiles.R
icd koder fishing expedition

@@ -145,6 +145,7 @@ extsort=print(extdf[order(extdf$Freq, decreasing = TRUE), ]   )
 
 present <- data.frame(xraysort[1:50,1:2],fastsort[1:50,1:2],opsort[1:50,1:2],ctsort[1:50,1:2],extsort[1:50,1:2])
 
+
 write.csv(present,"present.csv")
 #okej nu har vi en matris med alla x-ray-kolumner
 #borde kunna räkna nu

### 2022-09-27 manuscript.Rmd
originaldatan untampered

@@
 output:
+  pdf_document: default
   html_document: default
@@
 ACS - American college of surgeons
+
 ER - Emergency room
+
 HIC - High-income countries
+
 LMIC - Low- and middle-income countries
+
 PSM - propensity score matching
+
 RCT - Randomized case-control study
+
 TITCO -Towards Improved Trauma Care Outcomes
+
 TQIP - Trauma Care Improvement Program
@@
 Trauma is a substantial global health problem that disproportionately affects low and middle-income countries (LMIC), resulting in significant mortality and morbidity @traumaLMIC.
+4.8 million people die from trauma annually, 90% are in LMIC and 20% in India alone @onemilliondead.
 Despite having only 1% of the global vehicle fleet, a disproportionate 10% of global traffic casualties occur in India @indiantrauma.
-The annual mortality from injury in India amount to one million casualties @onemilliondead.
-Half of trauma mortality in India occur in-hospital, and in-hospital trauma mortality in Indian hospitals was double that of high-income countries @indiantrauma.
+Half of trauma mortality in India occur in-hospital.
+There is no national trauma registry in India, but isolated studies have shown that in-hospital mortality within 30 days of hospitalization is higher than similarly injured patients in HIC hospitals @indiantrauma.
+Globally, an estimated two million lives could be saved annually, if the appropriate and timely trauma care provided by hospitals in many high-income countries were available and provided in low to-middle-income countries (LMICs) @indiantrauma.
 
-Mortality due to accidents in India amounted to 421 104 people in 2019, which is 31.5 per 100.000 inhabitants.
-The traumatic mechanisms of injury from accidents covered in this thesis are road accidents, falls, assault and burns and these make up about half of the accident mortality @nationalcrime.
+Health care professionals in many LMIC testify that their medical assessment oftentimes is that a trauma patient should be admitted to the ICU, but are not because such resources are scarce, overloaded and unavailable.
+Their medical judgement is that the lives of these patients were salvageable, and that these patients died as a direct result of not being admitted to the ICU @stuffLMICdoctorssay.
 
@@
 
-Health care professionals in many LMIC testify that their medical assessment oftentimes is that a trauma patient should be admitted to the ICU, but are not because such resources are scarce, overloaded and unavailable.
-Their medical judgement is that the lives of these patients were salvageable, and that these patients died as a direct result of not being admitted to the ICU @stuffLMICdoctorssay.
-Globally, an estimated two million lives could be saved annually, if the appropriate and timely trauma care provided by hospitals in many high-income countries were available and provided in low to-middle-income countries (LMICs) @livessaved
-
 A diverse mix of public, for-profit and not-for-profit private actors provide health services in India, where private providers dominate.
 According to a WHO 2022 report, India has 76 709 non-ventilator supported and 39 476 ventilator-supported ICU beds @ICUbeds.
+Because of the type and severity of injuries, ICU resources are often essential in care of trauma patients @titco.
+
+<!-- KommentarJohanna: I suggest adding more details about the context in India, the health care system, how common trauma is in India, how many ICU beds there are, why ICU care is important for the trauma population and so on..
+-->
+<!-- KommentarMartin:You can definitely write more about the health care system in India, the trauma system, the availability of ICUs, how common ICU care is in trauma, why trauma patients need ICU care, how long they stay, outcomes etc -->
+
+<!-- Respons: Ja det är magert men det finns en orsak till det, jag har undvikit detta av två skäl
+ 1.Datan som står att finna om ICU vård resurserna i Indien är lite spretig, men enligt WHO så är den runt 8 per 100.000 och andra trovärdiga bedömare är där nånstans +- 50%.
+Det är i paritet med Sverige?
+Förklaringen står kanske att finna i att definitionen av en ICU bed inte är vedertagen men trots mycket läsande innantill har jag inte kunnat avfärda eller förklara detta.
+2.
+pkt 1 lyfter en fråga som ökar komplexiteten: det finns lite olika levels av ICU beds, vilken är det vi menar?
+Eftersom statistiken på ICU resurser är dåligt differentierad och TITCO datasetet inte specificerar det, ska frågan undvikas för att inte öka komplexiteten för sig själv?
+-->
 
@@
 The aim is to find discrepancies in treatment for patients presenting with similar conditions.
+This difference in in-hospital mortality is likely multi-factorial, and any factors have been proposed to explain it such as differences in triage and delay-of-care, as well as shortage of hospital resources including ICU beds @onemilliondead ,@traumaLMIC .
 TQIPs aim to ensure patients get equivalent care within and across hospitals, to benchmark performance of health care providers versus their peers and to improve trauma care by developing best practice guidelines @tqip.
@@
 These differences make a direct transferral of HIC TQIP models onto LMIC irrelevant or misleading and there is a need to develop TQIP assessment applicable to local conditions.
+
+<!-- Ni vil båda flytta fokus från detta, och jag förstår inte riktigt varför det är med kopplingen är luddig.
+Är det för att säga: det finns TQIPs som ger en fingervisning om hur ICU ska dimensioneras, bestyckas, bemannas och bedrivas Men det finns inga färdiga TQIPs som kan användas för att kvantifiera ICU-bristen i Indien och vi tror det är vanskligt att använda de befintliga eftersom de skapats under så olika förhållanden.
+Och därför måste vi börja från noll?
+-->
 
@@
 sources and methods of selection of participants -->
+
 Ethical considerations
@@
 
-This study does not requlre ethical permission, it refers to data collected and shared openly.
+This study does not require ethical permission, it refers to data collected and shared openly.
 
@@
 Hospital identification number
+
 Hospital name has been replaced with a identification number, and the conversion key has not been saved so there is not possible to reverse the conversion.
@@
 Patient identification numbers
+
 Patient names were never recorded, patients were identified by a identification number in the original unencrypted data set.
@@
 Date observations
+
 Dates recorded in the TITCO-I data set were shifted using a random offset of somewhere between 0 and 5 weeks for each patient, but without losing any relevant information.
@@
 
+Age
+
+Very old patients are few, they have been grouped together to prevent identification via the age parameter.
+
 In addition, 47 patient data were randomly deleted to prevent identification via the sequence number variable.
-
-Age
-Very old patients are few, they have been grouped together to prevent identification via the age parameter.
 
@@
 
-datatvätt:
-FRÅGA: quantitativa variablerna har många NA som jag tror egentligen är NO, ex om pat har haft syrgas vid spO2 mätning.
-kan jag tänka lite fritt här när det känns uppenbart eller ska jag förankra?
-också finns det ex obs där numeriska värdet för spO2 saknas men ändå data yes/no om pat fått syrgas och jag vet inte om sån data är meningsfull?
-
 The quantitative data include recorded parameters such as vital parameters, lab results, GCS total score and cumulative time in surgery or ICU, if applicable.
@@
 The TITCO dataset are observational data collected from a clinical environment and not an experimental setting.
-Medical interventions, examinations and ICU admission are not randomized, they are a consequence of a patients medical condition.
+Medical interventions, examinations and ICU admission are not randomized, they are a consequence of a patient's medical condition.
 There exists a causal effect between the independent and the dependent variables: a patient is sent to the ICU precisely because of the severity of their condition.
 
-There exist a potential bias in the routines, resources and patient makeup of the participating hospitals.
-Becasue they are all tertiary hospitals, about two thirds of the patients have been referred from other health care providers, meaning that there is a selection bias in the patient populations.
-Further, hospitals with dedicated trauma wards have a propensity to receive transferred patients or police escorted patients with worse condition and prognosis.
+There exists a potential bias in the routines, resources and patient makeup of the participating hospitals.
+Because they are all tertiary hospitals, about two thirds of the patients have been referred from other health care providers, meaning that there is a selection bias in the patient populations.
+Further, hospitals with dedicated trauma wards have a propensity to receive transferred or police escorted patients with worse condition and prognosis.
 
 As stated in chapter Settings, most patients have low socioeconomic status.
-Low socioeconomic status is associated with higher level of malnutrition and comorbidities and would have a negtive effect on vitals and the capacity of patients to tolerate certain interventions.
+Low socioeconomic status is associated with higher level of malnutrition and comorbidities and would have a negative effect on vitals and the capacity of patients to tolerate certain interventions.
 This circumstance doesn't affect the analysis and internal validity, but reduce the generalizability of the results.
@@
 
-Because the purpose of this thesis is to statistically assess how many patients that should have gone to the ICU but did not, these particular selection biases are not expected to skew the outcome, and therefore does not have to be adressed.
+Because the purpose of this thesis is to statistically assess how many patients that should have gone to the ICU but did not, these particular selection biases are not expected to skew the outcome, and therefore does not have to be addressed.
 
@@
 
+<!--  
 Propensity score matching in itself introduces a source of bias from unfit assumptions.
@@
 The 60% shall be used to produce the propensity score matching, and the remaining 40% shall be used to verify it.
+-->
+
+<!-- Fråga: vad är distinktionen mellan bias och error med PSM egentligen?
+är det error eller bias som uppstår om jag ex skulle välja ICU-sjukhuset som prediktor?
+Är det bias att ha för generösa calipers?
+Som jag förstår finns risk att få vilket resultat som helst med val metod för matching och limitations ?
+Och att jag väljer calipers som producerawr resultat som jag tycker känns "rimliga"?
+Ex jag kanske väntar mig att ICU bristen är 20% för jag föreställer mig att de patienter jag söker är de som är "på ribban" alltså de matchar de "friskaste" 20% av dem som var på ICU.
+Det känns mindre sannolikt att en patient inte fått ICU om den är jämförbar med någon som var en av de mest skadade 20% på ICU.
+-->
 
@@
 <!-- Explain how the study size was arrived at -->
-The data of 16000 trauma patients were collected at the participating hospitals over 2.5 years, as described in chapter Settings.
+Data of 16000 trauma patients were collected at the participating hospitals over 2.5 years, as described in chapter Settings.
 
@@
 
-The theory being examined in this masters thesis is that there exist a subset of the TITCO dataset that did not go to the ICU, but are very similar to the patients that did go to the ICU, in the covariates that caused ICU-treated patients to receive this intervention.
+The theory being examined in this master's thesis is that there exist a subset of the TITCO dataset that did not go to the ICU, but are very similar to the patients that did go to the ICU, in the covariates that caused ICU-treated patients to receive this intervention.
 The thesis is that these patients were not admitted to the ICU only for lack of such resources, not by a medical assessment.
@@
 Patients were admitted to the ICU as a consequence of their medical condition.
-Performing a randomized case-control study and randomly assigning or withholding a live-saving treatment would obviously be immoral, but we believe that in the TITCO data set exist patients who were not admitted into the ICU by volition but for scarcity of such resources.
+Performing a randomized case-control study and randomly assigning or withholding a live-saving treatment would obviously be immoral, but we believe that in the TITCO data set exist patients who were not admitted into the ICU not by volition but for scarcity of such resources.
 Such a group would be akin to an non-experimental control group and PSM can be used to identify and assemble it.
@@
 The propensity score model is developed using the Training Data and tested using the Validation Data.
-The workflow contain a series of steps and the work flow should understood as linear, but iterative.
+The workflow contains a series of steps and the work flow should understood as linear, but iterative.
 The output of each step is the input for later steps.
-Therefore the output of the last step reflect the cumulative error from unfit assumptions made the second and fourth step.
+Therefore the output of the last step contain the cumulative error from unfit assumptions made the second and fourth step.
 The output is tested against Validation Data.
@@
 
+<!-- ÅTERKOM inte klart bara gissat -->
+
 The first step of PSM is to randomly split the data set into two parts, 60% and 40%.
-The 60% part is a Training Part used to delvelop and train the PSM, and the 40% part is the Validation Part.
+The 60% part is a Training Part used to develop and train the PSM, and the 40% part is the Validation Part.
 This is done because the inherent bias of the PSM method described in chapter Bias.
@@
 This selection is based on clinical experience as well as analysis of the Training Data.
-In the data analysis all properties of the patients that were sent to the ICU will be compared to the patents who did not, in order to identify properties that distinguishes this cohort., Familiarity with the collection of source data is crucial to minimize the risk to choose irrelevant predictors and introduce error.
+In the data analysis all properties of the patients that were sent to the ICU will be compared to the patents who did not, in order to identify properties that distinguishes this cohort.
+Familiarity with the collection of source data is crucial to minimize the risk to choose irrelevant predictors and introduce error.
 
@@
 In PSM members of the subset that did not get ICU care but had similar predictors - and therefore similar propensity scores - as the subgroup that did get ICU care, are matched to patients who did go to the ICU.
-There are a few different ways to perform this matching but they are all imperfect because propensity scores are unique and perfect matching is unattainable There are control mechanisms and limitations but a risk of applying too fix or too loose matching criteria the outcome could inadvertedly vastly under- or overestimate the  need of ICU beds remain.
+There are a few different ways to perform this matching but they are all imperfect because propensity scores are unique and perfect matching is unattainable There are control mechanisms and limitations but a risk of applying too fix or too loose matching criteria the outcome could inadvertently  vastly under- or overestimate the  need of ICU beds remain.
+The same procedure is repeated on the Validation Data using the same predictors and the same matching criteria, and this should produce a cohort comparable to the one from Training Data.
+The proportion of this subset versus the subset that did receive ICU care is also the proportion of ICU bed shortage.
+If, for every patient that did receive ICU care, there was another patient with similar predictors that wasn’t admitted due to scarcity of ICU resources, that would mean the ICU resources are 50% of what they ought to be.
 
-The same procedure is repeated on the Validation Data using the same predictors and the same matching criteria, and this should produce a cohort comparable to the one trom Training Data.
-
-(b) Describe any methods used to examine subgroups and interactions
+<!-- (b) Describe any methods used to examine subgroups and interactions -->
 
@@
 
+Predictors are independent variables with strong causality with the dependent varaiable.
 Predictor variables will be selected using both clinical judgement and data analysis.
+Because causality is presumed, predictors must precede ICU admission.
+For example, though there is a strong correlation between ICU admission and mortality, mortality can not be a predictor because it did not cause ICU admission.
 
-Clinical experience compels inclusion of vital parameters, irregardless if there is any statistical motivation. Such variables are heart rate, systolic blood pressure and respiration rate.
+Clinical experience compels inclusion of vital parameters, regardless if there is any statistical motivation. Such variables are heart rate, systolic blood pressure and respiration rate.
 
-The TITCO-I data set is analyzed mathemathically in pursuit of suitable predictor variables.
+The TITCO-I data set is analysed mathematically  in pursuit of suitable predictor variables.
 The data set is divided into two subgroups, the cohort that were admitted to the ICU and the cohort that was not.
@@
 p- and t-values when comparing these cohorts to show if there are any statistically significant differences.
+It is plausible to construct new variables from quantitative data, such as trends.
 
-For the qualitative data, the propensity of each parameter to get a patient admitted to the ICU is in relation to the proportion of patients that were admitted to the ICU from the whole TITCO-I data set, about a third.
+Timer data can be used to construct quantitative variables delay-of-care.
+Second and third delays, corresponding to the delays between incident and hospital admission and between admission and first survey have been proposed to have high relevance for patient outcome @indiantrauma .
+
+For the qualitative data, the propensity of each parameter for ICU admission is compared to the proportion of patients that were admitted to the ICU from the whole TITCO-I data set, about a third.
 If considerably more than a third of each qualitative parameter are admitted to the ICU, then that parameter is of interest.
-
-Some qualitative parameters such as intubation are recorded after both 1 and 24 hours, if applicable.
-The record at 1 hour has two options indicating the patient is intubated, "before arrival" meaning the patient was intubated at a their first health care facility and "yes" meaning the patient was intubated within the first hour.
-Because these two options both mean the patient was intubated after 1 hour, an aggregate parameter has been created for this, and a further parameter "intubated after 24 hours" that summarize the patients who were intubated on arrival, the patients who were intubated within the first hour and the patients who were intubated between 1 and 24 hours.
-
-FRÅGA: det finns pat som är intuberade "before arrival" men som inte är noterade som transferred från ett annat sjukhus.
-De har kommit med motor rickshaw.
-Är det rimligt att anta att de är transferred från ett sjukhus iallafall för vem har annars intuberat dessa patienter?
-
-<!-- 
-The parameter "taken to the operation theatre" is a heterogenous qualitative parameter, where only a small proportion of patients had been taken within the first hour but a substantial proportion of patients had been taken to the operation theatre after 24 hours.
-There is a need to consider the free text field operation type to filter out for operation types with higher propensity for ICU admission from operations with lower propensity.
--->
-The parameter "taken to the operation theatre" is a heterogenous qualitative parameter, and operations performed in the first hour tend to be more acute than operations performed within the first day of hospitalization.
-There is a need to consider the free text field operation type to filter out for operation types with higher propensity for ICU admission from operations with lower propensity.
-FRÅGA: borde man ta operation within 24 hours och komplettera med text based operationstyp?
+It has been relevant to combine qualitative parameters into aggregate qualitative varaibles.
+For example, the variable intubation at 1 hour has two options indicating the patient is intubated: "before arrival" meaning the patient was intubated at a their first health care facility and "yes" meaning the patient was intubated within the first hour.
+Because these two options both mean the patient was intubated after 1 hour, an aggregate parameter has been created for this.
+Because the data set also contain information of whether a patient was intubated within the first and 24th hour the parameter "intubated after 24 hours" were created that summarize the patients who were intubated on arrival, the patients who were intubated within the first hour and the patients who were intubated between 1 and 24 hours.
 
@@
 These text based data fields will be examined for entries that are more prevalent in the ICU cohort than the non-ICU cohort, and combined with clinical experience to select suitable predictors.
+
+It is also plausible to include aggregate variables, by combining unrelated variables of the same type or across varaible types.
+Selection and construction of such aggregate parameters to be explored will be based on clinical experience not data analysis.
+An example of this is the heterogenous qualitative parameter "taken to the operation theatre".
+Many operation types are strongly correlated to ICU admission - especially if they were performed within the first hour of admission - but most of the operations performed are lower priority orthopedic operations with no correlation to ICU admission.
+There is a need to combine operation timing and free text fields "operation type" and "intraoperative findings" to filter out operations with higher propensity for ICU admission.
+<!-- FRÅGA prediktorval: borde man ta operation within 24 hours och komplettera med text based operationstyp?
+-->
 
@@
 setwd("C:/Users/maria/Downloads/Packt Learning RStudio for R Statistical Computing 2012 RETAIL eBook-repackb00k/unmet-ICU-beds")
-load("table3.rdata")
+load("table5.rdata")
 
-knitr::kable(table3, caption = "Means and standard devations of quantitative variables of ICU and non-ICU admitted patients, respectively")
+knitr::kable(table5, caption = "Means, standard devations of quantitative variables of ICU and non-ICU admitted patients, respectively and p- and t-values of comparison between the sets")
 ```
+<!-- FRÅGA: är p-värdena rimliga tycker du? de är väldigt låga och min känsla - med 5 hp statistik i ryggen - är att std-avvikelsen är för stor för det? -->
 
@@
 However, low scores in either category is a high propensity to admit a patient to the ICU, as demonstrated in the comparison of quantitative parameters.
-
-P-values and T-values of non-ICU vs ICU patients:
-```{r}
-setwd("C:/Users/maria/Downloads/Packt Learning RStudio for R Statistical Computing 2012 RETAIL eBook-repackb00k/unmet-ICU-beds")
-load("statresult.rdata")
-
-knitr::kable(statresult, caption = "P- and t-values of the quantitative values in comparison of non-ICU and ICU patients")
-```
-FRÅGA: är p-värdena rimliga tycker du? de är superlåga och min känsla - med 5 hp statistik i ryggen - är att std-avvikelsen är för stor för det?
 

### 2022-09-27 main.R

@@ -20,20 +20,20 @@
 ## clean R session.
 #noacsr::source_all_functions()
 
-#setwd("C:/Users/maria/Downloads/Packt Learning RStudio for R Statistical Computing 2012 RETAIL eBook-repackb00k/unmet-ICU-beds")
+setwd("C:/Users/maria/Downloads/Packt Learning RStudio for R Statistical Computing 2012 RETAIL eBook-repackb00k/unmet-ICU-beds")
 
-setwd("C:/Users/maria/Downloads/unmet-ICU-beds")
+#setwd("C:/Users/maria/Downloads/unmet-ICU-beds")
 
 library(tidyverse)
 library(dplyr)
 library(tableone)
 
 ## Import data
-headerproperties <- read.csv("headerproperties.csv")
-titcodata <- read.csv("titcodata.csv")
-#jag har manuellt ändrat och städat lite i dessa så jag sparar dem så jag kan committa dem:
-save(headerproperties, file = "headerproperties.RData")
-save(titcodata, file = "titcodata.RData")
+headerproperties <- read.csv("headerpropertieswtimers.csv")
+titcodata <- read.csv("titcodatawtimers.csv")
+#jag har manuellt ändrat och städat lite i dessa så jag sparar dem så jag kan committa dem
+save(headerproperties, file = "headerpropertieswtimers.RData")
+save(titcodata, file = "titcodatawtimers.RData")
 headerrows <- head(headerproperties)
 
 #FRÅGA: jag kan inte hitta hur man definierar eller filtrerar på row headers, men jag kan göra det här superjobbiga sättet
@@ -90,6 +90,9 @@ df_t <- data.table::transpose(quantitativelong)
 statresult<-data.frame(df_t,pvalues,tvalues) #jag vet inte hur jag får headers att stå p och t-värde, men det blir en tabell
 colnames(statresult)<-data.frame("variable short","p-värde","t-värde")
 
+table4<-data.frame(pvalues,tvalues) #jag vet inte hur jag får headers att stå p och t-värde, men det blir en tabell
+colnames(table4)<-data.frame("p-värde","t-värde")
+
 b=quantitatestatlicu[["ContTable"]][["Overall"]]
 myDf <- as.data.frame(b)
 ICU_mean=myDf$mean
@@ -109,6 +112,9 @@ mean_diffdf<- as.data.frame(mean_diff)
 #det här är mean och std av ICU vs no-ICU
[…]

### 2022-09-27 textfiles.R

@@ -145,7 +145,6 @@ extsort=print(extdf[order(extdf$Freq, decreasing = TRUE), ]   )
 
 present <- data.frame(xraysort[1:50,1:2],fastsort[1:50,1:2],opsort[1:50,1:2],ctsort[1:50,1:2],extsort[1:50,1:2])
 
-
 write.csv(present,"present.csv")
 #okej nu har vi en matris med alla x-ray-kolumner
 #borde kunna räkna nu

### 2022-09-27 manuscript.Rmd
flyttade kvant predictorn ur main den är en special operation tänkte jag

@@
 Despite having only 1% of the global vehicle fleet, a disproportionate 10% of global traffic casualties occur in India @indiantrauma.
-Half of trauma mortality in India occur in-hospital.
-There is no national trauma registry in India, but isolated studies have shown that in-hospital mortality within 30 days of hospitalization is higher than similarly injured patients in HIC hospitals @indiantrauma.
+
 Globally, an estimated two million lives could be saved annually, if the appropriate and timely trauma care provided by hospitals in many high-income countries were available and provided in low to-middle-income countries (LMICs) @indiantrauma.
+Half of trauma mortality in India occur in-hospital, and while there is no national trauma registry in India, isolated studies have shown that in-hospital mortality within 30 days of hospitalization is higher than similarly injured patients in HIC hospitals @indiantrauma.This difference in in-hospital mortality is likely multi-factorial, and any factors have been proposed to explain it such as differences in triage and delay-of-care, as well as shortage of hospital resources including ICU beds @onemilliondead ,@traumaLMIC .
 
@@
 
-A diverse mix of public, for-profit and not-for-profit private actors provide health services in India, where private providers dominate.
+A diverse mix of public, for-profit and not-for-profit private actors provide health services in India, where a majortiry of health care resources are held by private health care providers.
 According to a WHO 2022 report, India has 76 709 non-ventilator supported and 39 476 ventilator-supported ICU beds @ICUbeds.
@@
 
-Methods to examine trauma care and ensure equivalence
------------------------------------------------------
 In order to improve and ensure equivalence of intra-hospital care, trauma quality improvement programs (TQIP) have been developed.
-The pilot and most notable example is the TQIP developed by the American College of Surgeons ACS.
-Since the implementation of the ACS pilot, TQIP has evolved to a broad term that include many kinds of processes to improve the quality of trauma care, such as mortality and morbidity conferences, using audit filters or comparing risk adjusted outcomes over time and between centres.
-The aim is to find discrepancies in treatment for patients presenting with similar conditions.
-This difference in in-hospital mortality is likely multi-factorial, and any factors have been proposed to explain it such as differences in triage and delay-of-care, as well as shortage of hospital resources including ICU beds @onemilliondead ,@traumaLMIC .
-TQIPs aim to ensure patients get equivalent care within and across hospitals, to benchmark performance of health care providers versus their peers and to improve trauma care by developing best practice guidelines @tqip.
-
-TQIP often originate in high-income countries (HIC), they are either not applicable nor easily transferred to the hospital conditions of LMIC @frailty @gunshot  @onemilliondead.
-For trauma patients, there exist significant structural differences between HIC and LMIC.
-These differences are in factors such as injury mechanisms and severity, health care resources as well as patient make-up @frailty @indiantrauma @traumaLMIC @onemilliondead.
-These differences make a direct transferral of HIC TQIP models onto LMIC irrelevant or misleading and there is a need to develop TQIP assessment applicable to local conditions.
-
-<!-- Ni vil båda flytta fokus från detta, och jag förstår inte riktigt varför det är med kopplingen är luddig.
-Är det för att säga: det finns TQIPs som ger en fingervisning om hur ICU ska dimensioneras, bestyckas, bemannas och bedrivas Men det finns inga färdiga TQIPs som kan användas för att kvantifiera ICU-bristen i Indien och vi tror det är vanskligt att använda de befintliga eftersom de skapats under så olika förhållanden.
-Och därför måste vi börja från noll?
--->
+TQIP has evolved to a broad term that include many kinds of processes to improve the quality of trauma care, such as mortality and morbidity conferences, using audit filters or comparing risk adjusted outcomes over time and between centres.
+The TITCO data set is among the first trauma registry data bases established in India and can be used for TQIPs in an Indian context @titco.
+TQIP often originate in high-income countries (HIC), they are either not applicable nor easily transferred to the hospital conditions of LMIC @frailty @onemilliondead.
+For trauma patients, there exist significant structural differences between HIC and LMIC in factors such as injury mechanisms and severity, health care resources as well as patient make-up @frailty @indiantrauma @traumaLMIC @onemilliondead.
 
@@
 There exists a potential bias in the routines, resources and patient makeup of the participating hospitals.
-Because they are all tertiary hospitals, about two thirds of the patients have been referred from other health care providers, meaning that there is a selection bias in the patient populations.
+Because they are all tertiary hospitals, about two thirds of the patients have been referred from other health care providers meaning that there is a selection bias in the patient populations.
 Further, hospitals with dedicated trauma wards have a propensity to receive transferred or police escorted patients with worse condition and prognosis.
+The hospitals in this study were all top-tier tertial hospitals in their respective community, meaning they did not refer patients onward.
 
@@
 This circumstance doesn't affect the analysis and internal validity, but reduce the generalizability of the results.
-<!-- (ÅTERKOM: har inte råd att refereas?) -->
 
@@
 Hospitals with trauma wards could have lower admission criteria to the ICU than hospitals with no trauma ward.
-Because examining admission criteria to the ICU is part of the analysis it may be necessary to examine and address this bias by comparing ICU admitted patients between hospitals.
-Propensity score matching can be used to identify differences in ICU admission criteria between hospitals and find patients that were admitted but need not have been.
-These patients are excluded from predictor selection to reduce this bias.
-This is part of the development and training of the PSM model.
-
+Because examining admission criteria to the ICU is part of the analysis, any bias from differences in hospital admission criteria must be minimized.
+This is why the PSM model is trained using a training and validation subset as described in the chapter Statistical methods.
+If this proves insufficient and unsatisfactory it may be necessary to examine and address this bias by comparing ICU admitted patients between hospitals using the same PSM methodology, but this will go beyond the time frame and scope of this thesis.
 <!--  
@@
 Som jag förstår finns risk att få vilket resultat som helst med val metod för matching och limitations ?
-Och att jag väljer calipers som producerawr resultat som jag tycker känns "rimliga"?
+Och att jag väljer calipers som producerar resultat som jag tycker känns "rimliga"?
 Ex jag kanske väntar mig att ICU bristen är 20% för jag föreställer mig att de patienter jag söker är de som är "på ribban" alltså de matchar de "friskaste" 20% av dem som var på ICU.
 Det känns mindre sannolikt att en patient inte fått ICU om den är jämförbar med någon som var en av de mest skadade 20% på ICU.
+Men ICU-populationen kanske inte stratifierar så, ICU-patienterna är kanske snarare en pyramid där de flesta är "på ribban" och sen finns några få riktigt riktigt skadade i toppen av pyramiden som inte matchar icke-ICU populationen alls men inte är så viktig för de är så få.
+Vilket för mig vidare på tankar om val av matchningsmetod nearest neighbor osv.
+I vår analys vill vi inte hämta en specifik proportion that would defeat the purpose.
+Och men då är frågan om varje pat i de lägre "stratosfärerna" av ICU kan matchas till flera icke-ICU medan svårt skadade ICU-pat blir utan matchingar?
 -->
@@
 The propensity score model is developed using the Training Data and tested using the Validation Data.
-The workflow contains a series of steps and the work flow should understood as linear, but iterative.
+The workflow contains a series of steps.
 The output of each step is the input for later steps.
-Therefore the output of the last step contain the cumulative error from unfit assumptions made the second and fourth step.
-The output is tested against Validation Data.
+The model is developed using Testing Data and is tested with Validation Data.
 By reverting back and adjusting assumptions in an iterative process, the model is continuously improved and the accuracy and validity is improved.
 
-<!-- ÅTERKOM inte klart bara gissat -->
-
 The first step of PSM is to randomly split the data set into two parts, 60% and 40%.
-The 60% part is a Training Part used to develop and train the PSM, and the 40% part is the Validation Part.
+The 60% part is a Training Data used to develop the propensity score model, and the 40% part is the Validation Part.
 This is done because the inherent bias of the PSM method described in chapter Bias.
-Training Data is used in steps 2-4 and the output is then validated using Validation Data.
+Training Data is used in steps 2-3 and the resulting model is then validated using Validation Data.
 
@@
 
-Third step is to build a standard multivariable logistic regression model with ICU care as the outcome and the predictors selected in the second step as independent variables.
-This model is used to calculate the standard variable propensity score for all patients.
+Third step is to build a multivariable logistic regression model with ICU care as the outcome and the predictors selected in the second step as independent variables using Testing Data.
+
+Fourth step is to apply this model on patients from the Validation Data set.
+Using the model, calculate the standard variable propensity score for each patient in the Validation Data set and determine whether the patients that were admitted to the ICU in the validation set receive appropriate propensity scores by the model constructed in step 2-3.
 The propensity score determines the likelihood that a patient will receive ICU care, given their condition.
 
-Fourth step is PSM.
+<!-- 
+#fråga: är det såhär man gör menade du Johanna?
+vad är det egentligen som händer i steg 3 och vad består "träningen" av i steg 2-3?
+År det att man har valt sina variabler i steg 2 men vi vet inte om de är bra och vi vet inte deras inbördes relevans/vikt för den oberoende variabeln.
+Så vi petar in dem i steg 3 och klickar OK, och sen bygger R en multivariabelregressionsmodell åt en som viktar prediktorerna åt en så att modellen tilldelar patienterna propensity scores som stämmer bäst överens med deras ICU-status (som är känt för R).
+R tweakar/vikar/manipulerar modellen att passa träningsdatasetet man matat den med, och bygger en modell som mha prediktorerna är bäst på att identifiera de patienter som R vet hamnade på ICU?
+
+Det gör vi för att optimera modellens kapacitet att "förutse" vad som kommer hända med en "helt ny" patient med en uppsättning värden på prediktorerna? 
+Så har vi i Validation Data sparat en uppsättning "helt nya" patienter som vi prövar modellen på? Och ser om modellen är bra på att ge ICU-patienter rimliga propensity scores?
+Är det det du menar att modellen ska "träna" på ett set och testas på ett annat?
+
+Och om man switchar över redan i steg 4 finns det ingen validering sen mera för dålig PSM i steg 5? -->
+
+Fifth step is PSM.
 In PSM members of the subset that did not get ICU care but had similar predictors - and therefore similar propensity scores - as the subgroup that did get ICU care, are matched to patients who did go to the ICU.
-There are a few different ways to perform this matching but they are all imperfect because propensity scores are unique and perfect matching is unattainable There are control mechanisms and limitations but a risk of applying too fix or too loose matching criteria the outcome could inadvertently  vastly under- or overestimate the  need of ICU beds remain.
-The same procedure is repeated on the Validation Data using the same predictors and the same matching criteria, and this should produce a cohort comparable to the one from Training Data.
-The proportion of this subset versus the subset that did receive ICU care is also the proportion of ICU bed shortage.
-If, for every patient that did receive ICU care, there was another patient with similar predictors that wasn’t admitted due to scarcity of ICU resources, that would mean the ICU resources are 50% of what they ought to be.
+There are a few different ways to perform this matching but they are all imperfect because propensity scores are unique and perfect matching is unattainable.
+There are control mechanisms and limitations but a risk of applying too fix or too loose matching criteria and inadvertently vastly under- or overestimate the  need of ICU beds remain.
+<!-- (b) 1. visst finns dessa svagheter i propensity score matching?
+2. vad gör man åt dem? 
+3. är beskrivningen av åtgärder mot detta något som hör hemma här eller i results?-->
 

### 2022-09-27 quantpredictors.R

@@ -1,0 +1,73 @@
+setwd("C:/Users/maria/Downloads/Packt Learning RStudio for R Statistical Computing 2012 RETAIL eBook-repackb00k/unmet-ICU-beds")
+
+
+library(tidyverse)
+library(dplyr)
+library(tableone)
+
+headerproperties <- read.csv("headerpropertiestrans.csv") #från rad 70 har datan fär "type" hamnat i "note"-kolumnen. dessutom är spo2_o2_1 och 2 markerade som kvantitativa men de är Y/n så det har jag löst manuellt
+titcodata <- read.csv("titcodatafull.csv") #untampered version
+
+#vi måste hantera att age är numeric och quantiativ men >89 förekommer
+#acceptableage<-data.frame(0:89)
+titcodata=filter(titcodata, age !=">89")
+#resten har lästs som char pga tecknet, måste bytas till num
+titcodata[,5] <- as.numeric(titcodata[,5])
+
+#ur headerprops, filtrera fram alla rader i type-kolumnen som är är c("qualitative", "quantitative")), och skapa en array av motsvarande rader i name-kolumnen
+quantcolname <- headerproperties %>% filter(type %in% c("quantitative")) %>% select(name)
+df_t<- headerproperties %>% filter(type %in% c("quantitative")) %>% select(label)
+#name är header-raden i titco
+# # Välj kolumner ur datat baserat på listan av variabler
+#name-arrayen variables kan du nu applicera på titcodatan, och välja alla kolumner som finns nämnd i arrayen. variables$name byter objekttyp till chr 1:56 som måste till. all_of gör inget
+quantcols =titcodata %>% select(all_of(quantcolname$name))
+
+licu=filter(titcodata, licu >0.5)
+nolicu=filter(titcodata, licu <0.5)
+
+
+quantcolnametrans<-data.table::transpose(quantcolname)
+
+#välj kvant kolumnerna
+quantLicu <- licu[, which((names(licu) %in% quantcolnametrans)==TRUE)]
+quantNolicu <- nolicu[, which((names(nolicu) %in% quantcolnametrans)==TRUE)]
+quantTitco <- titcodata[, which((names(titcodata) %in% quantcolnametrans)==TRUE)]
+
[…]

### 2022-09-27 manuscript.Rmd
kväll 27e

--- 2022-09-27 manuscript.Rmd
@@
 As stated in chapter Settings, most patients have low socioeconomic status.
-Low socioeconomic status is associated with higher level of malnutrition and comorbidities and would have a negative effect on vitals and the capacity of patients to tolerate certain interventions.
-This circumstance doesn't affect the analysis and internal validity, but reduce the generalizability of the results.
-<!-- (ÅTERKOM: har inte råd att refereas?) -->
+Low socioeconomic status is associated with higher levels of malnutrition and comorbidities and would have a negative effect on vital parameters and the capacity of patients to tolerate both injuries and interventions.
+This circumstance doesn't affect the analysis and internal validity, but reduces the generalizability of the results.
 
@@
 Hospitals with trauma wards could have lower admission criteria to the ICU than hospitals with no trauma ward.
-Because examining admission criteria to the ICU is part of the analysis it may be necessary to examine and address this bias by comparing ICU admitted patients between hospitals.
-Propensity score matching can be used to identify differences in ICU admission criteria between hospitals and find patients that were admitted but need not have been.
-These patients are excluded from predictor selection to reduce this bias.
-This is part of the development and training of the PSM model.
+Because examining admission criteria to the ICU is part of the analysis, any bias from differences in hospital admission criteria must be minimized.
+This is why the PSM model is trained using a training and validation subset as described in the chapter Statistical methods.
+If this proves insufficient and unsatisfactory it may be necessary to examine and address this bias by comparing ICU admitted patients between hospitals using the same PSM methodology, but this will go beyond the time frame and scope of this thesis.
 
@@
 The thesis is that these patients were not admitted to the ICU only for lack of such resources, not by a medical assessment.
-
 Translated into this context this means that in this data set exist a selection bias, there is causality between the independent variables medical conditions and interventions and the dependent variable ICU admittance.
@@
 Such a group would be akin to an non-experimental control group and PSM can be used to identify and assemble it.
-If such a subset can be identified by PSM, then it naturally follows that there exists shortage of ICU beds in tertiary hospitals in India.
-It also follows that the shortage can be quantified by the proportionality of this subset versus the subset that did receive ICU care.
-If, for every patient that did receive ICU care, there was another patient with similar predictors that wasn’t admitted due to scarcity of ICU resources, that would mean the ICU resources are 50% of what they ought to be.
+If such a subset can be identified by PSM, that is indicative there exists shortage of ICU beds in tertiary hospitals in India.
+
+The proportion of the subset of patients that should hade received ICU care based on our prediction model versus the subset that did receive ICU care is also the proportion of ICU bed shortage.
+If, for every patient that did receive ICU care, there was another patient with similar predictors that should hade received ICU care based on our prediction model but did not due to scarcity of ICU resources, that would mean the ICU resources are 50% of what they ought to be.
 
 The propensity score model is developed using the Training Data and tested using the Validation Data.
-The workflow contains a series of steps and the work flow should understood as linear, but iterative.
+The workflow contains a series of steps.
 The output of each step is the input for later steps.
-Therefore the output of the last step contain the cumulative error from unfit assumptions made the second and fourth step.
-The output is tested against Validation Data.
+The model is developed using Testing Data and is tested with Validation Data.
 By reverting back and adjusting assumptions in an iterative process, the model is continuously improved and the accuracy and validity is improved.
 
-<!-- ÅTERKOM inte klart bara gissat -->
-
 The first step of PSM is to randomly split the data set into two parts, 60% and 40%.
-The 60% part is a Training Part used to develop and train the PSM, and the 40% part is the Validation Part.
-This is done because the inherent bias of the PSM method described in chapter Bias.
-Training Data is used in steps 2-4 and the output is then validated using Validation Data.
+The 60% part is Training Data used to develop the propensity score model, and the 40% part is the Validation Part.
+This is done because the inherent bias of the PSM method.
+The prediction model produced in step 3 is based on and fitted against the Training data and predictors are weighed so that the the model awards patients that did go to the ICU corresponding high propensity.
+This model can then be used to predict new patients propensity to go to the ICU based on their baseline predictor values, and by saving 40% in the Validation data set we have a set of patients that is new to the model on which we may test how well the model predicted their fate.
+Training Data is used in steps 2-3 and the resulting model is then validated using Validation Data.
 
@@
 
-Third step is to build a standard multivariable logistic regression model with ICU care as the outcome and the predictors selected in the second step as independent variables.
-This model is used to calculate the standard variable propensity score for all patients.
+Third step is to build a multivariable logistic regression model with ICU care as the outcome and the predictors selected in the second step as independent variables using Testing Data.
+
+Fourth step is to apply this model on patients from the Validation Data set, and see whether the model correctly prediced which patients in the Validation data received ICU treatment.
+Using the model, the standard variable propensity score is calculated for each patient in the Validation Data set.
+If the patients that were admitted to the ICU in the validation set receive appropriate propensity scores by the model constructed in step 2-3, the model is valid.
 The propensity score determines the likelihood that a patient will receive ICU care, given their condition.
+<!-- 
+#fråga: är det såhär man gör menade du Johanna?
+vad är det egentligen som händer i steg 3 och vad består "träningen" av i steg 2-3?
+År det att man har valt sina variabler i steg 2 men vi vet inte om de är bra och vi vet inte deras inbördes relevans/vikt för den oberoende variabeln.
+Så vi petar in dem i steg 3 och klickar OK, och sen bygger R en multivariabelregressionsmodell åt en som viktar prediktorerna åt en så att modellen tilldelar patienterna propensity scores som stämmer bäst överens med deras ICU-status (som är känt för R).
+R tweakar/vikar/manipulerar modellen att passa träningsdatasetet man matat den med, och bygger en modell som mha prediktorerna är bäst på att identifiera de patienter som R vet hamnade på ICU?
+Det gör vi för att optimera modellens kapacitet att "förutse" vad som kommer hända med en "helt ny" patient med en uppsättning värden på prediktorerna? 
+Så har vi i Validation Data sparat en uppsättning "helt nya" patienter som vi prövar modellen på? Och ser om modellen är bra på att ge ICU-patienter rimliga propensity scores?
+Är det det du menar att modellen ska "träna" på ett set och testas på ett annat?
+Och om man switchar över redan i steg 4 finns det ingen validering sen mera för dålig PSM i steg 5? -->
 
-Fourth step is PSM.
+Fifth step is PSM.
 In PSM members of the subset that did not get ICU care but had similar predictors - and therefore similar propensity scores - as the subgroup that did get ICU care, are matched to patients who did go to the ICU.
-There are a few different ways to perform this matching but they are all imperfect because propensity scores are unique and perfect matching is unattainable There are control mechanisms and limitations but a risk of applying too fix or too loose matching criteria the outcome could inadvertently  vastly under- or overestimate the  need of ICU beds remain.
-The same procedure is repeated on the Validation Data using the same predictors and the same matching criteria, and this should produce a cohort comparable to the one from Training Data.
-The proportion of this subset versus the subset that did receive ICU care is also the proportion of ICU bed shortage.
-If, for every patient that did receive ICU care, there was another patient with similar predictors that wasn’t admitted due to scarcity of ICU resources, that would mean the ICU resources are 50% of what they ought to be.
+There are a few different ways to perform this matching but they are all imperfect because propensity scores are unique and perfect matching is unattainable.
+There are control mechanisms and limitations but a risk of applying too fix or too loose matching criteria and inadvertently vastly under- or overestimate the  need of ICU beds remain.
+<!-- (b) Fråga: 1. visst finns dessa svagheter i propensity score matching?
+2. vad gör man åt dem? 
+3. är beskrivningen av åtgärder mot detta något som hör hemma här eller i results?-->
+
+The proportion of the subset of patients that probably should hade received ICU care based on our prediction model versus the subset that did receive ICU care is also the proportion of ICU bed shortage.
 

### 2022-09-27 main.R

@@ -36,85 +36,6 @@ save(headerproperties, file = "headerpropertieswtimers.RData")
 save(titcodata, file = "titcodatawtimers.RData")
 headerrows <- head(headerproperties)
 
-#FRÅGA: jag kan inte hitta hur man definierar eller filtrerar på row headers, men jag kan göra det här superjobbiga sättet
-typefilt=filter(headerrows, name == "type") #gör en dataframe av type-raden
-typetrans <- data.table::transpose(typefilt) #transponera den. jo. humor me. jag måste transponera eftersom jag bara vet hur man filtrerar på kolumnnamn
-timers=filter(typetrans, V1 == "time" | V1 == "date") #NU kan jag applicera den här arbetshästen
-counttimers=length(timers)
-#och det kan jag bara göra för att göra counts jag har ju tappat alla associationer inom datasetet
-
-#ladda vektorerna med kvantitativa och kvalitativa datan
-load("quantitativecolumnname.rdata")
-load("qualitativecolumnname.rdata")
-load("textcolumnname.rdata")
-load("quantitativelong.rdata")
-
-#sortera bort DOA
-alive=filter(titcodata, incl != 2)
-titcodata=alive
 
 ## Whatever you do next, maybe clean data?
-#definiera dataseten
-licu=filter(titcodata, licu >0.5)
-nolicu=filter(titcodata, licu <0.5)
 
-#välj kvant kolumnerna
-quantitativeLicu <- licu[, which((names(licu) %in% quantitativecolumnname)==TRUE)]
-quantitativeNolicu <- nolicu[, which((names(nolicu) %in% quantitativecolumnname)==TRUE)]
-quantitativeTitco <- titcodata[, which((names(titcodata) %in% quantitativecolumnname)==TRUE)]
-
-#gör stat analysen
-quantitatestatlicu=CreateTableOne(data = quantitativeLicu,includeNA=FALSE)
-quantitatestatnolicu=CreateTableOne(data = quantitativeNolicu,includeNA=FALSE)
-
-
-#gör t.test för alla kvant kolumner
-columncount=length(quantitativecolumnname) #hur många kolumner ska vi rulla
-pvalues<- data.frame(matrix(ncol = 1, nrow = columncount))
[…]

### 2022-09-27 textfiles.R

@@ -6,157 +6,172 @@ library(tidyverse)
 library(dplyr)
 library(tableone)
 
-titcodata <- read.csv("titcodatafull.csv")
-load("licu.rdata")
-titcodata=licu #fult
-
-#build strings for the ICU codes
-
-resultmatrix = data.frame(matrix(, nrow = 3, ncol = 11))
-resultmatrixext = data.frame(matrix(, nrow = 1, ncol = 12))
-resultmatrixct = data.frame(matrix(, nrow = 1, ncol = 13))
-resultarray = data.frame(matrix(, nrow = 1, ncol = 11))
-
-#bygg först en liten dataframe med alla ICD koder
-#nej vi måste ha tre pga om jag vill använda dem sen för att filtrera ut kolumner accepterar inte r missing values och de är tre olika längder
-#jag behöver bygga en array inte en matris av resultaten här annars vill det inte förstå, så
-rowno=1 #counter
-loop=0
-uniqueoptot= data.frame(matrix(, nrow = 11*200, ncol =1 )) #resultatmatris, extra stor
-for (a in 1:11) {
-  loop=loop+1
-xray=paste("xray_",a,"_icd",sep="")
-fast=paste("fast_",a,"_icd",sep="")
-op=paste("op_",a,"_icd",sep="")
-
-resultmatrix[1,a]=xray
-resultmatrix[2,a]=fast
-resultmatrix[3,a]=op
-
-#vilka unika har vi?
-uniqueop=unique(titcodata[op])
-startat=rowno
-endat=rowno+nrow(uniqueop)-1
-howmanyrows=nrow(uniqueop)
-uniqueoptot[startat:endat,1]<- data.frame(matrix(uniqueop[1:howmanyrows,1])) #data.frame(uniquexray[1:nrow(uniquexray),1])
-rowno=rowno+nrow(uniqueop)
-#DU VAR HÄR
-
-}
-
-uniqueexttot= data.frame(matrix(, nrow = 11*200, ncol =1 ))
-for (a in 1:12) {
-  ext=paste("e_",a,"_icd",sep="")
-  resultmatrixext[1,a]=e
+for (b in 2:2) {
+  titcodata <- read.csv("titcodatafull.csv")
+  if (a==1) {
+    licu=filter(titcodata, licu >0.5)
+    titcodata=licu #fult
+  }
+  if (a==2) {
+    nolicu=filter(titcodata, licu <0.5)
+    titcodata=nolicu #fult
+  }
   
[…]

### 2022-09-28 textfiles.R
sortering ICD koder

@@ -6,15 +6,19 @@ library(tidyverse)
 library(dplyr)
 library(tableone)
 
-for (b in 2:2) {
+for (b in 1:3) {
   titcodata <- read.csv("titcodatafull.csv")
-  if (a==1) {
+  if (b==1) {
     licu=filter(titcodata, licu >0.5)
     titcodata=licu #fult
   }
-  if (a==2) {
+  if (b==2) {
     nolicu=filter(titcodata, licu <0.5)
     titcodata=nolicu #fult
+  }
+  
+  if (b==3) {
+    #titcodata<- read.csv("titcodatafull.csv")
   }
   
   #build strings for the ICU codes
@@ -27,9 +31,13 @@ for (b in 2:2) {
   #bygg först en liten dataframe med alla ICD koder
   #nej vi måste ha tre pga om jag vill använda dem sen för att filtrera ut kolumner accepterar inte r missing values och de är tre olika längder
   #jag behöver bygga en array inte en matris av resultaten här annars vill det inte förstå, så
-  rowno=1 #counter
+  rownoxray=1 #counter
+  rownofast=1
+  rownoop=1
   loop=0
-  uniqueoptot= data.frame(matrix(, nrow = 11*200, ncol =1 )) #resultatmatris, extra stor
+  uniquexraytot= data.frame(matrix(, nrow = 11*200, ncol =3 )) #resultatmatris, extra stor
+  uniquefasttot= data.frame(matrix(, nrow = 11*200, ncol =3 )) #resultatmatris, extra stor
+  uniqueoptot= data.frame(matrix(, nrow = 11*200, ncol =3 )) #resultatmatris, extra stor
   for (a in 1:11) {
     loop=loop+1
     xray=paste("xray_",a,"_icd",sep="")
@@ -41,16 +49,34 @@ for (b in 2:2) {
     resultmatrix[3,a]=op
     
     #vilka unika har vi?
+    uniquexray=unique(titcodata[xray])
+    uniquefast=unique(titcodata[fast])
     uniqueop=unique(titcodata[op])
-    startat=rowno
-    endat=rowno+nrow(uniqueop)-1
-    howmanyrows=nrow(uniqueop)
-    uniqueoptot[startat:endat,1]<- data.frame(matrix(uniqueop[1:howmanyrows,1])) #data.frame(uniquexray[1:nrow(uniquexray),1])
[…]

### 2022-09-30 quantpredictors.R
underlagen i prediktorvalen

@@ -7,12 +7,171 @@ library(tableone)
 
 headerproperties <- read.csv("headerpropertiestrans.csv") #från rad 70 har datan fär "type" hamnat i "note"-kolumnen. dessutom är spo2_o2_1 och 2 markerade som kvantitativa men de är Y/n så det har jag löst manuellt
 titcodata <- read.csv("titcodatafull.csv") #untampered version
+titcodata <- read.csv("titcodatacleaned.csv") #cleaned version, see separate script
+
+
+#lägg till timerkolumnerna
+#det tog för lång tid att googla hur R hanterar HH:MM, så istället:
+
+ ss= data.frame(matrix(":00", nrow=nrow(titcodata), ncol =1 ))
+ hms=paste(titcodata$toi[1:nrow(titcodata)],ss[1:nrow(titcodata),1],sep= "")
+ hmstoi=data.frame(hms)
+ hms=paste(titcodata$toar[1:nrow(titcodata)],ss[1:nrow(titcodata),1],sep= "")
+ hmstoa=data.frame(hms)
+ hms=paste(titcodata$tom_1[1:nrow(titcodata)],ss[1:nrow(titcodata),1],sep= "")
+ hmstom1=data.frame(hms)
+# for (a in 1:nrow(titcodata)){
+# if(is.na(titcodata$toi[a])) {titcodata$toi[a]=FALSE} else {if(titcodata$toi[a]) {titcodata$toi[a]}}
+# hms=paste(titcodata$toi[a],ss[a,1],sep= "")
+# }
+
+hoursinj=substr(titcodata$toi, 1, 2)
+hoursinj=as.numeric(hoursinj)
+hoursinj=data.frame(hoursinj)
+minutesinj=substr(titcodata$toi, 4, 5)
+minutesinj=as.numeric(minutesinj)
+minutesinj=data.frame(minutesinj)
+hoursarr=substr(titcodata$toar, 1, 2)
+hoursarr=as.numeric(hoursarr)
+hoursarr=data.frame(hoursarr)
+minutesarr=substr(titcodata$toar, 4, 5)
+minutesarr=as.numeric(minutesarr)
+minutesarr=data.frame(minutesarr)
+hours1st=substr(titcodata$tom_1, 1, 2)
+hours1st=as.numeric(hours1st)
+hours1st=data.frame(hours1st)
+minutes1st=substr(titcodata$tom_1, 4, 5)
+minutes1st=as.numeric(minutes1st)
[…]

### 2022-09-30 textfiles.R

@@ -6,6 +6,10 @@ library(tidyverse)
 library(dplyr)
 library(tableone)
 
+headerproperties <- read.csv("headerpropertiestrans.csv") #från rad 70 har datan fär "type" hamnat i "note"-kolumnen. dessutom är spo2_o2_1 och 2 markerade som kvantitativa men de är Y/n så det har jag löst manuellt
+titcodata <- read.csv("titcodatafull.csv") #untampered version
+icdcodes <- read.csv("icdcodes.csv") #icd bibba
+
 for (b in 1:3) {
   titcodata <- read.csv("titcodatafull.csv")
   if (b==1) {
@@ -237,29 +241,41 @@ for (c in 1:nrow(xraysortICU)) {
        if (str1 == str2) {
          frequencies[c,1]=str1
          frequencies[c,2]=strfreqICU
+         strfreqTITCO=as.numeric(strfreqTITCO)
+         frequencies[c,3]=strfreqTITCO-strfreqICU #det funkade dåligt att göra samma sortering på no-ICU (summorna stämemr inte), men de är ju resten så
+         strfreqnoICU=strfreqTITCO-strfreqICU
+         strfreqICU=as.numeric(strfreqICU)
+         strfreqnoICU=as.numeric(strfreqnoICU)
+         if (strfreqnoICU>0) {
+         prop=strfreqICU/strfreqnoICU
+         }
+         if (strfreqnoICU==0) {
+           prop=0
+         }
+         frequencies[c,5]=prop
          frequencies[c,4]=strfreqTITCO
        }
       
       #hitta på vilken row denna finns i noICU ICD rad
-      findrow= which(xraysortnoICU$Var1 == str1, arr.ind = TRUE)
-      #översätt till ICD-kod
-      str2=xraysortnoICU[findrow,1]
-      str2 = as.character(str2) #byt till rätt form
-      strfreqnoICU=xraysortnoICU[c,2]
-      
-      if (length(str2>0)){
-      if (str1 == str2) {
-        frequencies[c,3]=strfreqnoICU
-        prop=strfreqICU/strfreqnoICU
-        frequencies[c,5]=prop
-      }
+      # findrow= which(xraysortnoICU$Var1 == str1, arr.ind = TRUE)
+      # #översätt till ICD-kod
[…]

### 2022-09-30 qualpredictors.R
beslutsunderlag urval kvalitativa prediktorer

@@ -1,0 +1,164 @@
+setwd("C:/Users/maria/Downloads/Packt Learning RStudio for R Statistical Computing 2012 RETAIL eBook-repackb00k/unmet-ICU-beds")
+
+
+library(tidyverse)
+library(dplyr)
+library(tableone)
+
+headerproperties <- read.csv("headerpropertiestrans.csv") #från rad 70 har datan fär "type" hamnat i "note"-kolumnen. dessutom är spo2_o2_1 och 2 markerade som kvantitativa men de är Y/n så det har jag löst manuellt
+titcodata <- read.csv("titcodatafull.csv") #untampered version
+
+headerproperties[2,6]="text" #gör mig av med patient ID för den har 16000 varianter 
+
+#ur headerprops, filtrera fram alla rader i type-kolumnen som är är c("qualitative", och skapa en array av motsvarande rader i name-kolumnen
+#filtrera fram array kolumnnamnen
+qualcolname <- headerproperties %>% filter(type %in% c("qualitative")) %>% select(name)
+#begripliga namnen
+quallong<- headerproperties %>% filter(type %in% c("qualitative")) %>% select(label)
+#och bygg en matrix med de kolumnerna
+qualcols =titcodata %>% select(all_of(qualcolname$name))
+
+#data frames med licuows och nonlicurows
+licu=filter(titcodata, licu >0.5)
+nolicu=filter(titcodata, licu <0.5)
+
+#transponera
+qualcolnametrans<-data.table::transpose(qualcolname)
+#välj kvant kolumnerna ut licu och nonlicurows och hela titco
+qualLicu <- licu[, which((names(licu) %in% qualcolnametrans)==TRUE)]
+qualNolicu <- nolicu[, which((names(nolicu) %in% qualcolnametrans)==TRUE)]
+qualTitco <- titcodata[, which((names(titcodata) %in% qualcolnametrans)==TRUE)]
+
+#gör stat analysen
+qualstatlicu=CreateTableOne(data = qualLicu,includeNA=FALSE)
+qualstatnolicu=CreateTableOne(data = qualNolicu,includeNA=FALSE)
+
+#man kan göra mer analys på detta men det kräver att man tolkar list-värden
+
+#unika värden, exempel (datatype character list)
+
+qualcolumncount=nrow(qualcolname) #hur många kolumner ska vi rulla
+f=0
[…]

### 2022-10-01 manuscript.Rmd
efter johannas kommentarer

@@
 output:
+  word_document: default
+  html_document: default
   pdf_document: default
-  html_document: default
-  word_document: default
 bibliography: bibliography.bib
@@
 
+<!-- 
+Definitions
+------------
+An ICU bed is a tiltable hospital bed but included in the term is also staffing and auxiliary equipment.
+The equipment is to administer oxygen and to monitor vitals.
+Some ICU beds are equipped with ventilators and if applicable length of ventilation has been registered.
+
+Staffing is one doctor in the ward, and 1:4 nurses per patient but there is not always strict adherence to staffing guidelines. 
+If the ICU ward is 100% full, patients might receive ventilation outside the ICU ward.
+Though the care these patients receive could very well be considered ICU care, they are not registered as such in the TITCO registry.
+-->
+
 Study design
@@
 exposed and unexposed -->
-The eligibility criteria for TITCO were patients who presented with injuries from trauma at the emergency departments of the participating hospitals with a mechanism of road traffic, railway, fall, burn or assault.
+The eligibility criteria for TITCO were all patients who presented with injuries from trauma at the emergency departments of the participating hospitals with a mechanism of road traffic, railway, fall, burn or assault.
 Patients who were dead on arrival were excluded, as were patients who only had isolated limb injuries.
@@
 <!-- Describe any efforts to address potential sources of bias -->
-The TITCO dataset are observational data collected from a clinical environment and not an experimental setting.
+The TITCO dataset is observational data collected from a clinical environment and not an experimental setting.
 Medical interventions, examinations and ICU admission are not randomized, they are a consequence of a patient's medical condition.
 There exists a causal effect between the independent and the dependent variables: a patient is sent to the ICU precisely because of the severity of their condition.
+There is also a selection bias in which parameters are recorded and which examinations are performed, introducing detection and observer bias.
 
-There exists a potential bias in the routines, resources and patient makeup of the participating hospitals.
-Because they are all tertiary hospitals, about two thirds of the patients have been referred from other health care providers, meaning that there is a selection bias in the patient populations.
-Further, hospitals with dedicated trauma wards have a propensity to receive transferred or police escorted patients with worse condition and prognosis.
+There exist a bias in the routines, resources and patient makeup of the participating hospitals.
+Because they are all tertiary hospitals, about two-thirds of the patients have been referred from other health care providers, meaning that there is a selection bias in the patient populations.
+Further, hospitals with dedicated trauma wards have a propensity to receive more transferred or police-escorted patients.
 
-As stated in chapter Settings, most patients have low socioeconomic status.
-Low socioeconomic status is associated with higher levels of malnutrition and comorbidities and would have a negative effect on vital parameters and the capacity of patients to tolerate both injuries and interventions.
-This circumstance doesn't affect the analysis and internal validity, but reduces the generalizability of the results.
+As stated in chapter Settings, most patients have low socioeconomic status. This circumstance doesn't affect the analysis and internal validity, but reduces the generalizability of the results.
 
+Biases like these are expected in data collected in a clinical setting, the analysis model i chosen to adress this.
 Because the purpose of this thesis is to statistically assess how many patients that should have gone to the ICU but did not, these particular selection biases are not expected to skew the outcome, and therefore does not have to be addressed.
 
-Hospitals with trauma wards could have lower admission criteria to the ICU than hospitals with no trauma ward.
+Hospitals with trauma wards could have different admission criteria to the ICU than hospitals with no trauma ward.
+The same goes for all equipment any hospital has as well as the proficiency of their physicians.
 Because examining admission criteria to the ICU is part of the analysis, any bias from differences in hospital admission criteria must be minimized.
@@
 
-Study size 
-----------
-
-<!-- Explain how the study size was arrived at -->
-Data of 16000 trauma patients were collected at the participating hospitals over 2.5 years, as described in chapter Settings.
-
 Statistical methods
@@
 Propensity score matching (PSM) is useful method for identifying and assembling a control group from within a data set where there is causality between the independent and dependent variables, and a randomized case-control (RCT) study to offset or balance this bias is impractical, immoral or otherwise improper.
-Propensity score is a statistical variable, understood as the likelihood of a subject receiving treatment conditional on observations of relevant covariates of that subject before treatment.
-In RCTs it is known from study design, but it is generally unknown in observational data.
-The rationale for PSM is that while it is unknown, this variable can be fairly well estimated using logistic regression.
 
-The theory being examined in this master's thesis is that there exist a subset of the TITCO dataset that did not go to the ICU, but are very similar to the patients that did go to the ICU, in the covariates that caused ICU-treated patients to receive this intervention.
-The thesis is that these patients were not admitted to the ICU only for lack of such resources, not by a medical assessment.
-Translated into this context this means that in this data set exist a selection bias, there is causality between the independent variables medical conditions and interventions and the dependent variable ICU admittance.
-Patients were admitted to the ICU as a consequence of their medical condition.
-Performing a randomized case-control study and randomly assigning or withholding a live-saving treatment would obviously be immoral, but we believe that in the TITCO data set exist patients who were not admitted into the ICU not by volition but for scarcity of such resources.
-Such a group would be akin to an non-experimental control group and PSM can be used to identify and assemble it.
-If such a subset can be identified by PSM, that is indicative there exists shortage of ICU beds in tertiary hospitals in India.
+The theory being examined in this master's thesis is that there exist a subset of the TITCO dataset that did not go to the ICU, but are very similar to the patients that did go to the ICU, in the covariates that caused ICU-treated patients to be admitted.
+The thesis is that these patients were not admitted to the ICU only for lack of ICU resources, not by a medical assessment.
+Such a group would be akin to an non-experimental control group that PSM can be used to identify and assemble.
 
-The proportion of the subset of patients that should hade received ICU care based on our prediction model versus the subset that did receive ICU care is also the proportion of ICU bed shortage.
+If such a group of patients can be identified by PSM, that is indicative there exists shortage of ICU beds in tertiary hospitals in India.
+The size of this group versus the size of the group that did receive ICU care indicate the magnitude of the ICU bed shortage.
 If, for every patient that did receive ICU care, there was another patient with similar predictors that should hade received ICU care based on our prediction model but did not due to scarcity of ICU resources, that would mean the ICU resources are 50% of what they ought to be.
 
-The propensity score model is developed using the Training Data and tested using the Validation Data.
+The propensity score model is developed using the Training Data and tested using the Test Data.
 The workflow contains a series of steps.
 The output of each step is the input for later steps.
-The model is developed using Testing Data and is tested with Validation Data.
+Steps three and four that involve development and validation of a multivariable regression model have several standard methods to test and improve model validity, and the matching process of step five have several means by which you can control and validate the process.
 By reverting back and adjusting assumptions in an iterative process, the model is continuously improved and the accuracy and validity is improved.
@@
 The first step of PSM is to randomly split the data set into two parts, 60% and 40%.
-The 60% part is Training Data used to develop the propensity score model, and the 40% part is the Validation Part.
-This is done because the inherent bias of the PSM method.
-The prediction model produced in step 3 is based on and fitted against the Training data and predictors are weighed so that the the model awards patients that did go to the ICU corresponding high propensity.
-This model can then be used to predict new patients propensity to go to the ICU based on their baseline predictor values, and by saving 40% in the Validation data set we have a set of patients that is new to the model on which we may test how well the model predicted their fate.
-Training Data is used in steps 2-3 and the resulting model is then validated using Validation Data.
+The 60% part is Training Data used to develop the propensity score model, and the 40% part is the Test Part.
+This is done to detect overfitting of the model to the Training Data.
+Training Data is used in steps 2-3 and the resulting model is then validated using Test Data.
 
@@
 
-Third step is to build a multivariable logistic regression model with ICU care as the outcome and the predictors selected in the second step as independent variables using Testing Data.
+Third step is to build a multivariable logistic regression model with ICU care as the outcome and the predictors selected in the second step as independent variables.
+The prediction model is based on the Training Data and predictors are weighed so that the the model awards patients that did go to the ICU corresponding high propensity.
+This model can then be used to predict new patients propensity to go to the ICU based on their baseline predictor values.
+By saving 40% in the Validation data set we have a set of patients that is new to the model.
+This allows to test how well the model correctly predicted ICU patients by awarding them high propensity scores.
 
-Fourth step is to apply this model on patients from the Validation Data set, and see whether the model correctly prediced which patients in the Validation data received ICU treatment.
-Using the model, the standard variable propensity score is calculated for each patient in the Validation Data set.
-If the patients that were admitted to the ICU in the validation set receive appropriate propensity scores by the model constructed in step 2-3, the model is valid.
+Fourth step is to apply this model on patients from the Test Data.
+Using the model, the standard variable propensity score is calculated for each patient in the Test Data set.
 The propensity score determines the likelihood that a patient will receive ICU care, given their condition.
+If the model is suitable, it will correctly identify patients that did go to the ICU by awarding them high propensity scores, but non-ICU patients that were similar to this cohort will also get high propensity scores.
+
 <!-- 
-#fråga: är det såhär man gör menade du Johanna?
-vad är det egentligen som händer i steg 3 och vad består "träningen" av i steg 2-3?
-År det att man har valt sina variabler i steg 2 men vi vet inte om de är bra och vi vet inte deras inbördes relevans/vikt för den oberoende variabeln.
-Så vi petar in dem i steg 3 och klickar OK, och sen bygger R en multivariabelregressionsmodell åt en som viktar prediktorerna åt en så att modellen tilldelar patienterna propensity scores som stämmer bäst överens med deras ICU-status (som är känt för R).
-R tweakar/vikar/manipulerar modellen att passa träningsdatasetet man matat den med, och bygger en modell som mha prediktorerna är bäst på att identifiera de patienter som R vet hamnade på ICU?
-Det gör vi för att optimera modellens kapacitet att "förutse" vad som kommer hända med en "helt ny" patient med en uppsättning värden på prediktorerna? 
-Så har vi i Validation Data sparat en uppsättning "helt nya" patienter som vi prövar modellen på? Och ser om modellen är bra på att ge ICU-patienter rimliga propensity scores?
-Är det det du menar att modellen ska "träna" på ett set och testas på ett annat?
+fråga: vi är ju ute efter ge vissa non-ICU pat höga propensity scores för att sedan kunna identifiera och filtrera fram dem i propensity score matchingen steg 5.
+Men hur identifierar man skillnaden mellan en patient som förmodligen borde fått ICU, och en patient modellen felaktigt spottat ut?
 Och om man switchar över redan i steg 4 finns det ingen validering sen mera för dålig PSM i steg 5? -->
@@
 Fifth step is PSM.
-In PSM members of the subset that did not get ICU care but had similar predictors - and therefore similar propensity scores - as the subgroup that did get ICU care, are matched to patients who did go to the ICU.
+In PSM members of the subset that did not get ICU care are matched to patients who did go to the ICU, matching based on their propensity scores.
 There are a few different ways to perform this matching but they are all imperfect because propensity scores are unique and perfect matching is unattainable.
@@
 3. är beskrivningen av åtgärder mot detta något som hör hemma här eller i results?-->
-
-The proportion of the subset of patients that probably should hade received ICU care based on our prediction model versus the subset that did receive ICU care is also the proportion of ICU bed shortage.
 

### 2022-10-01 qualpredictors.R

@@ -82,8 +82,8 @@ for (a in 1:qualcolumncount) {
   currentqual=qualcolname[a,1] #namn på kolumn
   currentlong=quallong[a,1]
   #currentcolumn=paste("titcodata$",currentqual, sep= "")
-  currentcol =titcodata %>% select(all_of(currentqual)) #hela kolumnen
-  currentcolchr=titcodata[1:nrow(titcodata),currentqual] #i chr form
+  currentcol =set2use %>% select(all_of(currentqual)) #hela kolumnen
+  currentcolchr=set2use[1:nrow(set2use),currentqual] #i chr form
   
   variants=qualstat[["CatTable"]][["Overall"]][[currentqual]][["level"]]
   variants=data.frame(variants)
@@ -122,7 +122,7 @@ for (a in 1:qualcolumncount) {
       freqtitco[loopno,1]=freqthisvar
     } #total in set
 
-    qualstatmatrix[loopno,6]=freqtitco
+    qualstatmatrix[loopno,6]=freqtitco[loopno,1]
     qualstatmatrix[loopno,7]=freqthisvar/(freqtitco[loopno,1]) #återkom
     
     qualstatmatrix[loopno,8]=percthisvar

### 2022-10-01 textfiles.R

@@ -223,8 +223,8 @@ for (b in 1:3) {
 #ok, nu vill jag skapa en matris som för varje ICD-kod som förekommer på ICU skapar en motsvarande 
 n=0
 
-frequencies= data.frame(matrix(, nrow = nrow(xraysortICU), ncol =5 ))
-colnames(frequencies)<-data.frame("ICD code","freq ICU","freq no-ICU","freq TITCO","propICUvsnoICU")
+frequencies= data.frame(matrix(, nrow = nrow(xraysortICU), ncol =6 ))
+colnames(frequencies)<-data.frame("ICD code","ICD label","freq ICU","freq no-ICU","freq TITCO","propICUvsnoICU")
 for (c in 1:nrow(xraysortICU)) {
        str1 <- xraysortICU[c,1]
        str1 = as.character(str1)
@@ -232,17 +232,22 @@ for (c in 1:nrow(xraysortICU)) {
        
        #hitta på vilken row denna finns i TITCOsetets ICD rad
       findrow= which(xraysortTITCO$Var1 == str1, arr.ind = TRUE)
+    
       #översätt till ICD-kod
       str2=xraysortTITCO[findrow,1]
       str2 = as.character(str2) #byt till rätt form
       strfreqTITCO=xraysortTITCO[c,2]
-      
+      findrowICD= which(icdcodes[1:nrow(icdcodes),1] == str1, arr.ind = TRUE)
+      ICDlabel=icdcodes[findrowICD,2]
        
        if (str1 == str2) {
          frequencies[c,1]=str1
-         frequencies[c,2]=strfreqICU
+         if (length(ICDlabel)>0) {
+         frequencies[c,2]=ICDlabel
+         }
+         frequencies[c,3]=strfreqICU
          strfreqTITCO=as.numeric(strfreqTITCO)
-         frequencies[c,3]=strfreqTITCO-strfreqICU #det funkade dåligt att göra samma sortering på no-ICU (summorna stämemr inte), men de är ju resten så
+         frequencies[c,4]=strfreqTITCO-strfreqICU #det funkade dåligt att göra samma sortering på no-ICU (summorna stämemr inte), men de är ju resten så
          strfreqnoICU=strfreqTITCO-strfreqICU
          strfreqICU=as.numeric(strfreqICU)
[…]

### 2022-10-01 manuscript.Rmd
html

@@
 output:
+  html_document: default
   word_document: default
-  html_document: default
   pdf_document: default
@@
 If the model is suitable, it will correctly identify patients that did go to the ICU by awarding them high propensity scores, but non-ICU patients that were similar to this cohort will also get high propensity scores.
-
 <!-- 
@@
 3. är beskrivningen av åtgärder mot detta något som hör hemma här eller i results?-->
-
 <!-- (b) Describe any methods used to examine subgroups and interactions -->

### 2022-10-03 johanna/functions.R
Some example code in R to read data, do minimal cleaning/modifications, split the dataset and do descriptive statistics.

@@ -1,0 +1,18 @@
+# Functions
+# Depends on stringr, readr and dplyr
+
+# Functions to import csv files from github base URLs.
+import_csv_from_github <- function(ghurl, delim = ",") {
+  # We replace the URL sections to get the raw-file.
+  raw_ghurl <- ghurl %>% stringr::str_replace("github.com", replacement = "raw.githubusercontent.com") %>% stringr::str_replace("blob/", replacement = "")
+  print(paste0("Reading csv file from: ", raw_ghurl))
+  if (delim == ",") {
+    # We use read_csv from the readr package as read.csv in base R cant handle very large files.
+    data <- readr::read_csv(raw_ghurl)
+  } else if (delim == ";") {
+    data <- readr::read_csv2(raw_ghurl)
+  } else {
+    stop("Please specify a delimiter, either , or ;")
+  }
+  return(data)
+}

### 2022-10-03 johanna/example_code.R
Some example code in R to read data, do minimal cleaning/modifications, split the dataset and do descriptive statistics.

@@ -1,0 +1,61 @@
+# Example code
+# Load packages
+library(tidyverse)
+library(tableone)
+library(readr)
+library(stringr)
+
+# Source the functions in functions.R
+source("johanna/functions.R")
+
+# 1. Import the data
+#
+# Read the dataset and codebook from github
+data <- import_csv_from_github("https://github.com/titco/titco-I/blob/master/titco-I-full-dataset-v1.csv")
+# data <- import_csv_from_github("https://github.com/titco/titco-I/blob/master/titco-I-limited-dataset-v1.csv")
+codebook <- import_csv_from_github("https://github.com/titco/titco-I/blob/master/codebook-titco-I-v1.csv")
+
+# 2. Do data-cleaning and create variables needed for analysis. Handling missing would also be added here.
+#
+# There are some variables that are a bit problematic to handle in the TITCO dataset, a few of them are updated here to show different
+# approaches on how to handle them.
+# age, >89 is converted to numbers only to be able to handle it as numeric. That is age 89 contain 89 and above.
+# spo2_o2_1 and spo2_o2_2 is in the codebook defined as quantitative variables, but in the data it's recorded as categorical Yes/no.
+
+# Set the variable spo2_o2_1 and spo2_o2_2 is updated in the codebook to be handled as qualitative.
+codebook <- codebook %>% rows_update(tibble(name = "spo2_o2_1", type = "qualitative")) %>%
+  rows_update(tibble(name = "spo2_o2_2", type = "qualitative"))
+
+# Updates to the dataset
+data <- data %>%
+  # Convert age to numeric
+  mutate(age = parse_number(age)) %>%
+  # Add id column to the dataset
+  mutate(id = row_number()) %>%
+  # Add our outcome, if patient has a recorded length in the ICU then treated_icu = 1
+  mutate(treated_icu = case_when(licu > 24 ~ 1,
+                                        licu < 25 ~ 0))
+
+# 3. Set seed and split into training and test samples.
+# The seed allows us to always get the same randomized sample when we run the code.
+set.seed(42)
+# Use 60% of the data for training
+train_data <- data %>% sample_frac(.60)
+# Use the remaining 40% as test data
+test_data <- anti_join(data, train_data, by = 'id')
+# Optional, Verify that no rows overlap, should return 0
+count(train_data %>% filter(id %in% test_data$id))
+
+# 4. Compare variables between ICU and non-ICU patients
+# Define the categorical and continuous variables
[…]

### 2022-11-06 manuscript.Rmd

@@
 ----
-
 Quantify the unmet need for ICU beds in tertiaty hospitals in urban India.
@@
 Their numbers in proportion to the number of ICU-admitted patients is indicative of the relative shortage of ICU beds.
+
+<!-- I would not focus too much on the TQIP aspect of this, instead I would focus on the likely shortage of ICU care in this setting and that there are few methods to quantify this -->
 
@@
 -------------
+
+<!-- You don't have to include all this in the thesis or a paper, definitely enough if you only provide a brief summary or cite the information on GitHub -->
 Hospital identification number
 
-Hospital name has been replaced with a identification number, and the conversion key has not been saved so there is not possible to reverse the conversion.
+Hospital name has been replaced with a identification number, and the conversion key has not been saved 
+so there is not possible to reverse the conversion.
 Despite this, a high degree of anonymization can not be expected because the hospitals do have some distinguishable features and these are reflected in the data collected.
@@
 -->
+
 Propensity score matching (PSM) is useful method for identifying and assembling a control group from within a data set where there is causality between the independent and dependent variables, and a randomized case-control (RCT) study to offset or balance this bias is impractical, immoral or otherwise improper.
